\chapter{The Doppler Effect}\label{ch:g12:doppler-effect}

An ambulance closes in, siren blazing, and sweeps past --- and at that instant the note
slides audibly downward, though the driver touched nothing. The siren is honest; the wave
is not: motion squeezes it ahead and stretches it behind. This chapter measures the squeeze,
then rides the same idea from sirens to radar guns, to blood in an artery, and out to a
planet betrayed by a wobble in starlight.

\section{The effect}

\begin{definition}[Doppler effect]\label{def:g12:doppler-effect:doppler}
The \emph{Doppler effect}\index{Doppler effect} is the change of the received frequency
of a wave when source and receiver move relative to each other: the received frequency
$f'$ is higher than the emitted frequency $f$ while they approach, lower while they
recede, and equal to $f$ only while their distance momentarily stops changing.
\end{definition}

\begin{remark}[Pitch, not loudness]\label{rem:g12:doppler-effect:pitch}
Two things change as the ambulance passes. The \emph{loudness} swells and fades because
the wave spreads with distance (\cref{ch:g12:sound-acoustics}); the \emph{pitch} shifts
because of the speed of approach or recession, and drops abruptly at the passing.
Loudness says ``how far''; pitch says ``how fast''.
\end{remark}

\section{A moving source: crowded wavefronts}

The key picture: once a crest leaves the source, the medium owns it. Each crest expands as
a circle at the wave speed $v$ (\cref{ch:g12:mechanical-waves}), centered on the point where
it was \emph{emitted} --- the medium neither knows nor cares that the source has moved on.

\begin{proposition}[Moving source]\label{prop:g12:doppler-effect:moving-source}
A source emits crests with period $T$ (frequency $f = 1/T$) while moving in a straight
line at speed $v_s < v$ through a medium in which the wave travels at speed $v$. An
observer at rest in the medium receives
\[
\lambda' = (v \mp v_s)\,T, \qquad f' = f\,\frac{v}{v \mp v_s},
\]
the upper signs for a source approaching, the lower for one receding.
\end{proposition}

\begin{proof}
Between two crests the first advances $vT$ while the source advances $v_s T$. Ahead, the
second crest is therefore emitted $v_s T$ \emph{closer} to the first, $\lambda' = (v - v_s)T$;
behind, the same $v_s T$ is added. The crests then sweep past a resting observer at the
medium's speed $v$, one every $\lambda'/v$ seconds, so $f' = v/\lambda' = f v/(v \mp v_s)$.
\end{proof}

\begin{omfigure}
\begin{tikzpicture}[font=\small]
  \foreach \i/\r in {0/3.0, 1/2.25, 2/1.5, 3/0.75} {
    \draw[omDef] ({0.3*\i},0) circle ({\r});
    \fill[gray] ({0.3*\i},0) circle (1.1pt);
  }
  \fill (1.2,0) circle (2pt);
  \draw[-Stealth, omThm, thick] (1.2,0) -- (1.75,0)
    node[below right=-2pt] {$\vect{v}_s$};
  \draw[Stealth-Stealth, thick] (2.55,0) -- (3.0,0);
  \node[above] at (2.78,0.05) {$\lambda'$};
  \draw[Stealth-Stealth, thick] (-3.0,0) -- (-1.95,0)
    node[midway, above] {$\lambda'$};
  \node[gray] at (0.15,0.55) {\scriptsize past positions};
\end{tikzpicture}

{\small Wavefronts of a source moving right: each circle is centered on the point where
it was emitted (gray dots), so the crests bunch ahead, $\lambda' = (v - v_s)T$, and
stretch behind, $\lambda' = (v + v_s)T$.}
\end{omfigure}

\begin{example}[The ambulance, in numbers]\label{ex:g12:doppler-effect:siren}
A siren at $f = \qty{700}{Hz}$ approaches at $v_s = \qty{25}{m/s}$; take
$v = \qty{340}{m/s}$ in air (\cref{ch:g12:sound-acoustics}). Approaching:
$f' = 700 \times 340/315 \approx \qty{756}{Hz}$; receding:
$f' = 700 \times 340/365 \approx \qty{652}{Hz}$. The pass drops the pitch by \qty{104}{Hz}
--- a ratio $\approx 1.16$, about two and a half semitones, unmistakable to any ear.
\end{example}

\begin{remark}[What motion does not change]\label{rem:g12:doppler-effect:speed-unchanged}
The wave still travels at $v$: the medium alone sets the speed
(\cref{ch:g12:mechanical-waves}). Motion of the source squeezes \emph{wavelengths}, never
speeds. And the squeeze has a limit: as $v_s \to v$ the crests ahead pile onto each other
and $f' \to \infty$ --- the traffic jam of wavefronts a supersonic aircraft drags as a shock.
\end{remark}

\section{A moving observer, and the slow-motion shortcut}

\begin{proposition}[Moving observer]\label{prop:g12:doppler-effect:moving-observer}
An observer moves at speed $v_o$ straight toward (or away from) a source at rest in the
medium. The received frequency is
\[ f' = f\left(1 \pm \frac{v_o}{v}\right) \quad (+\ \text{approaching},\ -\ \text{receding}). \]
\end{proposition}

\begin{proof}
The wave pattern itself is undisturbed: crests of spacing $\lambda = vT$ travel at $v$. An
observer running at them meets crests at the relative speed $v + v_o$, one every
$\lambda/(v + v_o)$ seconds: $f' = (v + v_o)/\lambda = f(1 + v_o/v)$. Running away, $v - v_o$.
\end{proof}

\begin{definition}[Radial velocity]\label{def:g12:doppler-effect:radial}
The \emph{radial velocity}\index{radial velocity} $v_r$ of a source relative to an
observer is the component of their relative velocity along the line joining them --- the
rate at which the distance shrinks or grows. Only this component shifts the frequency;
motion \emph{across} the line of sight gives no shift at this order.
\end{definition}

\begin{proposition}[Slow motion: one formula for everybody]\label{prop:g12:doppler-effect:slow}
When the radial speed is small, $v_r \ll v$, it no longer matters who moves:
\[
\frac{\Delta f}{f} = \frac{f' - f}{f} \approx +\frac{v_r}{v} \ \text{(approach)},
\qquad
\frac{\Delta f}{f} \approx -\frac{v_r}{v} \ \text{(recession)}.
\]
\end{proposition}

\begin{proof}
The observer formulas are already exactly $1 \pm v_o/v$. For the source, write $x = v_s/v$:
the identity $\frac{1}{1 - x} = 1 + x + \frac{x^2}{1 - x}$ shows that $f'/f$ differs from
$1 + x$ by a term of order $x^2$ --- for $x = 0.1$, a $1\%$ correction on a $10\%$ shift.
To first order all four formulas collapse onto $1 \pm v_r/v$.
\end{proof}

\begin{omfigure}
\begin{tikzpicture}
\begin{axis}[omaxis, width=8.8cm, height=4.8cm,
    xlabel={$t$ (\unit{s})}, ylabel={$f'$ (\unit{Hz})},
    xmin=-8, xmax=8, ymin=408, ymax=472,
    ytick={420,440,460}, xtick={-8,-4,0,4,8}]
  \addplot[dashed, omExo, domain=-8:8, samples=2] {460.2};
  \addplot[dashed, omExo, domain=-8:8, samples=2] {421.4};
  \addplot[dotted, omProp, thick, domain=-8:8, samples=2] {440};
  \addplot[omDef, thick, domain=-8:8, samples=120]
    {440*340/(340 + 15*(15*x)/sqrt((15*x)^2 + 64))};
\end{axis}
\end{tikzpicture}

{\small Frequency heard as a \qty{440}{Hz} siren passes at \qty{15}{m/s}, \qty{8}{m}
from the microphone: plateaus at $fv/(v \mp v_s)$, and the true $f$ (dotted) crossed
exactly at closest approach, where the motion is purely transverse.}
\end{omfigure}

\begin{method}[Reading a pass]\label{met:g12:doppler-effect:pass}
A recording of a passing source shows two plateaus: $f_1$ (approach) and $f_2$
(recession). Then
\[
f = \frac{2 f_1 f_2}{f_1 + f_2}, \qquad
v_s = v\,\frac{f_1 - f_2}{f_1 + f_2}.
\]
Indeed $f_1 + f_2 = \frac{2fv^2}{v^2 - v_s^2}$ and $f_1 - f_2 = \frac{2fv\,v_s}{v^2 - v_s^2}$:
divide for $v_s$, and compute $2f_1 f_2/(f_1 + f_2)$ for $f$. Two frequency readings yield
both the true pitch and the speed --- no stopwatch, no ruler.
\end{method}

\section{Echoes that measure speed}

Bounce a wave off a moving target and the Doppler effect strikes twice.

\begin{proposition}[Reflection doubles the shift]\label{prop:g12:doppler-effect:double}
A wave of frequency $f$ is sent at a reflector approaching head-on at speed $u \ll v$.
The echo returns with
\[ f' = f\,\frac{v + u}{v - u}, \qquad \text{hence} \qquad \Delta f \approx \frac{2u}{v}\,f . \]
\end{proposition}

\begin{proof}
Two shifts in series. As a moving \emph{observer} the reflector receives $f(1 + u/v)$;
re-emitting what it receives, it is now a moving \emph{source}, so the echo arrives at
$f' = f(1 + u/v)/(1 - u/v) \approx f(1 + 2u/v)$ for $u \ll v$.
\end{proof}

\begin{example}[The radar gun]\label{ex:g12:doppler-effect:radar}
A traffic radar emits microwaves at $f = \qty{24.15}{GHz}$ ($v = c = \qty{3.00e8}{m/s}$)
and superposes the echo on the outgoing wave; the beat (\cref{ch:g12:sound-acoustics})
directly sounds out $\Delta f = 2uf/c \approx \qty{161}{Hz}$ per \unit{m/s} of car speed
--- a trivially measurable audio frequency carved out of a \qty{24}{GHz} carrier. The
factor $2$ is not optional: forgetting it flatters every driver by half.
\end{example}

\begin{example}[Doppler ultrasound]\label{ex:g12:doppler-effect:ultrasound}
A medical probe sends $f = \qty{5.0}{MHz}$ ultrasound into an artery
($v \approx \qty{1540}{m/s}$ in tissue); red blood cells reflect it. Blood at
$u = \qty{0.50}{m/s}$ returns $\Delta f = 2 \times 0.50 \times \num{5.0e6}/1540 \approx
\qty{3.2}{kHz}$ --- an \emph{audible} beat: the cardiologist literally hears the blood
accelerate at each heartbeat, then measures $u$ from $\Delta f$.
\end{example}

\section{Doppler in starlight}

Light is a wave (\cref{ch:g12:light-as-wave}), its spectrum striped with sharp lines at
wavelengths fixed by each chemical element (\cref{ch:g10:light-spectra}) --- a
laboratory-calibrated ruler printed on every star.

\begin{definition}[Redshift and blueshift]\label{def:g12:doppler-effect:redshift}
When the spectral lines of a source appear at wavelengths \emph{longer} than in the
laboratory, the spectrum is \emph{redshifted}\index{redshift}; when they appear \emph{shorter},
it is \emph{blueshifted}\index{blueshift} --- shifted toward the red or the blue end of
the visible spectrum.
\end{definition}

\begin{proposition}[Doppler shift of light]\label{prop:g12:doppler-effect:light}
For a source receding at radial speed $v_r \ll c$, every wavelength is observed
stretched by
\[ \frac{\Delta \lambda}{\lambda} = \frac{\lambda' - \lambda}{\lambda} \approx \frac{v_r}{c} \quad \text{(redshift)}, \]
and compressed by the same amount, $\Delta\lambda/\lambda \approx -v_r/c$, for an approaching
source (blueshift).
\end{proposition}
\admitted

\begin{remark}[Where the honest formula lives]\label{rem:g12:doppler-effect:honest}
Light needs no medium, so the sound derivation cannot be copied --- there is no ``at rest
in the medium''. The exact formula comes from special relativity, later this year
(\cref{ch:g12:special-relativity}); the corrections are of order $(v_r/c)^2$, negligible for
planes, stars and nearby galaxies. Below we use the slow-motion formula with a clear conscience.
\end{remark}

\begin{omfigure}
\begin{tikzpicture}[font=\small]
  \foreach \y/\xl/\c/\t in {2.4/1.8/black/laboratory,
      1.2/0.6/omDef/approaching star, 0/3.72/omThm/receding star} {
    \fill[gray!18] (0,\y) rectangle (6,\y+0.55);
    \draw[\c, very thick] (\xl,\y) -- (\xl,\y+0.55);
    \node[right] at (6.15,\y+0.28) {\t};
  }
  \draw[dashed, gray] (1.8,-0.15) -- (1.8,2.95);
  \draw[-Stealth] (-0.2,-0.55) -- (6.4,-0.55)
    node[below left] {$\lambda$ (\unit{nm})};
  \foreach \x/\l in {0/656.0, 3/656.5, 6/657.0}
    \draw (\x,-0.48) -- (\x,-0.62) node[below] {\l};
\end{tikzpicture}

{\small The same hydrogen line in three spectra: at \qty{656.30}{nm} in the laboratory,
blueshifted for an approaching star, redshifted for a receding one. The shift, read
against the dashed reference, gives the radial velocity.}
\end{omfigure}

\begin{example}[Weighing a star's escape]\label{ex:g12:doppler-effect:vega}
In the spectrum of the bright star Vega, the hydrogen line of laboratory wavelength
\qty{656.30}{nm} is measured at \qty{656.27}{nm}: $\Delta\lambda = \qty{-0.03}{nm}$, a
blueshift, so Vega approaches at $v_r = c\,|\Delta\lambda|/\lambda \approx \qty{1.4e4}{m/s}$
--- \qty{14}{km/s}, read off a shift of one part in twenty thousand.
\end{example}

\begin{example}[Wobbling stars: binaries and exoplanets]\label{ex:g12:doppler-effect:wobble}
When two stars orbit each other, their lines swing periodically blue and red --- a
\emph{spectroscopic binary}, orbits measured without ever resolving the pair. A planet does
the same to its star, faintly: both circle their common center of mass, so the star's radial
velocity oscillates by a few tens of \unit{m/s} --- a shift $\Delta\lambda/\lambda \sim 10^{-7}$.
Detecting it is how the first planet around a Sun-like star was found; the weekend problem
reruns that discovery, numbers and all.
\end{example}

\begin{remark}[Galaxy redshifts]\label{rem:g12:doppler-effect:expansion}
The spectra of distant galaxies are all \emph{red}shifted, and the more distant, the more
shifted: the universe's distances are stretching. For nearby galaxies
$v_r = c\,\Delta\lambda/\lambda$ reads off the recession speed; for the farthest, shifts
outgrow what the slow-motion formula can honestly handle, and the accounting is taken up
again from \cref{ch:g12:special-relativity} onward, and in the university volumes.
\end{remark}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:g12:doppler-effect:1}
A fire-engine siren emits at \qty{700}{Hz} and drives at \qty{25}{m/s} ($v = \qty{340}{m/s}$).
Compute the frequency heard by a pedestrian (a) ahead of it; (b) behind it. (c) What does
the driver hear?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g12:doppler-effect:2}
A friend claims: ``the pitch drops as the ambulance leaves because the sound gets weaker
with distance.'' Untangle the two effects being confused, and say what each depends on.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g12:doppler-effect:3}
A cyclist rides at \qty{6.0}{m/s} straight toward a stationary siren emitting at \qty{700}{Hz}.
What frequency do they hear? And riding straight away? Which formula applies, and why not
the moving-source one?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g12:doppler-effect:4}
A train horn emits at \qty{400}{Hz} while the train runs at \qty{30}{m/s}. Compute the
wavelength ahead of the train, behind it, and at rest. Which observer receives which?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g12:doppler-effect:5}
A phone playing a \qty{440}{Hz} tone is carried by a runner at \qty{5.0}{m/s}. Use the
slow-motion formula to estimate the shift heard by someone the runner approaches. Compare
it to a semitone (about $6\%$ in frequency): would a musician notice?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:doppler-effect:6}
A bat flies at \qty{6.0}{m/s} straight at a wall, emitting at \qty{50.0}{kHz}. Show that
the echo it hears returns at $f' = f(v + v_b)/(v - v_b)$, then compute $f'$ and the shift.
Why is this the radar-gun formula in disguise?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:doppler-effect:7}
A radar gun at \qty{24.15}{GHz} measures a beat of \qty{4.0}{kHz} from an approaching car.
Find the car's speed in \unit{m/s} and \unit{km/h}. What beat would a car at exactly
\qty{50}{km/h} give?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:doppler-effect:8}
A \qty{5.0}{MHz} Doppler ultrasound probe aimed along an artery ($v = \qty{1540}{m/s}$ in
tissue) hears a beat of \qty{2.6}{kHz}. Find the blood speed. Why must the probe be angled
\emph{along} the flow rather than perpendicular to it?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:doppler-effect:9}
Take $f = \qty{1000}{Hz}$, $v = \qty{340}{m/s}$ and a closing speed of \qty{34}{m/s}.
Compute $f'$ exactly when (a) the source moves, (b) the observer moves. Compare both to
the slow-motion prediction $f(1 + v_r/v)$ and explain the size of the disagreement.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:doppler-effect:10}
The recording of a scooter passing a microphone shows plateaus at \qty{465}{Hz} and
\qty{415}{Hz}. Using \cref{met:g12:doppler-effect:pass}, find the horn's true frequency and
the scooter's speed in \unit{km/h}. Why is the true frequency \emph{not} their average?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:doppler-effect:11}
In a star's spectrum, the hydrogen line of laboratory wavelength \qty{656.3}{nm} is observed
at \qty{656.5}{nm}. Redshift or blueshift? Compute the star's radial velocity and direction.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g12:doppler-effect:12}
A star wobbles at \qty{55}{m/s} because of an unseen planet. Compute the amplitude of the
wavelength swing of a \qty{500}{nm} line, as a length and as a fraction of $\lambda$. Order
of magnitude: why did planet hunting have to wait for spectrographs stable to one part in
$10^{7}$?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g12:doppler-effect:13}
Every line in a galaxy's spectrum is found stretched by $2.4\%$. Compute its recession speed.
Astronomers find that such speeds grow in proportion to distance, for every direction of the
sky: what picture does this suggest? Is the slow-motion formula still trustworthy here?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g12:doppler-effect:14}
The Sun rotates: one edge of its disk approaches us at about \qty{2.0}{km/s} while the other
recedes equally fast. What does this do to the \qty{656.3}{nm} line in light from the whole
disk --- shift it or broaden it? Compute the effect in \unit{nm}.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g12:doppler-effect:15}
A police car at \qty{40}{m/s}, siren at \qty{700}{Hz}, chases a truck driving at \qty{30}{m/s}
in the same direction. (a) Justify $f' = f(v - v_o)/(v - v_s)$ for the truck driver.
(b) Compute $f'$. (c) Show that if the truck matched the car's speed the shift would vanish,
and say why that is as it should be.
\end{exercise}

\section{Problem: The Hunt for an Exoplanet}

\begin{problem}[{Weekend problem --- the hunt for an exoplanet: a train horn calibrates
the ear, a radar gun rehearses the double shift, and by Sunday night a star's spectrum,
wobbling by one part in five million, has weighed a planet nobody has ever seen}]
\label{pb:g12:doppler-effect:1}
An astronomy club spends a weekend on one idea --- the Doppler shift --- rehearsed on Earth,
then aimed at a Sun-like star. Data: $v = \qty{340}{m/s}$ for sound in air,
$c = \qty{3.00e8}{m/s}$, $G = \qty{6.67e-11}{N.m^2/kg^2}$; the star has mass
$M = \qty{2.0e30}{kg}$. Admitted, from the mechanics of circular orbits
(\cref{ch:g12:satellites-kepler}): a planet of mass $m$ on a circular orbit of radius $r$
has period and speed given by $r^3 = \frac{G M P^2}{4\pi^2}$ and $V = \frac{2\pi r}{P}$,
and the star, circling the common center of mass, wobbles at speed $K = \frac{m}{M}\,V$.

\medskip
\noindent\textbf{Part I --- Friday: the level crossing.}
A train passes at constant speed, horn on; a phone records $f_1 = \qty{471}{Hz}$ while it
approaches, $f_2 = \qty{411}{Hz}$ after it passes.
\begin{enumerate}
  \item Explain, with the crowded-wavefront picture, why $f_1 > f_2$ although the horn never changes.
  \item Write the two equations linking $f_1$, $f_2$ to the true frequency $f$ and train speed $u$.
  \item Show that $u = v\,\dfrac{f_1 - f_2}{f_1 + f_2}$ and compute it in \unit{m/s} and \unit{km/h}.
  \item Show that $f = \dfrac{2 f_1 f_2}{f_1 + f_2}$ and compute it. Why is $f$ slightly
        \emph{below} the average of $f_1$ and $f_2$?
  \item Compute the wavelengths ahead of and behind the train, and the at-rest wavelength.
\end{enumerate}

\medskip
\noindent\textbf{Part II --- Saturday: the radar gun.}
A traffic patrol demonstrates its radar: $f = \qty{24.125}{GHz}$, and the gun reads the
beat between the outgoing wave and the echo.
\begin{enumerate}[resume]
  \item Explain why the echo of a moving car is shifted \emph{twice} --- name the role the
        car plays in each shift.
  \item Show that for a car at speed $u \ll c$ the beat is $\Delta f \approx \dfrac{2u}{c}\,f$.
  \item A car returns $\Delta f = \qty{3.55}{kHz}$: compute $u$.
  \item Convert to \unit{km/h} and compare to the \qty{80}{km/h} limit posted there.
  \item How many hertz of beat correspond to \qty{1}{km/h}? Comment: what makes such a tiny
        relative shift of a \qty{24}{GHz} wave easy to measure?
\end{enumerate}

\medskip
\noindent\textbf{Part III --- Sunday: the star that wobbles.}
The club downloads the measured radial velocity of a Sun-like star, night after night,
plotted below.

\begin{omfigure}
\begin{tikzpicture}
\begin{axis}[omaxis, width=8.8cm, height=4.6cm,
    xlabel={$t$ (days)}, ylabel={$v_r$ (\unit{m/s})},
    xmin=0, xmax=8.4, ymin=-72, ymax=72,
    xtick={0,2.1,4.2,6.3,8.4},
    xticklabels={$0$,$2.1$,$4.2$,$6.3$,$8.4$}, ytick={-55,0,55}]
  \addplot[dashed, omExo, domain=0:8.4, samples=2] {55};
  \addplot[dashed, omExo, domain=0:8.4, samples=2] {-55};
  \addplot[omDef, thick, domain=0:8.4, samples=140] {55*sin(360*x/4.2)};
\end{axis}
\end{tikzpicture}

{\small The star's radial velocity over twelve nights: a clean sinusoid --- the
signature of a body on a circular orbit, seen edge-on.}
\end{omfigure}

\begin{enumerate}[resume]
  \item What is actually measured, night after night, in the star's \emph{spectrum}? State
        the relation used to turn it into $v_r$.
  \item Read the amplitude $K$ and the period $P$ of the wobble off the curve.
  \item Compute the corresponding swing $\Delta\lambda$ of a \qty{550}{nm} line, and
        $\Delta\lambda/\lambda$. Compare with Part I: how much harder is this measurement
        than the train's?
  \item Why does an unseen planet make the star wobble at all? What does the sinusoidal
        shape indicate about the orbit?
  \item Why does this method measure only the \emph{radial} part of the star's motion, and
        what would be seen if the orbit were face-on to us?
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- Sunday night: weighing the invisible.}
Take the orbit circular and edge-on, so $K$ is the star's full orbital speed.
\begin{enumerate}[resume]
  \item From $P$ and the admitted relation $r^3 = G M P^2/(4\pi^2)$, compute the planet's
        orbital radius $r$. Compare it to the Earth--Sun distance, \qty{1.5e11}{m}.
  \item Compute the planet's orbital speed $V = 2\pi r/P$.
  \item From $K = (m/M)\,V$, compute the planet's mass $m$.
  \item Compare $m$ to Jupiter ($\qty{1.9e27}{kg}$) and to Earth ($\qty{6.0e24}{kg}$), and
        $r$ to the Sun--Mercury distance ($\qty{5.8e10}{m}$). What sort of world is this?
  \item If the orbit were tilted rather than edge-on, would the true mass be larger or
        smaller than your value? Conclude in one sentence: what did a periodic shift of one
        part in five million just deliver?
\end{enumerate}
\end{problem}
