\chapter{Satellites and Planetary Motion}\label{ch:g12:satellites-kepler}

The dish on the balcony is bolted down, yet all day it stares at one
fixed point of the sky, \qty{36000}{km} up; the space station
crosses the evening sky in minutes; a constellation of flying clocks
tells your phone where you stand. None of these machines has an
engine running: they are all simply \emph{falling}. This chapter
computes how fast and how high each must fall --- then weighs
planets with the same law.

\section{Falling around the Earth}

Everything rests on the law measured earlier
(\cref{ch:g10:universal-gravitation}): the Earth, of mass $M$, pulls
a \omterm{def:g10:universal-gravitation:satellite}{satellite} of mass $m$ whose center is a distance $r$ from its own
with a \omterm{def:g10:inertia:force}{force} directed toward the Earth's center, of magnitude
\[ F = G\,\frac{m M}{r^2}, \qquad G = \qty{6.67e-11}{N.m^2/kg^2}. \]
Mind what $r$ is: the distance \emph{between centers}. A \omterm{def:g10:universal-gravitation:satellite}{satellite} at
altitude $h$ has $r = R + h$, with $R = \qty{6.37e6}{m}$ the Earth's
radius --- forgetting the $R$ is the classic blunder of this chapter.

\begin{remark}[Newton's cannon, revisited]\label{rem:g12:satellites-kepler:cannon}
Fired horizontally, a cannonball lands; faster, it lands farther;
at the right speed the ground curves away exactly as fast as the
ball falls, and the fall closes on itself: an \omterm{def:g10:universal-gravitation:satellite}{orbit}. A \omterm{def:g10:universal-gravitation:satellite}{satellite}
needs no engine --- only the right \emph{sideways} speed.
\end{remark}

\begin{omfigure}
\begin{tikzpicture}[scale=1.0, font=\small]
  \fill[omExo!15] (0,0) circle (1.5); \draw[omExo, thick] (0,0) circle (1.5);
  \fill[omExo!60] (-0.22,1.44) -- (0,1.85) -- (0.22,1.44) -- cycle;
  \coordinate (L) at (0,1.85);
  \draw[omProp, thick] (L) .. controls (0.7,1.85) .. (50:1.5);
  \draw[omProp, thick] (L) .. controls (1.4,1.85) and (2.0,0.9) .. (0:1.5);
  \draw[-Stealth, omDef, thick] (0,1.85) arc (90:-260:1.85);
  \draw[-Stealth, omThm, thick] (L) .. controls (1.7,2.0) and (3.2,2.2)
    .. (4.3,2.9);
  \node[omDef, anchor=east] at (-1.85,1.1) {orbit};
  \node[omThm, anchor=west] at (3.35,2.35) {escape};
  \node[omProp, anchor=west] at (1.35,0.55) {falls};
\end{tikzpicture}

{\small Newton's cannon: fired ever faster, the ball lands farther
(orange); at about \qty{7.9}{km/s} its fall matches the Earth's
curvature and it \omterm{def:g10:universal-gravitation:satellite}{orbits} (blue); faster still, it escapes (red).}
\end{omfigure}

\section{The circular orbit: one speed, one period}

\begin{theorem}[Orbital speed]\label{thm:g12:satellites-kepler:speed}
A \omterm{def:g10:universal-gravitation:satellite}{satellite} in \omterm{def:g12:kinematics-2d:ucm}{uniform circular} \omterm{def:g10:universal-gravitation:satellite}{orbit} of radius $r$ around a body of
mass $M$ travels at the speed
\[ v = \sqrt{\frac{G M}{r}}, \]
which does not depend on the \omterm{def:g10:universal-gravitation:satellite}{satellite}'s mass.
\end{theorem}

\begin{proof}
Gravity is the only \omterm{def:g10:inertia:force}{force}, so \omterm{thm:g12:newtons-laws:second}{Newton's second law}
(\cref{thm:g12:newtons-laws:second}) reads $m\vect a = \vect F$;
project it on the \omterm{def:g12:kinematics-2d:frenet}{Frenet frame} (\cref{def:g12:kinematics-2d:frenet}).
The \omterm{def:g10:inertia:force}{force} aims at the center: its tangential component is zero, so
the speed is constant; its normal component is the whole $GmM/r^2$,
and circular motion has $a_n = v^2/r$
(\cref{thm:g12:kinematics-2d:centripetal}). Hence $m v^2/r =
GmM/r^2$, and --- $m$ dividing out --- $v^2 = GM/r$.
\end{proof}

\begin{omfigure}
\begin{tikzpicture}[scale=1.0, font=\small]
  \fill[omExo!15] (0,0) circle (0.85); \draw[omExo, thick] (0,0) circle (0.85);
  \node at (0,0) {Earth}; \draw[dashed, omExo] (0,0) circle (2.3);
  \draw[thin] (0,0) -- (40:2.3) node[midway, above left] {$r$};
  \fill (40:2.3) circle (2.2pt) node[above right] {$S$};
  \draw[-Stealth, omThm, very thick] (40:2.3) -- (40:1.35)
    node[pos=0.85, below right] {$\vect F$};
  \draw[-Stealth, omDef, thick] (40:2.3) -- ++(130:1.15)
    node[above] {$\vect v$};
\end{tikzpicture}

{\small The whole mechanism of an \omterm{def:g10:universal-gravitation:satellite}{orbit}: velocity tangent, \omterm{def:g10:inertia:force}{force}
toward the center --- gravity never slows the \omterm{def:g10:universal-gravitation:satellite}{satellite}, it only
bends its path, forever.}
\end{omfigure}

\begin{proposition}[Orbital period]\label{prop:g12:satellites-kepler:period}
The \omterm{def:g10:universal-gravitation:satellite}{satellite} completes its \omterm{def:g10:universal-gravitation:satellite}{orbit} in the \omterm{def:g12:kinematics-2d:ucm}{period}
\[ T = \frac{2\pi r}{v} = 2\pi \sqrt{\frac{r^3}{G M}}. \]
\end{proposition}

\begin{proof}
One lap is the circumference $2\pi r$ at constant speed; insert
$v = \sqrt{GM/r}$: $T = 2\pi r\sqrt{r/(GM)} = 2\pi\sqrt{r^3/(GM)}$.
\end{proof}

Neither $v$ nor $T$ contains $m$: a screw shaken loose from the
space station \omterm{def:g10:universal-gravitation:satellite}{orbits} right alongside it. And both fall with $r$:
\emph{the higher the \omterm{def:g10:universal-gravitation:satellite}{orbit}, the slower the \omterm{def:g10:universal-gravitation:satellite}{satellite}}, in speed
($v \propto 1/\sqrt r$) and in angle ($T \propto r^{3/2}$).

\begin{example}[The space station]\label{ex:g12:satellites-kepler:iss}
The International Space Station flies at altitude
$h \approx \qty{400}{km}$, so $r = \num{6.37e6} + \num{4.0e5} =
\qty{6.77e6}{m}$; with $GM = \num{6.67e-11} \times \num{5.97e24} =
\qty{3.98e14}{m^3/s^2}$, $v = \sqrt{\num{3.98e14}/\num{6.77e6}}
\approx \qty{7.7e3}{m/s}$ and $T = 2\pi r/v \approx \qty{5.5e3}{s}
\approx \qty{92}{min}$: a lap of the planet every hour and a half,
about sixteen sunrises a day for the astronauts.
\end{example}

\begin{example}[The Moon falls like the apple]\label{ex:g12:satellites-kepler:moon}
The Moon, at $r = \qty{3.84e8}{m}$, \omterm{def:g10:universal-gravitation:satellite}{orbits} in $T = \qty{27.3}{d} =
\qty{2.36e6}{s}$, so $v = 2\pi r/T \approx \qty{1.02e3}{m/s}$ and
$a = v^2/r \approx \qty{2.7e-3}{m/s^2}$ --- precisely $g/60^2$, and
the Moon sits $60$ Earth radii away: the pull that drops the apple
bends the Moon, weakened by the inverse square. Each second the
Moon ``falls'' about \qty{1.4}{mm} toward us, its sideways
kilometer carrying it around --- the check Newton made in 1666.
\end{example}

\section{Kepler's laws}

Newton did not guess the inverse square: he extracted it from three
regularities that Johannes Kepler distilled, between 1609 and 1619,
from Tycho Brahe's twenty years of naked-eye planet positions ---
the best data of the pre-telescope world.

\begin{definition}[Ellipse]\label{def:g12:satellites-kepler:ellipse}
An \emph{ellipse}\index{ellipse} is the set of points whose
distances to two fixed points, the \emph{foci}\index{focus (of an
ellipse)}, have a constant sum; half its longest diameter is the
\emph{semi-major axis}\index{semi-major axis} $a$, and a circle is
the case where the foci merge. A planet's \omterm{def:g10:universal-gravitation:satellite}{orbit} points nearest and
farthest from the Sun are its \emph{perihelion}\index{perihelion}
and \emph{aphelion}\index{aphelion}.
\end{definition}

\begin{theorem}[Kepler's laws]\label{thm:g12:satellites-kepler:kepler}
\index{Kepler's laws}%
For the planets orbiting the Sun (mass $M$):
\begin{enumerate}
  \item each \omterm{def:g10:universal-gravitation:satellite}{orbit} is an \omterm{def:g12:satellites-kepler:ellipse}{ellipse} with the Sun at one focus;
  \item the Sun--planet segment sweeps equal areas in equal times;
  \item the ratio of squared \omterm{def:g12:kinematics-2d:ucm}{period} to cubed \omterm{def:g12:satellites-kepler:ellipse}{semi-major axis} is the
        same for all planets: $T^2/a^3 = 4\pi^2/(G M)$.
\end{enumerate}
The same laws govern any family of \omterm{def:g10:universal-gravitation:satellite}{satellites} around any central
body, with $M$ the central mass. \admitted
\end{theorem}

\begin{remark}[What we can prove this year]\label{rem:g12:satellites-kepler:proof}
For a \emph{circular} \omterm{def:g10:universal-gravitation:satellite}{orbit} the third law is ours: squaring
\cref{prop:g12:satellites-kepler:period} gives
$T^2/r^3 = 4\pi^2/(GM)$ --- one constant for every \omterm{def:g10:universal-gravitation:satellite}{satellite} of the
same center. That \omterm{def:g12:satellites-kepler:ellipse}{ellipses}, equal areas and the general $a$ follow
from the inverse square is shown, with more calculus, in the Year~1
volume. Meanwhile the second law can be read: the planet moves
fastest at \omterm{def:g12:satellites-kepler:ellipse}{perihelion}, where the short sweep segment must hurry.
\end{remark}

\begin{omfigure}
\begin{tikzpicture}[scale=0.85, font=\small,
    declare function={re(\t)=2.025/(1+0.5*cos(\t));}]
  \draw[thick] plot[domain=0:360, samples=120, smooth]
    ({1.35+re(\x)*cos(\x)}, {re(\x)*sin(\x)});
  \fill[omDef!30] (1.35,0) -- plot[domain=-40:40, samples=25]
    ({1.35+re(\x)*cos(\x)}, {re(\x)*sin(\x)}) -- cycle;
  \fill[omDef!30] (1.35,0) -- plot[domain=175:185, samples=10]
    ({1.35+re(\x)*cos(\x)}, {re(\x)*sin(\x)}) -- cycle;
  \fill[omThm] (1.35,0) circle (2.6pt) node[below=2pt] {Sun};
  \draw[-Stealth, omDef, thick]
    ({1.35+re(40)*cos(40)}, {re(40)*sin(40)}) -- ++(-0.55,0.42);
  \draw[-Stealth, omDef, thick]
    ({1.35+re(175)*cos(175)}, {re(175)*sin(175)}) -- ++(-0.12,-0.5);
  \node[right] at (2.75,0) {$P$}; \node[left] at (-2.72,0) {$A$};
\end{tikzpicture}

{\small Kepler's second law: in one month near \omterm{def:g12:satellites-kepler:ellipse}{perihelion} $P$ the
planet sweeps a short, fat sector; near \omterm{def:g12:satellites-kepler:ellipse}{aphelion} $A$, a long, \omterm{def:g11:lenses-and-eye:lens}{thin}
one --- equal areas: it hurries when close, dawdles when far.}
\end{omfigure}

\begin{omfigure}
\begin{tikzpicture}
\begin{axis}[omaxis, width=8.8cm, height=5.8cm,
    xmode=log, ymode=log,
    xlabel={semi-major axis $a$ (astronomical units)},
    ylabel={period $T$ (years)},
    xmin=0.25, xmax=50, ymin=0.15, ymax=400]
  \addplot[omThm, thick] coordinates {(0.3,0.1643) (40,253)};
  \addplot[only marks, mark=*, mark size=1.7pt, omDef] coordinates
    {(0.387,0.241) (0.723,0.615) (1,1) (1.524,1.881) (5.203,11.86)
     (9.537,29.46) (19.19,84.02) (30.07,164.8)};
  \node[below right, font=\scriptsize] at (axis cs:0.387,0.241) {Mercury};
  \node[below right, font=\scriptsize] at (axis cs:1,1) {Earth};
  \node[below right, font=\scriptsize] at (axis cs:5.203,11.86) {Jupiter};
  \node[below right, font=\scriptsize] at (axis cs:30.07,164.8) {Neptune};
\end{axis}
\end{tikzpicture}

{\small The eight planets on log--log axes: a straight line of slope
$3/2$, i.e.\ $T^2 \propto a^3$ --- Kepler's third law, which in
these \omterm{def:g10:orders-of-magnitude:unit}{units} (\omterm{def:g10:orders-of-magnitude:lightyear}{astronomical units}, years) reads simply $T^2 = a^3$.}
\end{omfigure}

\section{Orbits at work}

\begin{definition}[Geostationary orbit]\label{def:g12:satellites-kepler:geo}
A \omterm{def:g10:universal-gravitation:satellite}{satellite} is \emph{geostationary}\index{geostationary orbit} when
it hangs over one fixed point of the ground. Its \omterm{def:g10:universal-gravitation:satellite}{orbit} must be
circular, equatorial, and of \omterm{def:g12:kinematics-2d:ucm}{period} one \emph{sidereal
day}\index{sidereal day}, $T = \qty{86164}{s}$ (\qty{23}{h}
\qty{56}{min} \qty{4}{s}): the time the Earth takes to turn once
\emph{relative to the stars}. (The \qty{24}{h} day is relative to
the Sun, toward which the Earth also advances a degree a day.)
\end{definition}

\begin{example}[The geostationary altitude]\label{ex:g12:satellites-kepler:geoalt}
Kepler's third law, read backward, dictates the radius:
$r = \left(GMT^2/4\pi^2\right)^{1/3} =
\left(\num{3.98e14} \times (\num{86164})^2/4\pi^2\right)^{1/3}
\approx \qty{4.22e7}{m}$ --- an altitude $h = r - R \approx
\qty{3.58e7}{m} \approx \qty{35800}{km}$, traveled at $v = 2\pi r/T
\approx \qty{3.1}{km/s}$. Every relay that ``hangs still'' sits on
this one circle over the equator --- there is no other.
\end{example}

\begin{example}[The navigation constellation]\label{ex:g12:satellites-kepler:gps}
Satellite-navigation systems fly some thirty \omterm{def:g10:universal-gravitation:satellite}{satellites} at
$r \approx \qty{2.66e7}{m}$ (altitude $\approx \qty{20200}{km}$),
where \cref{prop:g12:satellites-kepler:period} gives $T \approx
\qty{4.32e4}{s}$: half a \omterm{def:g12:satellites-kepler:geo}{sidereal day}, so each \omterm{def:g10:universal-gravitation:satellite}{satellite} retraces
its ground track daily. Your receiver compares the arrival times of
their clock \omterm{def:g10:signals-and-waves:signal}{signals} against \omterm{def:g10:universal-gravitation:satellite}{orbits} known to meters.
\end{example}

\begin{method}[Weighing a world from its satellite]\label{met:g12:satellites-kepler:weigh}
\index{weighing a planet}%
To measure the mass of a planet or a star: find any \omterm{def:g10:universal-gravitation:satellite}{satellite} of it
(a moon, a probe, a planet), measure its \omterm{def:g10:universal-gravitation:satellite}{orbit} radius $r$ and \omterm{def:g12:kinematics-2d:ucm}{period}
$T$, and invert the third law:
\[ M = \frac{4\pi^2 r^3}{G\,T^2}. \]
This weighs the \emph{central} body only --- the \omterm{def:g10:universal-gravitation:satellite}{satellite}'s mass
cancelled in \cref{thm:g12:satellites-kepler:speed}. Every data-card
mass came this way: the Earth from the Moon, the Sun from the
Earth's own \omterm{def:g10:universal-gravitation:satellite}{orbit}.
\end{method}

\begin{remark}[Higher is slower --- but costs more]\label{rem:g12:satellites-kepler:energy}
Since $v = \sqrt{GM/r}$, the Moon crawls at \qty{1}{km/s} while the
station races at \qty{7.7}{km/s}. Yet the higher \omterm{def:g10:universal-gravitation:satellite}{orbit} is the more
\emph{expensive}: climbing against gravity stores potential \omterm{def:g10:energy-conservation:energy}{energy}
(the account is settled in \cref{ch:g12:work-and-energy}), and the
gain in $E_p$ outweighs the loss in $E_k$. Hence a rocket burns
\emph{forward} to reach a higher, slower \omterm{def:g10:universal-gravitation:satellite}{orbit} --- and a \omterm{def:g10:universal-gravitation:satellite}{satellite}
braked by \omterm{def:g11:lenses-and-eye:lens}{thin} air spirals \emph{down} and speeds up.
\end{remark}

\begin{remark}[Data card]\label{rem:g12:satellites-kepler:data}
Unless stated otherwise: $G = \qty{6.67e-11}{N.m^2/kg^2}$; Earth:
$M = \qty{5.97e24}{kg}$, $R = \qty{6.37e6}{m}$, \omterm{def:g12:satellites-kepler:geo}{sidereal day}
\qty{86164}{s}; Moon's \omterm{def:g10:universal-gravitation:satellite}{orbit}: $r = \qty{3.84e8}{m}$, $T =
\qty{27.3}{d}$; Sun: $M = \qty{1.99e30}{kg}$; Earth--Sun distance
\qty{1.496e11}{m}; one year $= \qty{3.156e7}{s}$; Mars: $M =
\qty{6.42e23}{kg}$, $R = \qty{3.39e6}{m}$.
\end{remark}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:g12:satellites-kepler:1}
A \qty{1000}{kg} \omterm{def:g10:universal-gravitation:satellite}{satellite} sits on the launch pad, then flies at
altitude \qty{400}{km}. Compute the Earth's pull in both places:
what percentage survives aloft --- and why do occupants float?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g12:satellites-kepler:2}
For a (theoretical) \omterm{def:g10:universal-gravitation:satellite}{satellite} skimming the Earth's surface
($r = R$), compute the orbital speed and the \omterm{def:g12:kinematics-2d:ucm}{period}. Compare with
Newton's cannonball (\cref{rem:g12:satellites-kepler:cannon}).
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g12:satellites-kepler:3}
A \omterm{def:g10:universal-gravitation:satellite}{satellite} is moved from a circular \omterm{def:g10:universal-gravitation:satellite}{orbit} of radius $r$ to one of
radius $2r$. By what factor do its speed and \omterm{def:g12:kinematics-2d:ucm}{period} change? Do the
answers depend on its mass?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g12:satellites-kepler:4}
Redo the station's numbers: from $r = \qty{6.77e6}{m}$, compute
$v$, $T$, and the number of \omterm{def:g10:universal-gravitation:satellite}{orbits} completed in \qty{24}{h}.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g12:satellites-kepler:5}
State Kepler's three laws. Where along its very stretched \omterm{def:g12:satellites-kepler:ellipse}{ellipse}
is a comet fastest, and which law says so? Does a \omterm{def:g10:universal-gravitation:satellite}{satellite}'s
\omterm{def:g12:kinematics-2d:ucm}{period} depend on its own mass?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:satellites-kepler:6}
Weigh the Earth: from the Moon's \omterm{def:g10:universal-gravitation:satellite}{orbit} radius and \omterm{def:g12:kinematics-2d:ucm}{period} (data
card), compute the Earth's mass and compare with the card's value.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:satellites-kepler:7}
A company proposes a \omterm{def:g10:universal-gravitation:satellite}{satellite} hovering permanently above Paris
(latitude $49^\circ$ north). From the direction of gravity, explain
why every circular \omterm{def:g10:universal-gravitation:satellite}{orbit} is centered on the Earth's center; conclude
the proposal is impossible. Where \emph{can} a \omterm{def:g10:universal-gravitation:satellite}{satellite} hover?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:satellites-kepler:8}
Phobos circles Mars at $r = \qty{9.38e6}{m}$ in $T = \qty{7}{h}
\qty{39}{min}$. Deduce the mass of Mars; compare with the data card.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:satellites-kepler:9}
Mars \omterm{def:g10:universal-gravitation:satellite}{orbits} the Sun at $a = \qty{2.279e11}{m}$ in \qty{687}{d}.
Compute $T^2/a^3$ for Mars and for the Earth (data card), check
Kepler's third law, and deduce the mass of the Sun.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:satellites-kepler:10}
Newton's Moon test: from the data card, compute the Moon's speed and
its \omterm{thm:g12:kinematics-2d:centripetal}{centripetal acceleration} $a = v^2/r$. Show it equals $g/60^2$,
and say why that convinced Newton that one law rules apple and Moon.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:satellites-kepler:11}
In Sun \omterm{def:g10:orders-of-magnitude:unit}{units} (\omterm{def:g10:orders-of-magnitude:lightyear}{astronomical units}, years) Kepler's third law reads
$T^2 = a^3$: the straight line of the course's log--log plot. Place
on it the asteroid Ceres, $a = \qty{2.77}{au}$, and Halley's comet,
$a = \qty{17.8}{au}$: compute both \omterm{def:g12:kinematics-2d:ucm}{periods}. Halley last passed
\omterm{def:g12:satellites-kepler:ellipse}{perihelion} in 1986 --- when is it due back?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g12:satellites-kepler:12}
\omterm{def:g11:lenses-and-eye:lens}{Thin} air brushes the station, yet its speed \emph{increases} as it
loses altitude. Compute $v$ at altitudes \qty{400}{km} and
\qty{350}{km}, then resolve the paradox: \omterm{def:g10:inertia:inventory}{friction} slows things down
--- where does the extra speed come from?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g12:satellites-kepler:13}
A telescope finds an exoplanet circling its star at
$r = \qty{7.48e9}{m}$ every \qty{3.50}{d}. Weigh the star, in
\omterm{def:g10:orders-of-magnitude:unit}{kilograms} and in Suns. Which mass --- star's or planet's ---
\emph{cannot} be obtained this way, and why?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g12:satellites-kepler:14}
Mars settlers want an ``areostationary'' relay over their base.
Mars turns relative to the stars in \qty{24}{h} \qty{37}{min}:
compute the \omterm{def:g10:universal-gravitation:satellite}{orbit} radius and the altitude above the Martian
surface (data card).
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g12:satellites-kepler:15}
Halley's comet passes \omterm{def:g12:satellites-kepler:ellipse}{perihelion} at $r_p = \qty{8.77e10}{m}$ moving
at \qty{54.5}{km/s}; \omterm{def:g12:satellites-kepler:ellipse}{aphelion} is at $r_a = \qty{5.25e12}{m}$. At
both extremes the velocity is perpendicular to the Sun--comet
segment, and Kepler's second law then \omterm{def:g10:inertia:force}{forces} $r_p v_p = r_a v_a$
(equal sweep triangles of area $\tfrac12 r v\,\Delta t$). Deduce the
\omterm{def:g12:satellites-kepler:ellipse}{aphelion} speed and comment on the ratio.
\end{exercise}

\section{Problem: A Seat Over the Equator}

\begin{problem}[{Weekend problem --- putting a \omterm{def:g10:universal-gravitation:satellite}{satellite} over the
equator: why the only parking spot in the sky is a circle
\qty{35800}{km} up, how it compares with the station racing below,
and how the same law weighs Jupiter on the way}]
\label{pb:g12:satellites-kepler:1}
A telecom company wants a relay that its ground dishes, bolted down
once, never have to chase: a \omterm{def:g12:satellites-kepler:geo}{geostationary} \omterm{def:g10:universal-gravitation:satellite}{satellite}. You are the
mission analyst; use the data card. For Part~IV: Jupiter's moon Io
\omterm{def:g10:universal-gravitation:satellite}{orbits} at $r = \qty{4.22e8}{m}$ in $T = \qty{1.77}{d}$, and Jupiter
turns relative to the stars in \qty{9}{h} \qty{56}{min}.

\medskip
\noindent\textbf{Part I --- The rules of hanging still.}
\begin{enumerate}
  \item The \omterm{def:g10:universal-gravitation:satellite}{satellite} must stay over one fixed ground point. What
        must its \omterm{def:g12:kinematics-2d:ucm}{period} be --- and why the \omterm{def:g12:satellites-kepler:geo}{sidereal day}
        \qty{86164}{s} rather than \qty{24}{h}?
  \item In circular motion the \omterm{def:g11:forces-and-motion:netforce}{net force} points at the circle's
        center; here the only \omterm{def:g10:inertia:force}{force} is gravity. Deduce that the
        center of any circular \omterm{def:g10:universal-gravitation:satellite}{orbit} is the Earth's center.
  \item A circle hovering above Paris's latitude would be centered
        on the Earth's axis, north of the center. Conclude: in which
        plane must a \omterm{def:g12:satellites-kepler:geo}{geostationary orbit} lie?
  \item Write \omterm{thm:g12:newtons-laws:second}{Newton's second law} in the \omterm{def:g12:kinematics-2d:frenet}{Frenet frame} for a circular
        \omterm{def:g10:universal-gravitation:satellite}{orbit} of radius $r$ and derive $v = \sqrt{GM/r}$.
  \item Deduce $T = 2\pi\sqrt{r^3/(GM)}$ and rewrite it as Kepler's
        third law $T^2/r^3 = 4\pi^2/(GM)$.
\end{enumerate}

\medskip
\noindent\textbf{Part II --- The one circle that \omterm{def:g11:work-of-force:work}{works}.}
\begin{enumerate}[resume]
  \item Solve the third law for $r$ and compute the \omterm{def:g12:satellites-kepler:geo}{geostationary}
        radius.
  \item Deduce the altitude above the ground.
  \item Compute the orbital speed from $v = 2\pi r/T$.
  \item Check it against $v = \sqrt{GM/r}$.
  \item The company's \omterm{def:g10:universal-gravitation:satellite}{satellite} has mass \qty{4.0e3}{kg}; a rival's
        weighs twice that. Compare their \omterm{def:g12:satellites-kepler:geo}{geostationary} radii, and
        say what the extra mass changes at launch.
\end{enumerate}

\medskip
\noindent\textbf{Part III --- The station below.}
\begin{enumerate}[resume]
  \item The space station flies at $r = \qty{6.77e6}{m}$: compute
        its speed.
  \item Compute its \omterm{def:g12:kinematics-2d:ucm}{period}, in seconds and minutes.
  \item How many \omterm{def:g10:universal-gravitation:satellite}{orbits} does it complete in \qty{24}{h} --- about
        how many sunrises do the astronauts watch per day?
  \item Compute the ratio of radii $r_{\text{geo}}/r_{\text{ISS}}$,
        and verify that the ratio of \omterm{def:g12:kinematics-2d:ucm}{periods} is that number to the
        \omterm{def:g11:work-of-force:power}{power} $3/2$, as Kepler's third law demands.
  \item Which \omterm{def:g10:universal-gravitation:satellite}{satellite} moves faster, and which took more \omterm{def:g10:energy-conservation:energy}{energy} per
        \omterm{def:g10:orders-of-magnitude:unit}{kilogram} to emplace? Reconcile the two in one sentence.
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- Weighing Jupiter on the side.}
\begin{enumerate}[resume]
  \item From Kepler's third law applied to Io, express Jupiter's
        mass $M_J$ in terms of $G$, $r$ and $T$.
  \item Compute $M_J$.
  \item How many Earth masses is that? Roughly what fraction of the
        Sun's mass?
  \item A ``jovistationary'' relay should hang over one point of
        Jupiter's clouds: compute its \omterm{def:g10:universal-gravitation:satellite}{orbit} radius; compare with
        Io's.
  \item Report to the board, two sentences: the altitude where the
        company must park its relay, and the mass of Jupiter that
        the very same law delivered for free.
\end{enumerate}
\end{problem}
