\chapter{Special Relativity: Time Dilation}\label{ch:g12:special-relativity}

Every GPS \omterm{def:g10:universal-gravitation:satellite}{satellite} carries an atomic clock good to one second in three
million years --- and before launch, engineers deliberately detune it: left
honest, it would gain 38 millionths of a second per day, and by tomorrow
evening your phone would place you eleven kilometres from where you stand.
The clock is not at fault --- time itself ticks differently for the moving and
the still: this chapter follows Einstein from one stubborn fact to that conclusion.

\section{One speed refuses to add}

Speeds depend on the frame and compose by addition
(\cref{ch:g10:relative-motion}): a passenger walking at \qty{4}{km/h} toward the
front of a train doing \qty{300}{km/h} passes the trackside observer at \qty{304}{km/h}.

\begin{definition}[Inertial frame]\label{def:g12:special-relativity:inertial}
An \emph{inertial frame}\index{inertial frame} is a \omterm{def:g10:relative-motion:frame}{reference frame}
(\cref{ch:g10:relative-motion}) in which a force-free body keeps a constant
velocity: the ground, near enough, and every frame in uniform
straight-line motion relative to it.
\end{definition}

Push the rule to the extreme: the train's headlight beam should pass the
ground observer at $c + \qty{300}{km/h}$. In 1887 Michelson and Morley
compared the speed of light along and across the Earth's \qty{30}{km/s}
orbital rush, hunting such differences; to exquisite precision they found
none --- the same $c$ in every direction, in every season. Einstein, in 1905,
believed the experiment and rebuilt time and space around it, on two postulates
judged by their consequences --- tested for a century now without a single failure.

\begin{theorem}[The postulates of special relativity]\label{thm:g12:special-relativity:postulates}
\index{special relativity}\index{postulates of relativity}
\begin{enumerate}
  \item \emph{Relativity principle}: the laws of physics take the same
        form in every \omterm{def:g12:special-relativity:inertial}{inertial frame}; none is ``truly at rest''.
  \item \emph{Invariance of $c$}: light in vacuum travels at
        $c = \qty{3.00e8}{m/s}$ in every \omterm{def:g12:special-relativity:inertial}{inertial frame}, whatever the
        motion of source or observer.
\end{enumerate}
\end{theorem}
\admitted

\section{Farewell to ``the same instant''}

\begin{proposition}[Relativity of simultaneity]\label{prop:g12:special-relativity:simultaneity}
\index{simultaneity}
Two \omterm{def:g12:special-relativity:proper-time}{events} simultaneous in one \omterm{def:g12:special-relativity:inertial}{inertial frame} are, in general, not
simultaneous in another: ``now, everywhere'' is frame-dependent.
\end{proposition}

\begin{proof}
Einstein's train. Two lightning bolts strike the two ends of a moving
train, scorching the track. The platform observer $M$, midway between the
marks, receives the two flashes together; each crossed the same distance at
the same speed $c$, so the strikes were simultaneous. The passenger $M'$,
midway along the train, rides \emph{toward} the front flash, which reaches
him first; but in his frame too both flashes covered equal distances (half
a train) at the same speed $c$ --- second postulate! --- so the front
strike happened first. Each verdict is correct in its own frame; the
disagreement, nanoseconds for a train, is why no one had noticed.
\end{proof}

\begin{omfigure}
\begin{tikzpicture}[scale=0.82, font=\small]
  \draw[thick] (0,0) -- (10.6,0);
  \fill (5.3,0) circle (2pt); \node[below] at (5.3,-0.05) {$M$};
  \node[below] at (1.6,-0.05) {$A$}; \node[below] at (9.0,-0.05) {$B$};
  \draw[omDef, thick] (1.6,0.55) rectangle (9.0,1.45);
  \fill[omDef] (5.3,1.45) circle (2pt); \node[above] at (5.3,1.5) {$M'$};
  \draw[-Stealth, omDef, thick] (9.2,1.0) -- (10.4,1.0) node[above left] {$\vect v$};
  \draw[omThm, thick] (1.25,2.7) -- (1.7,2.15) -- (1.45,2.05) -- (1.75,1.5);
  \draw[omThm, thick] (8.65,2.7) -- (9.1,2.15) -- (8.85,2.05) -- (9.15,1.5);
  \draw[dashed, omExo] (2.35,0) arc (0:60:0.75) (3.1,0) arc (0:60:1.5);
  \draw[dashed, omExo] (8.25,0) arc (180:120:0.75) (7.5,0) arc (180:120:1.5);
\end{tikzpicture}

{\small Einstein's train: the flashes from $A$ and $B$ reach $M$ together; $M'$
rides toward $B$'s flash, meets it first --- and concludes it struck first.}
\end{omfigure}

\section{The light clock and the $\gamma$ factor}

\begin{definition}[Event, proper time]\label{def:g12:special-relativity:proper-time}
An \emph{event}\index{event} is a happening at one place and one instant
--- a tick, a flash, a decay. When a single clock is present at two events,
the interval it reads between them is the
\emph{proper time}\index{proper time} $\Delta t_0$. A frame in which that
clock moves measures a generally different interval $\Delta t$.
\end{definition}

\begin{theorem}[Time dilation]\label{thm:g12:special-relativity:dilation}
\index{time dilation}\index{gamma factor@$\gamma$ (gamma) factor}
A clock moving at constant speed $v$ through an \omterm{def:g12:special-relativity:inertial}{inertial frame} is measured
there to run slow: between two \omterm{def:g12:special-relativity:proper-time}{events} at the clock,
\[
\Delta t = \gamma\, \Delta t_0, \qquad
\gamma = \frac{1}{\sqrt{1 - v^2/c^2}} \geq 1 .
\]
\end{theorem}

\begin{proof}
Build the simplest clock imaginable: two facing mirrors a distance $L$
apart, a \omterm{def:g12:quantum-world:photon}{photon} bouncing between them; one round trip is one tick, and at
rest a tick lasts $\Delta t_0 = 2L/c$. Now bolt the clock to a ship
crossing at speed $v$, mirrors perpendicular to the motion, and time one
tick from the ground: while the \omterm{def:g12:quantum-world:photon}{photon} climbs, the mirrors slide sideways,
so the \omterm{def:g12:quantum-world:photon}{photon} flies a slanted path --- two motions at once, like the
\omterm{def:g12:projectile-motion:projectile}{projectile} of \cref{ch:g12:kinematics-2d}. If the tick lasts $\Delta t$,
each half-tick advances $v\,\Delta t/2$ horizontally and $L$ \omterm{def:g10:universal-gravitation:weight}{vertically}; and ---
the crucial step --- the second postulate fixes the \omterm{def:g12:quantum-world:photon}{photon}'s ground speed at
$c$, so the slanted half-path measures $c\,\Delta t/2$. Pythagoras:
\[
\Bigl(\frac{c\,\Delta t}{2}\Bigr)^{\!2} = L^2 + \Bigl(\frac{v\,\Delta t}{2}\Bigr)^{\!2}
\;\Longrightarrow\;
\Delta t = \frac{2L/c}{\sqrt{1 - v^2/c^2}} = \gamma\,\Delta t_0 .
\]
And any clock riding beside it --- wristwatch, heartbeat, decaying \omterm{def:g11:fundamental-interactions:ladder}{nucleus} ---
must stay in step, or the comparison would betray the ship's motion and violate
the first postulate.
\end{proof}

\begin{omfigure}
\begin{tikzpicture}[scale=0.9, font=\small]
  \draw[thick] (-0.5,0) -- (0.5,0); \draw[thick] (-0.5,2) -- (0.5,2);
  \draw[omDef, thick, -Stealth] (0,0.05) -- (0,1.95);
  \draw[Stealth-Stealth] (0.85,0) -- (0.85,2) node[midway, right] {$L$};
  \node at (0.3,-0.75) {at rest: tick $\Delta t_0 = 2L/c$};
\end{tikzpicture}
\qquad
\begin{tikzpicture}[scale=0.9, font=\small]
  \foreach \x in {0,1.4,2.8} { \draw[gray] (\x-0.4,0) -- (\x+0.4,0);
    \draw[gray] (\x-0.4,2) -- (\x+0.4,2); }
  \draw[omDef, thick, -Stealth] (0,0.03) -- (1.4,1.97);
  \draw[omDef, thick, -Stealth] (1.4,1.97) -- (2.8,0.03);
  \draw[dashed, omExo] (1.4,0) -- (1.4,2);
  \draw[-Stealth, thick] (3.5,1.0) -- (4.4,1.0) node[above left] {$\vect v$};
  \node[above left] at (0.9,0.9) {$\frac{c\,\Delta t}{2}$};
  \node[right] at (1.4,0.75) {$L$};
  \draw[Stealth-Stealth] (0,-0.35) -- (1.4,-0.35) node[midway, below] {$v\,\Delta t/2$};
  \node at (1.5,-1.15) {moving: the photon flies the hypotenuse};
\end{tikzpicture}

{\small The light clock. At rest the \omterm{def:g12:quantum-world:photon}{photon} climbs $L$; from the ground it
flies the hypotenuse at the same $c$ --- so the tick must take longer.}
\end{omfigure}

\begin{example}[Gamma in numbers]\label{ex:g12:special-relativity:gammatable}
\begin{center}\small
\begin{tabular}{lccccccc}
$v/c$    & 0.01     & 0.1   & 0.5   & 0.8   & 0.9   & 0.99 & 0.999\\
$\gamma$ & 1.00005 & 1.005 & 1.155 & 1.667 & 2.294 & 7.09 & 22.4
\end{tabular}
\end{center}
Nothing happens for pages, then everything: $\gamma$ hugs $1$ up to a third
of light speed, passes $2$ near $0.9c$, and blows up as $v \to c$.
\end{example}

\begin{omfigure}
\begin{tikzpicture}
\begin{axis}[omaxis, width=8.6cm, height=5.4cm,
    xlabel={$v/c$}, ylabel={$\gamma$},
    xmin=0, xmax=1.08, ymin=0, ymax=8,
    xtick={0,0.2,0.4,0.6,0.8,1}, ytick={1,2,...,7}]
  \addplot[omDef, thick, domain=0:0.992, samples=200] {1/sqrt(1-x^2)};
  \addplot[omExo, dashed, thick] coordinates {(1,0) (1,8)};
\end{axis}
\end{tikzpicture}

{\small $\gamma = 1/\sqrt{1 - v^2/c^2}$: indistinguishable from $1$ at everyday
speeds, \omterm{def:g11:lenses-and-eye:lens}{diverging} at the asymptote $v = c$ --- the speed no massive body reaches.}
\end{omfigure}

\begin{remark}[Why daily life never noticed]\label{rem:g12:special-relativity:everyday}
For $v \ll c$, write $x = v^2/c^2$: since $(1-x)(1+x) \approx 1$ and
$(1 + x/2)^2 \approx 1 + x$, we get $\gamma \approx 1 + v^2/(2c^2)$. A car
at \qty{130}{km/h} has $\gamma - 1 \approx \num{7e-15}$ (one second lost
per four million years of driving); a jet manages $\num{3.5e-13}$, one
second per ninety thousand years aloft. Our clocks were simply too coarse.
\end{remark}

\section{The verdict of experiment}

\begin{example}[Muons reach the ground]\label{ex:g12:special-relativity:muons}
Cosmic rays striking the upper atmosphere create \emph{muons} ---
short-lived particles of proper lifetime $\Delta t_0 = \qty{2.2}{\micro s}$
--- about \qty{15}{km} up, moving at nearly $c$. Without relativity a muon
covers $c\,\Delta t_0 \approx \qty{660}{m}$ per lifetime: the descent takes
some $23$ lifetimes, and the surviving fraction
$\mathrm{e}^{-23} \approx \num{e-10}$ should make sea-level muons
vanishingly rare. Yet detectors count about one per square centimetre per
minute. \omterm{thm:g12:special-relativity:dilation}{Time dilation} resolves it: arriving with $\gamma \approx 30$, the
muons live $30 \times \qty{2.2}{\micro s} = \qty{66}{\micro s}$ on our
clocks --- \omterm{def:g12:projectile-motion:range}{range} $\approx \qty{20}{km}$, the whole atmosphere.
\end{example}

\begin{omfigure}
\begin{tikzpicture}[scale=0.85, font=\small]
  \draw[thick] (0,0) -- (8.6,0) node[below left] {ground};
  \draw[dashed, omExo] (0,5) -- (8.6,5);
  \node[above] at (2.1,5.02) {muons born at \qty{15}{km}};
  \draw[-Stealth, thick] (1.1,6.1) node[above right] {cosmic ray} -- (1.55,5.05);
  \draw[omThm, very thick, -Stealth] (3.4,5) -- (3.4,4.78);
  \node[right, omThm] at (3.5,4.62)
    {no dilation: $c\,\Delta t_0 \approx \qty{660}{m}$};
  \draw[omDef, very thick, -Stealth] (6.6,5) -- (6.6,0.06);
  \node[left, omDef, align=right] at (6.5,2.2)
    {$\gamma \approx 30$:\\ $\approx \qty{20}{km}$};
\end{tikzpicture}

{\small The muon's journey: one classical lifetime dies out \qty{660}{m}
below the birth altitude; the dilated lifetime spans the atmosphere.}
\end{omfigure}

Confirmations pile up. In 1971 Hafele and Keating flew caesium atomic
clocks (\cref{ch:g12:oscillators-and-time}) around the world on airliners;
on landing they differed from ground clocks by the predicted fractions of
a microsecond. Accelerators bank on dilation daily, their particles
boosted to $\gamma$ in the thousands surviving laps they could never
classically complete. And GPS is a running experiment with a price tag.

\begin{remark}[The relativity bill of GPS]\label{rem:g12:special-relativity:gps}
A GPS \omterm{def:g10:universal-gravitation:satellite}{satellite} \omterm{def:g10:universal-gravitation:satellite}{orbits} at about \qty{3.9}{km/s}, so \omterm{thm:g12:special-relativity:postulates}{special relativity}
slows its clock by some \qty{7}{\micro s} per day. Gravity, weaker at
altitude, \omterm{def:g11:work-of-force:work}{works} the other way and wins: general relativity (Einstein's
theory of \omterm{def:g11:fundamental-interactions:four}{gravitation}, told in the university volumes) speeds it by about
\qty{45}{\micro s} per day; net, \qty{38}{\micro s} per day fast.
Positioning converts clock readings to
distances at speed $c$: an uncorrected day would smear positions by
$c \times \qty{38}{\micro s} \approx \qty{11}{km}$ --- hence the detuned
clocks of the opening paragraph.
\end{remark}

\begin{proposition}[Length contraction]\label{prop:g12:special-relativity:contraction}
\index{length contraction}
A body of length $L_0$ in its rest frame is measured shorter along its
direction of motion in a frame where it moves at speed $v$:
$L = L_0/\gamma$.
\end{proposition}
\admitted

\begin{remark}[The muon's own story]\label{rem:g12:special-relativity:muonframe}
\omterm{prop:g12:special-relativity:contraction}{Length contraction} is \omterm{thm:g12:special-relativity:dilation}{time dilation}'s flip side (derived honestly in the
university volumes): the muon's clock ticks normally in its own frame, but
the atmosphere rushing up is contracted to $\qty{15}{km}/30 =
\qty{500}{m}$, crossed well within \qty{2.2}{\micro s} --- both frames
agree the muon reaches the ground.
\end{remark}

\section{Mass, energy, and the classical world regained}

\begin{theorem}[Mass--energy equivalence]\label{thm:g12:special-relativity:emc2}
\index{mass--energy equivalence}
A body of mass $m$ possesses, by its mass alone, the \omterm{def:g10:energy-conservation:energy}{energy}
\[ E = m c^2 . \]
Mass is a concentrated form of \omterm{def:g10:energy-conservation:energy}{energy}; \omterm{def:g10:energy-conservation:energy}{energy}, conversely, has mass.
\end{theorem}
\admitted

\begin{remark}[Where you have met it]\label{rem:g12:special-relativity:emc2seen}
This is the ledger behind \omterm{def:g10:energy-conservation:forms}{nuclear energy} (\cref{ch:g12:nuclear-energy}):
the \omterm{def:g12:nuclear-energy:defect}{mass defect} of a reaction, times $c^2$, is the \omterm{def:g10:energy-conservation:energy}{energy} released. One
gram of mass is \qty{9.0e13}{J} --- a full day's output of a large \omterm{def:g11:work-of-force:power}{power}
station. The honest derivation belongs to the university volumes; the
reactors and the stars run on it regardless.
\end{remark}

\begin{remark}[Newton was not wrong]\label{rem:g12:special-relativity:newton}
Set $v \ll c$: $\gamma \to 1$, times agree, lengths uncontract,
simultaneity turns absolute --- the mechanics of the last three years
reappears, intact, as the low-speed limit. Relativity did not demolish
Newton; it drew the boundary of his empire. That is how physics grows: by
refining theories, each new one handing back the old as a limiting case.
\end{remark}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:g12:special-relativity:1}
Compute $\gamma$, to three \omterm{def:g10:orders-of-magnitude:sigfig}{significant figures}, for $v = 0.1c$, $0.6c$, $0.8c$, $0.995c$.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g12:special-relativity:2}
A ship cruises at $0.6c$; its metronome beats every \qty{10.0}{s} by the ship's
clock. What interval does the ground measure? Which is the \omterm{def:g12:special-relativity:proper-time}{proper time}, and why?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g12:special-relativity:3}
At what speed (as a fraction of $c$) is $\gamma = 2$? And $\gamma = 10$?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g12:special-relativity:4}
A car drives at \qty{130}{km/h}. Compute $v/c$, then $\gamma - 1$
(\cref{rem:g12:special-relativity:everyday}); how long must it drive for
its clock to fall \qty{1}{s} behind?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g12:special-relativity:5}
In Einstein's train experiment, explain --- citing the postulate used ---
why the passenger concludes the \emph{front} strike came first, and why
neither observer is mistaken.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:special-relativity:6}
Charged pions live $\Delta t_0 = \qty{2.6e-8}{s}$. A beam leaves an
accelerator at $0.99c$. Compute $\gamma$, the lifetime in the lab, the
classical \omterm{def:g12:projectile-motion:range}{range} $v\,\Delta t_0$, and the actual mean \omterm{def:g12:projectile-motion:range}{range}.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:special-relativity:7}
From the $\gamma$ curve, read the speeds giving $\gamma = 1.5$ and
$\gamma = 3$ (check by computation). What does the vertical asymptote at
$v = c$ say about accelerating a massive body?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:special-relativity:8}
A light clock has mirrors $L = \qty{1.5}{m}$ apart. Compute its tick at
rest, then its tick as measured from the ground when it cruises at $0.8c$.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:special-relativity:9}
Who measures the \omterm{def:g12:special-relativity:proper-time}{proper time} between: (a) two ticks of a wristwatch --- the
wearer, or a passer-by? (b) birth and decay of a muon --- the muon, or the
lab? (c) leaving Earth and reaching Mars --- pilot, or mission control?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:special-relativity:10}
The space station \omterm{def:g10:universal-gravitation:satellite}{orbits} at \qty{7.7}{km/s}. Compute
$\gamma - 1 \approx v^2/2c^2$, then by how much an astronaut's clock trails
a ground clock after six months (\qty{1.58e7}{s}).
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g12:special-relativity:11}
Compute the \omterm{def:g10:energy-conservation:energy}{energy} content $E = mc^2$ of \qty{1.0}{g} of matter. Compare it
with the daily output of a \qty{1.0}{GW} \omterm{def:g11:work-of-force:power}{power} station, and with the
\qty{8.5e5}{J} \omterm{def:g11:mechanical-energy:kinetic}{kinetic energy} of a car at \qty{130}{km/h}.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g12:special-relativity:12}
Show from $\gamma = 1/\sqrt{1 - v^2/c^2}$ that $v = c\sqrt{1 - 1/\gamma^2}$;
deduce the speed of the $\gamma = 30$ muons, to five \omterm{def:g10:orders-of-magnitude:sigfig}{significant figures} in
\omterm{def:g10:orders-of-magnitude:unit}{units} of $c$.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g12:special-relativity:13}
Retell the muon's journey from its own frame: how thick is the atmosphere
at $\gamma = 30$, how long does it sweep past at $0.99944c$, and does that
fit within one proper lifetime?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g12:special-relativity:14}
A linear accelerator pushes electrons to $\gamma = \num{1.0e5}$ along a
\qty{3.2}{km} tube. How long does the trip take in the lab? How long in the
electron's frame, and how ``long'' is the tube there?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g12:special-relativity:15}
Proxima Centauri lies $4.2$ \omterm{def:g10:orders-of-magnitude:lightyear}{light-years} away (a \omterm{def:g10:orders-of-magnitude:lightyear}{light-year} is the distance
light covers in a year). A probe cruises there at $0.9c$: how long does the
trip take on Earth's calendar, and on the probe's clock?
\end{exercise}

\section{Problem: The Muon and the Twin}

\begin{problem}[{Weekend problem --- the muon's journey and the astronaut's
twin: a particle outliving its clock, a light clock rebuilt, a traveler
younger than her twin, and the bookkeeping that keeps GPS honest}]
\label{pb:g12:special-relativity:1}
Data: muon proper lifetime $\Delta t_0 = \qty{2.2}{\micro s}$; muons born
about \qty{15}{km} up at nearly $c$; sea-level flux about one muon per
\unit{cm^2} per minute; $c = \qty{3.00e8}{m/s}$.

\medskip
\noindent\textbf{Part I --- The muon's impossible arrival.}
\begin{enumerate}
  \item Without relativity, how far does a muon travel in one lifetime?
  \item How many lifetimes does the \qty{15}{km} descent then take?
  \item After $n$ lifetimes a fraction $\mathrm{e}^{-n}$ survives:
        estimate it, confront the measured flux, and pass verdict on
        classical physics in one sentence.
  \item These muons in fact arrive with $\gamma \approx 30$. Compute their
        lifetime as measured from the ground and the resulting \omterm{def:g12:projectile-motion:range}{range}.
  \item Compute their speed, using $v = c\sqrt{1 - 1/\gamma^2}$, to five
        \omterm{def:g10:orders-of-magnitude:sigfig}{significant figures} in \omterm{def:g10:orders-of-magnitude:unit}{units} of $c$.
\end{enumerate}

\medskip
\noindent\textbf{Part II --- The light clock, rebuilt.}
A light clock ($L = \qty{1.5}{m}$) rides a ship at $v = 0.6c$.
\begin{enumerate}[resume]
  \item Compute the tick $\Delta t_0$ read on board.
  \item Seen from the ground, a tick lasts $\Delta t$: give the horizontal
        advance per half-tick, the \omterm{def:g12:quantum-world:photon}{photon}'s vertical climb, and --- naming
        the postulate that fixes it --- the slanted half-path's length.
  \item Apply Pythagoras to the half-tick triangle and solve for
        $\Delta t$, showing that $\Delta t = \gamma\,\Delta t_0$.
  \item Compute $\gamma$ and $\Delta t$ numerically.
  \item Which observer measures the \omterm{def:g12:special-relativity:proper-time}{proper time} between two ticks? Justify
        from the definition.
\end{enumerate}

\medskip
\noindent\textbf{Part III --- The traveler and her twin.}
An astronaut leaves her twin on Earth for a round trip at $0.8c$; Earth
clocks time the whole trip at \qty{10.0}{years}.
\begin{enumerate}[resume]
  \item Compute $\gamma$ at $0.8c$ (exact fraction, then decimal).
  \item How much time elapses on the astronaut's clock?
  \item How far out (Earth frame, in \omterm{def:g10:orders-of-magnitude:lightyear}{light-years}) did the turnaround lie?
  \item What is the age difference on reunion, and which twin is younger?
  \item ``But motion is relative --- the astronaut could claim Earth did
        the traveling!'' Resolve the objection in one sentence.
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- The GPS bill.}
A GPS \omterm{def:g10:universal-gravitation:satellite}{satellite} \omterm{def:g10:universal-gravitation:satellite}{orbits} at radius $r = \qty{2.66e7}{m}$, \omterm{def:g12:oscillators-and-time:oscillator}{period} $T = \qty{12}{h}$.
\begin{enumerate}[resume]
  \item Compute its orbital speed $v = 2\pi r/T$.
  \item Compute $\gamma - 1 \approx v^2/2c^2$.
  \item Deduce, in microseconds, how far the \omterm{def:g10:universal-gravitation:satellite}{satellite} clock falls behind
        ground clocks per day (\qty{86400}{s}).
  \item If this drift alone went uncorrected, positions err by
        $c \times$ (drift): compute the error after one day.
  \item Gravity, weaker at altitude, runs the clock \emph{fast} by
        \qty{45}{\micro s} per day (general relativity): compute the net
        drift and the daily position error in kilometres --- the number
        that obliges engineers to detune every GPS clock.
\end{enumerate}
\end{problem}

\bigskip
\noindent
Here ends the physics of school --- not the physics. In the university
volumes, mechanics returns armed with calculus; electricity and magnetism
reveal themselves as one structure in which light, and relativity itself,
was hiding all along; the quantum world, glimpsed last chapter, becomes a
working theory of matter. Each theory will be obliged, like relativity, to
hand back what you now know as a limiting case. Carry it forward: it is
the part that will not change.
