\chapter{Light Spectra and the Message of Light}\label{ch:g10:light-spectra}

A beam of starlight is the only sample of a star we will ever hold, and it
turns out to be a generous one. Spread the beam into its spectrum --- a
glass prism does it, a curtain of raindrops does it --- and the light
becomes a message: a continuous rainbow background announces the star's
temperature, and a barcode of fine lines spells out, element by element,
what the star is made of. This chapter learns to read both parts of the
message --- on laboratory lamps first, then on the Sun.

\section{White light and its spectrum}

\begin{definition}[Decomposition of white light]\label{def:g10:light-spectra:white}
When a narrow beam of sunlight crosses a glass prism, it emerges spread
into a continuous band of colors, from red to violet. This band is the
\emph{spectrum}\index{spectrum} of the light. Light that contains all the
visible colors, like sunlight or the light of an ordinary bulb, is called
\emph{white light}\index{white light}. Light of a single pure color, which
no prism can split any further, is
\emph{monochromatic}\index{monochromatic light} (a laser beam, for
instance); a mixture of several colors is
\emph{polychromatic}\index{polychromatic light}. An instrument that
performs the decomposition and reads off the composition of a light is a
\emph{spectroscope}\index{spectroscope}.
\end{definition}

\begin{omfigure}
\begin{tikzpicture}[scale=1.0, font=\small]
  \draw[thick, fill=blue!6] (0,0) -- (1.4,2.6) -- (2.8,0) -- cycle;
  \draw[line width=1.6pt] (-2.4,1.95) -- (0.75,1.4)
    node[pos=0.22, above] {white light};
  \draw[gray] (0.75,1.4) -- (2.05,1.05);
  \draw[red, thick]              (2.05,1.05) -- (5.4,0.30);
  \draw[orange, thick]           (2.05,1.05) -- (5.36,0.08);
  \draw[yellow!75!black, thick]  (2.05,1.05) -- (5.31,-0.14);
  \draw[green!65!black, thick]   (2.05,1.05) -- (5.25,-0.36);
  \draw[blue, thick]             (2.05,1.05) -- (5.17,-0.58);
  \draw[violet, thick]           (2.05,1.05) -- (5.08,-0.80);
  \node[right] at (5.4,0.30)  {red (deviated least)};
  \node[right] at (5.08,-0.80) {violet (deviated most)};
\end{tikzpicture}

{\small A prism fans a beam of white light out into its spectrum; on a
screen, the fan paints a continuous rainbow band.}
\end{omfigure}

\begin{remark}[Prisms and raindrops]\label{rem:g10:light-spectra:rainbow}
The prism works because glass bends violet light a little more than red
light --- \emph{why} it does so belongs to the study of refraction, in
\cref{ch:g10:refraction}; here we only use the effect. Water does the same
job: each raindrop of a passing shower acts as a tiny prism, receiving
white sunlight and returning it sorted by color. A rainbow is the spectrum
of the Sun, drawn across the sky.
\end{remark}

\begin{definition}[Wavelength, the identity card of a radiation]\label{def:g10:light-spectra:wavelength}
Each monochromatic radiation is characterized by one number, its
\emph{wavelength}\index{wavelength} $\lambda$, measured in nanometers
($\qty{1}{nm} = \qty{e-9}{m}$). The wavelength is the identity card of the
radiation: give $\lambda$ and you have said everything there is to say
about the pure color. The eye responds to wavelengths between about
$\qty{400}{nm}$ (violet) and $\qty{800}{nm}$ (red) --- the \emph{visible
spectrum}\index{visible spectrum}. Beyond the two ends the radiation
continues, invisible to us: \emph{ultraviolet}\index{ultraviolet} below
$\qty{400}{nm}$, \emph{infrared}\index{infrared} above $\qty{800}{nm}$.
What the wavelength actually measures --- the length of one ripple of a
light wave --- is explained in \cref{ch:g10:signals-and-waves}; in this
chapter, the number itself is what matters.
\end{definition}

\begin{example}[Reading wavelengths]\label{ex:g10:light-spectra:classify}
A green laser pointer emits at $\qty{532}{nm}$: inside the visible range,
in the green. A germicidal mercury lamp emits at $\qty{254}{nm}$:
ultraviolet, invisible (and dangerous to the eye precisely because
invisible). The diode of a television remote emits at $\qty{940}{nm}$:
infrared --- point it at a phone camera, whose sensor sees a little beyond
$\qty{800}{nm}$, and the invisible flashes appear on the screen.
\end{example}

\section{Continuous spectra and temperature}

\begin{definition}[Continuous spectrum]\label{def:g10:light-spectra:continuous}
Dense, hot matter --- the filament of a bulb, molten steel, the surface of
a star --- glows, and its light contains \emph{every} wavelength over a
wide range: an unbroken band of color with nothing missing. Such a
spectrum is a \emph{continuous spectrum}\index{continuous spectrum}. It
carries no trace of the chemical nature of the body: a white-hot iron bar
and a white-hot copper bar glow alike. What it does encode --- entirely
--- is the temperature.
\end{definition}

\begin{definition}[Absolute temperature]\label{def:g10:light-spectra:kelvin}
For radiation laws, temperature is counted not from the freezing point of
water but from the coldest state possible,
$\qty{-273}{\degreeCelsius}$. The unit is the \emph{kelvin}\index{kelvin}
(symbol $\unit{K}$), with the same size of degree:
\[
T \text{ (in } \unit{K}) = \theta \text{ (in } \unit{\degreeCelsius}) + 273 .
\]
Thus a room at $\qty{20}{\degreeCelsius}$ is at $\qty{293}{K}$, and the
surface of the Sun, near $\qty{5500}{\degreeCelsius}$, is near
$\qty{5800}{K}$.
\end{definition}

\begin{proposition}[Hotter means brighter and bluer (Wien's law)]\label{prop:g10:light-spectra:wien}
When the temperature of a dense glowing body rises,
\begin{enumerate}
  \item it radiates more at \emph{every} wavelength: the whole spectrum
        brightens;
  \item the wavelength $\lambda_{\max}$ at which it radiates most
        strongly slides toward the short-wavelength (blue) side,
        according to
        \[
        \lambda_{\max} \times T = \qty{2.90e-3}{m.K},
        \]
        where $T$ is the absolute temperature in kelvins.
\end{enumerate}
\end{proposition}
\admitted

\begin{remark}[Where this law comes from]\label{rem:g10:light-spectra:wienorigin}
Wien's law summarizes careful measurements on furnaces and filaments made
at the end of the nineteenth century. Deriving it requires the quantum
theory of radiation --- this spectrum is, in fact, precisely what
classical physics could \emph{not} explain, and the puzzle that forced
Planck to invent the quantum. The honest derivation is carried out in the
Year 3 volume.
\end{remark}

\begin{omfigure}
\begin{tikzpicture}[scale=1.0, font=\small]
  \foreach \y in {0, 1.1, 2.2}
    \foreach \xa/\xb/\c in {0/0.6/violet, 0.6/1.35/blue,
        1.35/2.4/green!75!black, 2.4/2.85/yellow, 2.85/3.45/orange,
        3.45/6/red}
      \fill[\c] (\xa,\y) rectangle (\xb,\y+0.55);
  \foreach \xa/\xb/\o in {0/0.6/0.85, 0.6/1.35/0.7, 1.35/2.4/0.5,
      2.4/2.85/0.35, 2.85/3.45/0.25, 3.45/6/0.1}
    \fill[black, opacity=\o] (\xa,0) rectangle (\xb,0.55);
  \foreach \xa/\xb/\o in {0/0.6/0.5, 0.6/1.35/0.35, 1.35/2.4/0.2,
      2.4/2.85/0.12, 2.85/3.45/0.08, 3.45/6/0.08}
    \fill[black, opacity=\o] (\xa,1.1) rectangle (\xb,1.65);
  \node[left] at (-0.15, 0.275) {$\qty{3000}{K}$};
  \node[left] at (-0.15, 1.375) {$\qty{4500}{K}$};
  \node[left] at (-0.15, 2.475) {$\qty{6000}{K}$};
  \draw[-Stealth] (0,-0.5) -- (6.7,-0.5)
    node[right] {$\lambda$ (\unit{nm})};
  \foreach \x/\l in {0/400, 1.5/500, 3/600, 4.5/700, 6/800}
    \draw (\x,-0.42) -- (\x,-0.58) node[below] {\l};
\end{tikzpicture}

{\small The visible band of a continuous spectrum at three temperatures:
the cooler the source, the dimmer the band --- and the blue end fades
first.}
\end{omfigure}

\begin{example}[Taking the Sun's temperature]\label{ex:g10:light-spectra:suntemp}
The Sun's continuous spectrum is most intense near
$\lambda_{\max} = \qty{500}{nm}$. Wien's law gives its surface
temperature:
\[
T = \frac{\qty{2.90e-3}{m.K}}{\qty{5.00e-7}{m}} = \qty{5800}{K},
\]
about $\qty{5500}{\degreeCelsius}$ --- measured from $150$ million
kilometers away, with a prism and a division.
\end{example}

\begin{omfigure}
\begin{tikzpicture}
\begin{axis}[omaxis, width=9.4cm, height=5.6cm,
    xlabel={$\lambda$ (\unit{nm})}, ylabel={intensity (arbitrary units)},
    xmin=0, xmax=2150, ymin=0, ymax=9.5,
    xtick={400, 800, 1200, 1600, 2000}, ytick=\empty]
  \fill[gray, opacity=0.12] (axis cs:400,0) rectangle (axis cs:800,9.5);
  \node[gray, anchor=south] at (axis cs:600,8.4) {visible};
  \addplot[omThm, thick, domain=180:2050, samples=160]
    {3e16 * x^(-5) / (exp(2400/x) - 1)};
  \addplot[omDef, thick, domain=180:2050, samples=160]
    {3e16 * x^(-5) / (exp(3600/x) - 1)};
  \draw[dashed, omThm] (axis cs:480,0) -- (axis cs:480,8.0);
  \draw[dashed, omDef] (axis cs:725,0) -- (axis cs:725,1.05);
  \node[omThm, anchor=south west] at (axis cs:620,6.4)
    {hot star, $\qty{6000}{K}$};
  \node[omDef, anchor=south west] at (axis cs:950,1.3)
    {cool star, $\qty{4000}{K}$};
\end{axis}
\end{tikzpicture}

{\small Intensity against wavelength for two stars: the hotter one
radiates more at every wavelength, and its peak $\lambda_{\max}$ (dashed)
sits at a shorter wavelength.}
\end{omfigure}

\section{Line spectra of excited gases}

\begin{definition}[Emission line spectrum]\label{def:g10:light-spectra:emission}
A gas at low pressure, excited by an electric discharge (as in a neon tube
or a sodium street lamp), glows too --- but through a spectroscope its
light is nothing like a rainbow. Only a few isolated wavelengths are
present: thin bright lines on a black background, called \emph{spectral
lines}\index{spectral line}. Such a spectrum is an \emph{emission line
spectrum}\index{emission spectrum}\index{line spectrum}.
\end{definition}

\begin{example}[Hydrogen and sodium]\label{ex:g10:light-spectra:hydrogen}
Excited hydrogen emits exactly four lines in the visible: a red line at
$\qty{656}{nm}$, a blue-green line at $\qty{486}{nm}$, and two violet
lines at $\qty{434}{nm}$ and $\qty{410}{nm}$. Excited sodium emits
essentially one strong yellow-orange line at $\qty{589}{nm}$ --- which is
why older street lamps bathe whole neighborhoods in that single color:
their light contains almost nothing else.
\end{example}

\begin{proposition}[The fingerprint of the elements]\label{prop:g10:light-spectra:fingerprint}
Each chemical element emits its own fixed catalog of spectral lines, the
same in every laboratory and at every epoch, and no two elements share
the same catalog. A line spectrum therefore identifies the emitting
element, as a fingerprint identifies a person.
\end{proposition}
\admitted

\begin{remark}[Why lines?]\label{rem:g10:light-spectra:whylines}
Why an atom emits only certain wavelengths --- and why those wavelengths
betray the element --- is explained by the quantum theory of the atom, in
the final year of this volume: each element owns a ladder of energy
levels, and each line marks one jump between rungs. For now we use the
fingerprint as detectives do: without asking the fingers how they grew.
\end{remark}

\section{Absorption spectra}

\begin{definition}[Absorption spectrum]\label{def:g10:light-spectra:absorption}
Send white light \emph{through} a cool, low-pressure gas and disperse
what comes out: the continuous rainbow reappears, but scored by thin dark
lines. The gas has removed certain wavelengths from the passing light.
Such a spectrum --- a continuous background minus a set of lines --- is
an \emph{absorption spectrum}\index{absorption spectrum}.
\end{definition}

\begin{proposition}[Emission and absorption lines coincide]\label{prop:g10:light-spectra:kirchhoff}
A gas absorbs exactly the wavelengths that it is able to emit: the dark
lines of an element's absorption spectrum sit at the same wavelengths as
the bright lines of its emission spectrum. The fingerprint of
\cref{prop:g10:light-spectra:fingerprint} can therefore be read in the
dark as well as in the bright.
\end{proposition}
\admitted

\begin{remark}[One mechanism, two signs]\label{rem:g10:light-spectra:mechanism}
The coincidence is no accident: absorbing at $\lambda$ climbs the same
energy rung that emitting at $\lambda$ descends --- the quantum chapters
at the end of this volume make this precise. Established by Kirchhoff and
Bunsen in 1859, it is the key that unlocks the entire next section.
\end{remark}

\begin{omfigure}
\begin{tikzpicture}[scale=1.0, font=\small]
  \fill[black] (0,1.5) rectangle (6,2.1);
  \draw[violet, line width=1.6pt] (0.15,1.5) -- (0.15,2.1);
  \draw[blue, line width=1.6pt]   (0.51,1.5) -- (0.51,2.1);
  \draw[cyan, line width=1.6pt]   (1.29,1.5) -- (1.29,2.1);
  \draw[red, line width=1.6pt]    (3.84,1.5) -- (3.84,2.1);
  \node[left] at (-0.15,1.8) {emission};
  \foreach \xa/\xb/\c in {0/0.6/violet, 0.6/1.35/blue,
      1.35/2.4/green!75!black, 2.4/2.85/yellow, 2.85/3.45/orange,
      3.45/6/red}
    \fill[\c] (\xa,0) rectangle (\xb,0.6);
  \foreach \x in {0.15, 0.51, 1.29, 3.84}
    \draw[black, line width=1.6pt] (\x,0) -- (\x,0.6);
  \node[left] at (-0.15,0.3) {absorption};
  \foreach \x in {0.15, 0.51, 1.29, 3.84}
    \draw[dashed, gray] (\x,0.6) -- (\x,1.5);
  \draw[-Stealth] (0,-0.5) -- (6.7,-0.5)
    node[right] {$\lambda$ (\unit{nm})};
  \foreach \x/\l in {0/400, 1.5/500, 3/600, 4.5/700, 6/800}
    \draw (\x,-0.42) -- (\x,-0.58) node[below] {\l};
\end{tikzpicture}

{\small The two spectra of hydrogen: bright emission lines (top) and dark
absorption lines (bottom) at exactly the same wavelengths, $410$, $434$,
$486$ and $\qty{656}{nm}$.}
\end{omfigure}

\begin{example}[A flask of sodium vapor]\label{ex:g10:light-spectra:sodiumflask}
White light crosses a flask containing cool sodium vapor. The transmitted
spectrum is a full rainbow with one dark line at $\qty{589}{nm}$ ---
precisely where a sodium lamp shines brightest. Looking at the flask from
the side in a dark room, one sees the stolen light re-emitted: a faint
yellow-orange gleam at that same wavelength.
\end{example}

\section{Spectral analysis: reading a star}

\begin{definition}[Spectral analysis]\label{def:g10:light-spectra:analysis}
\emph{Spectral analysis}\index{spectral analysis} is the identification
of the chemical elements contained in a source (a flame, a lamp, a star)
by matching the lines of its spectrum against the laboratory catalogs of
the elements.
\end{definition}

\begin{method}[Reading a stellar spectrum]\label{met:g10:light-spectra:reading}
A star is a ball of dense hot matter (the surface) wrapped in a thinner,
cooler atmosphere. Its light therefore arrives as a continuous spectrum
scored by absorption lines --- and each part is read separately:
\begin{enumerate}
  \item Locate the peak $\lambda_{\max}$ of the continuous background;
        Wien's law (\cref{prop:g10:light-spectra:wien}) gives the surface
        temperature $T = \qty{2.90e-3}{m.K} / \lambda_{\max}$.
  \item List the wavelengths of the dark lines.
  \item Match them against the catalogs: an element may be declared
        present only if \emph{all} its strong lines appear.
  \item Conclude: the absorption happens in the star's atmosphere, so the
        elements identified are those of the atmosphere.
\end{enumerate}
\end{method}

\begin{example}[The Sun's recipe]\label{ex:g10:light-spectra:sunrecipe}
The solar spectrum carries thousands of dark lines, mapped by Fraunhofer
in 1814. Among the strongest: $410$, $434$, $486$ and $\qty{656}{nm}$ ---
the full hydrogen fingerprint --- together with the sodium line at
$\qty{589}{nm}$ and a calcium pair at $393$ and $\qty{397}{nm}$. The
Sun's atmosphere contains hydrogen above all, plus traces of elements
familiar from Earth's rocks. Modern measurements refine the headline: the
Sun is roughly three quarters hydrogen by mass.
\end{example}

\begin{remark}[The element found in the sky first]\label{rem:g10:light-spectra:helium}
During the eclipse of 1868, a bright line at $\qty{587.6}{nm}$ appeared
in the spectrum of the Sun's atmosphere --- matching no known element. By
\cref{prop:g10:light-spectra:fingerprint} an unknown fingerprint means an
unknown element, so one was announced and named after the Greek sun,
\emph{helios}: helium. It was finally isolated on Earth in 1895 --- the
only chemical element discovered in space before being found at home.
\end{remark}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:g10:light-spectra:1}
For each radiation, say whether it is ultraviolet, visible or infrared,
and give its color if visible: $\qty{254}{nm}$ (mercury lamp),
$\qty{480}{nm}$, $\qty{589}{nm}$, $\qty{656}{nm}$, $\qty{1064}{nm}$
(cutting laser).
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g10:light-spectra:2}
A narrow beam of sunlight falls on a glass prism.
\begin{enumerate}
  \item Describe what appears on a screen placed behind the prism.
  \item Which end of the band has been deviated the most?
  \item The sunlight is replaced by the beam of a green laser pointer.
        What appears on the screen now, and what does this show about
        laser light?
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g10:light-spectra:3}
Match each source to the type of spectrum it produces (continuous, line
emission, or absorption): (a) the filament of an incandescent bulb;
(b) a neon advertising tube; (c) molten iron in a foundry; (d) a sodium
street lamp; (e) sunlight received on Earth.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g10:light-spectra:4}
About rainbows:
\begin{enumerate}
  \item What plays the role of the prism?
  \item A rainbow is only seen with the Sun behind the observer and rain
        ahead. What does each ingredient supply?
  \item Explain why a rainbow proves that sunlight is white light.
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g10:light-spectra:5}
A blacksmith heats an iron nail: it glows dull red, then bright orange,
then almost white.
\begin{enumerate}
  \item Rank the three stages by temperature.
  \item Which two effects of \cref{prop:g10:light-spectra:wien} does the
        sequence illustrate?
  \item Justify the expression ``white-hot is hotter than red-hot''.
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:light-spectra:6}
The Sun's continuous spectrum peaks at $\lambda_{\max} = \qty{500}{nm}$.
\begin{enumerate}
  \item Compute the surface temperature of the Sun in kelvins.
  \item Convert it to degrees Celsius.
  \item Check that the peak lies inside the visible range, and state its
        color.
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:light-spectra:7}
A candle flame glows at about $\qty{1800}{K}$; an electric hotplate set
to ``warm'' sits at about $\qty{500}{K}$.
\begin{enumerate}
  \item Compute $\lambda_{\max}$ for each source, and give its domain.
  \item The hotplate is clearly hot --- a hand held above it feels the
        radiation --- yet it looks black. Explain.
  \item The candle's peak is also outside the visible range. Why can we
        see the flame at all?
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:light-spectra:8}
Laboratory catalogs give, in the visible: hydrogen
$\{410, 434, 486, 656\}\,\unit{nm}$; sodium $\{589\}\,\unit{nm}$; helium
$\{447, 502, 588, 668\}\,\unit{nm}$. Two discharge lamps are analyzed.
Lamp A shows lines at $434$, $486$ and $\qty{656}{nm}$, plus a faint
violet line near the edge of visibility; lamp B shows a single strong
line at $\qty{589}{nm}$. Identify the gas in each lamp, justifying
carefully why lamp B does not contain helium.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:light-spectra:9}
White light is sent through a flask of cool sodium vapor.
\begin{enumerate}
  \item Describe the spectrum of the transmitted light.
  \item At which wavelength does its dark line sit, and why exactly
        there?
  \item The lamp is switched off, the room darkened, and the vapor is
        now excited by a discharge. Describe the new spectrum.
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:light-spectra:10}
A high-resolution spectrum of the Sun shows dark lines at $410$, $434$,
$486$, $517$, $518$ and $\qty{656}{nm}$. Catalogs: hydrogen
$\{410, 434, 486, 656\}\,\unit{nm}$; magnesium $\{517, 518\}\,\unit{nm}$.
\begin{enumerate}
  \item Which elements do these lines reveal?
  \item In which part of the Sun are the lines produced, and why are
        they dark rather than bright?
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:light-spectra:11}
Your body has a surface temperature of about $\qty{37}{\degreeCelsius}$.
\begin{enumerate}
  \item Convert to kelvins and compute the $\lambda_{\max}$ of your own
        thermal radiation.
  \item In which domain does it fall?
  \item Deduce why nobody glows visibly in a dark room, and how a
        thermal camera finds people anyway.
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g10:light-spectra:12}
The tungsten filament of an incandescent bulb runs at $\qty{2700}{K}$.
\begin{enumerate}
  \item Compute $\lambda_{\max}$ and give its domain.
  \item Explain why such bulbs make poor lamps but excellent little
        heaters.
  \item What filament temperature would place $\lambda_{\max}$ at
        $\qty{550}{nm}$, in mid-visible? Tungsten melts at
        $\qty{3700}{K}$; conclude why no filament bulb can imitate
        daylight, and why lighting moved to other technologies.
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g10:light-spectra:13}
The spectrum of a star shows a continuous background peaking at
$\qty{420}{nm}$, scored by dark lines at $393$, $397$, $410$, $434$,
$486$ and $\qty{656}{nm}$. Catalogs: hydrogen
$\{410, 434, 486, 656\}\,\unit{nm}$; calcium $\{393, 397\}\,\unit{nm}$;
sodium $\{589\}\,\unit{nm}$.
\begin{enumerate}
  \item Compute the star's surface temperature. Is it hotter or cooler
        than the Sun?
  \item Which elements does its atmosphere certainly contain?
  \item Is sodium abundant in this atmosphere? Justify.
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g10:light-spectra:14}
A discharge lamp contains a mixture of two gases. Its spectrum shows
lines at $410$, $434$, $447$, $486$, $502$, $588$, $656$ and
$\qty{668}{nm}$. Using the catalogs of \cref{exo:g10:light-spectra:8}:
\begin{enumerate}
  \item Identify the two gases.
  \item A classmate claims the $\qty{588}{nm}$ line proves the lamp
        also contains sodium (line at $\qty{589}{nm}$), since a cheap
        spectroscope cannot separate wavelengths $\qty{1}{nm}$ apart.
        Without a better instrument, why is helium a sufficient
        explanation of this line --- and what measurement would settle
        the question of sodium for good?
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g10:light-spectra:15}
In the constellation Orion, Betelgeuse glows visibly red and Rigel
blue-white. Their continuous spectra peak at about $\qty{850}{nm}$ and
$\qty{240}{nm}$ respectively.
\begin{enumerate}
  \item Compute both surface temperatures.
  \item Recover the ratio of the two temperatures directly from the two
        wavelengths, without recomputing either temperature.
  \item Explain the two colors, given that \emph{neither} peak lies in
        the visible range.
  \item Both stars are bright to the naked eye. Why does a peak outside
        the visible range not make a star invisible?
\end{enumerate}
\end{exercise}

\section{Problem: Reading starlight}

\begin{problem}[{Weekend problem --- reading starlight: the philosopher
who said ``never'', and how a prism takes a star's temperature and reads
its chemical recipe}]
\label{pb:g10:light-spectra:1}
In 1835 the philosopher Auguste Comte needed an example of knowledge
forever beyond human reach, and chose it carefully: the chemical
composition of the stars. Nobody would ever bottle a star. In 1859
Kirchhoff and Bunsen matched the dark lines of the Sun to the bright
lines of laboratory flames, and the unreachable knowledge arrived by
return of light. This problem retraces the whole path: laboratory
fingerprints, glowing bodies, the Sun's dark lines --- and finally a
star whose temperature and composition you will determine from your
desk.

\medskip
\noindent\textbf{Part I --- Fingerprints on file.}
\begin{enumerate}
  \item Give the visible range of wavelengths in nanometers and the
        color at each end. Where does $\qty{550}{nm}$ sit?
  \item A hydrogen discharge lamp shows lines at $410$, $434$, $486$
        and $\qty{656}{nm}$. Give the color of each line.
  \item A sodium lamp emits essentially one line at $\qty{589}{nm}$.
        What does the lamp look like to the eye, and why is it hopeless
        to judge the color of a parked car under such a street light?
  \item Explain why a set of spectral lines identifies an element ---
        and why \emph{all} the strong lines of an element must be found
        before declaring it present.
  \item Predict the spectrum seen (a) when white light crosses a flask
        of cool sodium vapor, and (b) when the same vapor, alone in a
        dark room, is excited by a discharge.
\end{enumerate}

\medskip
\noindent\textbf{Part II --- Hot bodies and their colors.}
\begin{enumerate}[resume]
  \item State the two effects of raising the temperature of a dense
        glowing body on its continuous spectrum.
  \item A halogen filament runs at $\qty{3400}{K}$. Compute
        $\lambda_{\max}$ and give its domain.
  \item The Sun's continuous spectrum peaks at $\qty{500}{nm}$. Compute
        its surface temperature.
  \item That peak sits in the green --- yet the Sun does not look
        green. Explain.
  \item A poker left in the forge is first felt from a distance without
        being seen to glow, then glows dull red, then orange-white.
        Explain the whole sequence with question 6.
\end{enumerate}

\medskip
\noindent\textbf{Part III --- The Sun's dark lines.}
\begin{enumerate}[resume]
  \item The solar spectrum is a continuous band crossed by thousands of
        fine dark lines. Which part of the Sun produces the continuous
        background, and which part produces the lines?
  \item Strong solar lines sit at $393$, $397$, $410$, $434$, $486$,
        $589$ and $\qty{656}{nm}$. Catalogs: hydrogen
        $\{410, 434, 486, 656\}\,\unit{nm}$; sodium $\{589\}\,\unit{nm}$;
        calcium $\{393, 397\}\,\unit{nm}$. Which elements are present in
        the Sun's atmosphere?
  \item The hydrogen lines are by far the strongest. What does this
        suggest about the Sun's composition --- and what do modern
        measurements say?
  \item In 1868, a line at $\qty{587.6}{nm}$ observed in the Sun's
        atmosphere matched no catalog. Reconstruct the reasoning that
        justified announcing a new element, and say how the story ended.
  \item Why are the solar lines dark, when the same atoms in a
        laboratory discharge give bright lines? During a total eclipse,
        for a few seconds, only the thin atmosphere of the Sun stays
        visible against a dark sky --- and the same lines flash
        \emph{bright}. Explain.
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- A star on the desk.}
The spectrum of the bright star Vega is placed before you: a continuous
background peaking at $\qty{290}{nm}$, scored by strong dark lines at
$410$, $434$, $486$ and $\qty{656}{nm}$.
\begin{enumerate}[resume]
  \item Compute Vega's surface temperature.
  \item The peak lies in the ultraviolet. Predict the star's apparent
        color, and reconcile ``peak outside the visible'' with ``bright
        star''.
  \item Identify the element dominating Vega's atmosphere.
  \item Compare Vega with the Sun: temperature ratio, color, and
        brightness per unit of glowing surface.
  \item Finale --- fill in the identity form that Comte declared
        impossible: temperature, dominant element, apparent color of
        Vega, all read from one beam of light. How many years did
        ``never'' last?
\end{enumerate}
\end{problem}
