\chapter{Pressure: From Sport to Diving}\label{ch:g10:pressure}

A skier glides over snow that a walker sinks into; a \qty{5}{N} push
drives a needle through leather a fist cannot dent. Same forces,
wildly different effects: the missing quantity is \emph{pressure},
force divided by the area receiving it. This chapter defines it,
follows it down a pool, up a barometer, into a squeezed syringe ---
and ends forty metres under the sea, where it turns vital.

\section{Pressure: force spread over an area}

\begin{definition}[Pressure]\label{def:g10:pressure:pressure}
When a force of magnitude $F$ (in \unit{N}) presses perpendicularly on a
surface of area $S$ (in \unit{m^2}), the \emph{pressure}\index{pressure}
exerted on that surface is
\[
P = \frac{F}{S}.
\]
Its unit, the \unit{N/m^2}, is the \emph{pascal}\index{pascal}
(\unit{Pa}) --- a tiny unit, an apple's weight spread over a square
metre, so \unit{kPa} and \unit{MPa} are the working units.
\end{definition}

\begin{example}[Heel versus snowshoe]\label{ex:g10:pressure:snowshoe}
A \qty{60}{kg} person weighs $F = 60 \times 9.81 \approx \qty{590}{N}$.
On one stiletto heel of \qty{1.0}{cm^2} = \qty{1.0e-4}{m^2}:
$P = 590/\num{1.0e-4} \approx \qty{5.9e6}{Pa} = \qty{5.9}{MPa}$. On two
snowshoes of \qty{0.20}{m^2} each:
$P = 590/0.40 \approx \qty{1.5}{kPa}$. The \emph{same} weight presses
$4000$ times harder under the heel: heels pockmark wooden floors,
snowshoes float on powder snow.
\end{example}

\begin{omfigure}
\begin{tikzpicture}[scale=1.0, font=\small]
  \draw[thick] (-1.3,0) -- (1.3,0);
  \draw[-Stealth, omThm, very thick] (0,2.1) -- (0,0.6)
    node[midway, right] {$F = \qty{590}{N}$};
  \fill[omDef!30] (-0.09,0) rectangle (0.09,0.55);
  \draw[omDef] (-0.09,0) rectangle (0.09,0.55);
  \node[below] at (0,-0.1) {$S = \qty{1.0}{cm^2}$};
  \node[below] at (0,-0.6) {$P \approx \qty{5.9}{MPa}$};
\end{tikzpicture}\qquad
\begin{tikzpicture}[scale=1.0, font=\small]
  \draw[thick] (-1.7,0) -- (1.7,0);
  \draw[-Stealth, omThm, very thick] (0,2.1) -- (0,0.6)
    node[midway, right] {$F = \qty{590}{N}$};
  \fill[omDef!30] (-1.4,0) rectangle (1.4,0.16);
  \draw[omDef] (-1.4,0) rectangle (1.4,0.16);
  \node[below] at (0,-0.1) {$S = \qty{0.40}{m^2}$};
  \node[below] at (0,-0.6) {$P \approx \qty{1.5}{kPa}$};
\end{tikzpicture}

{\small The same force on two areas: the pressure, not the force, decides whether the surface yields.}
\end{omfigure}

\begin{remark}[Concentrate or spread]\label{rem:g10:pressure:tools}
A sewing needle pushed with \qty{5}{N} on a tip of \qty{0.01}{mm^2} =
\qty{1e-8}{m^2} exerts $\qty{5e8}{Pa} = \qty{500}{MPa}$ --- enough to
part leather fibres. Piercing tools concentrate force; load-bearing
designs (skis, caterpillar tracks, foundations) spread it. Sport is
applied pressure management: skates bite into ice, skis glide on snow.
\end{remark}

\section{Pressure in a liquid at rest}

Why does water press harder lower down? Isolate an imaginary vertical
column of water of cross-section $S$, from the surface to depth $h$:
volume $Sh$, mass $\rho S h$ ($\rho$ the liquid's density, in
\unit{kg/m^3}), weight $\rho S h g$. The bottom of the column supports
that weight \emph{plus} the atmosphere's push $P_0 S$ on the top.
Dividing the total force by $S$: the pressure at depth $h$ should be
$P_0 + \rho g h$ --- and it is, everywhere in the liquid.

\begin{theorem}[Pressure at depth]\label{thm:g10:pressure:depth}
In a liquid of density $\rho$ at rest, open to the air, the pressure at
depth $h$ below the surface is
\[
P = P_0 + \rho g h,
\]
where $P_0$ is the atmospheric pressure at the surface. The pressure
depends only on the depth, not on the container's shape, and at a
given point the liquid presses equally hard in all directions.
\end{theorem}
\admitted

\begin{remark}
The column argument only makes the law plausible; the honest derivation
(any shape, all directions) is in the Year 1 volume. The water's own
term $\rho g h$ is the \emph{gauge} pressure --- what a diver's
pressure gauge displays.
\end{remark}

\begin{example}[Ten metres of water]\label{ex:g10:pressure:tenmetres}
In fresh water ($\rho = \qty{1000}{kg/m^3}$), at $h = \qty{10}{m}$:
$\rho g h = 1000 \times 9.81 \times 10 \approx \qty{98}{kPa}$, so
$P = 101 + 98 \approx \qty{199}{kPa}$. Ten metres of water add almost
exactly one atmosphere: the diver's rule of thumb, \emph{one extra
atmosphere per ten metres}.
\end{example}

\begin{omfigure}
\begin{tikzpicture}[scale=1.0, font=\small]
  \shade[top color=omDef!8, bottom color=omDef!50] (0,0) rectangle (3.4,-3.4);
  \node[above] at (1.7,0.05) {surface: $P_0 = \qty{101}{kPa}$};
  \draw[-Stealth, omThm, thick] (1.7,-1.1) -- ++(1.0,0) node[right] {\qty{199}{kPa}};
  \draw[-Stealth, omThm, thick] (1.7,-2.2) -- ++(1.5,0) node[right] {\qty{297}{kPa}};
  \draw[-Stealth, omThm, thick] (1.7,-3.3) -- ++(2.0,0) node[right] {\qty{395}{kPa}};
  \draw (-0.15,-1.1) -- (0,-1.1) node[left=8pt] {\qty{10}{m}};
  \draw (-0.15,-2.2) -- (0,-2.2) node[left=8pt] {\qty{20}{m}};
  \draw (-0.15,-3.3) -- (0,-3.3) node[left=8pt] {\qty{30}{m}};
\end{tikzpicture}

{\small Fresh water at rest: pressure grows linearly with depth, one extra atmosphere every ten metres.}
\end{omfigure}

\section{The ocean of air}

\begin{definition}[Atmospheric pressure]\label{def:g10:pressure:atmospheric}
We live at the bottom of an ocean of air, whose weight creates the
\emph{atmospheric pressure}\index{atmospheric pressure}, at sea level
\[
P_0 \approx \qty{101}{kPa} = \qty{1.01e5}{Pa}.
\]
On each square centimetre this is $F = \num{1.01e5} \times \num{1e-4}
\approx \qty{10}{N}$: the weight of a one-kilogram mass on every
\unit{cm^2} of your skin, your desk, everything.
\end{definition}

\begin{example}[The invisible load]\label{ex:g10:pressure:sheet}
An A4 sheet of paper (\qty{21.0}{cm} $\times$ \qty{29.7}{cm}, so
$S \approx \qty{0.0624}{m^2}$) receives from the air above it
$F = \num{1.01e5} \times 0.0624 \approx \qty{6.3e3}{N}$ --- the weight
of a small car. Nothing tears: the air below pushes
\emph{up} just as hard. We notice the atmosphere only when one side
loses it --- a suction cup, a straw, an aircraft cabin.
\end{example}

\begin{remark}[Torricelli's barometer]\label{rem:g10:pressure:torricelli}
In 1643 Torricelli inverted a sealed tube of mercury over a mercury
bath. The column fell to the height at which $\rho g h = P_0$:
$h = \num{1.01e5}/(13600 \times 9.81) \approx \qty{0.76}{m}$, leaving
above it the first vacuum ever made. The height tracks the weather and
shrinks with altitude: a barometer weighs the air above you (water
would need \qty{10.3}{m}: \cref{exo:g10:pressure:9}).
\end{remark}

\section{Gases: bombardment and Boyle's law}

A gas too presses on its container: it is a swarm of molecules in
ceaseless random motion, each collision gives the wall a tiny outward
push, and billions of billions of impacts per second blur into the
steady force we measure as gas pressure. Squeeze the gas into half the
volume: the molecules strike twice as often, so the pressure should
double. It does.

\begin{proposition}[Boyle's law]\label{prop:g10:pressure:boyle}
\index{Boyle's law}
For a fixed amount of gas held at constant temperature, pressure and
volume are inversely proportional:
\[
P \, V = \text{constant},
\qquad\text{i.e.}\qquad
P_1 V_1 = P_2 V_2 .
\]
\end{proposition}
\admitted

\begin{remark}
An experimental fact at this level --- the Year 1 volume derives it
from the collision picture. The experiment (a sealed syringe of air,
a pressure sensor, \qty{20}{\degreeCelsius}):
\begin{center}
\begin{tabular}{lccccc}
$V$ (\unit{mL}) & 60 & 50 & 40 & 30 & 20 \\
$P$ (\unit{kPa}) & 100 & 120 & 150 & 200 & 300 \\
$P \times V$ (\unit{kPa.mL}) & 6000 & 6000 & 6000 & 6000 & 6000
\end{tabular}
\end{center}
\end{remark}

\begin{omfigure}
\begin{tikzpicture}
\begin{axis}[omaxis, width=7.6cm, height=5.6cm,
    xlabel={$V$ (\unit{mL})}, ylabel={$P$ (\unit{kPa})},
    xmin=0, xmax=70, ymin=0, ymax=340,
    domain=18.5:68, samples=100]
  \addplot[omThm, thick] {6000/x};
  \addplot[only marks, mark=*, omDef]
    coordinates {(60,100) (50,120) (40,150) (30,200) (20,300)};
\end{axis}
\end{tikzpicture}

{\small The syringe data fall on the hyperbola $P = 6000/V$: halve the volume, double the pressure.}
\end{omfigure}

\begin{example}[A squeezed syringe]\label{ex:g10:pressure:boyle}
Air occupies \qty{4.0}{L} at \qty{100}{kPa}; compressed slowly (so the
temperature stays constant) to \qty{250}{kPa}, it occupies
$V_2 = P_1 V_1 / P_2 = 400/250 = \qty{1.6}{L}$.
\end{example}

\section{Diving: pressure put to work}

\begin{method}[Absolute pressure at depth]\label{met:g10:pressure:dive}
For a diver at depth $h$ in the sea ($\rho \approx \qty{1025}{kg/m^3}$,
so $\rho g h \approx \qty{101}{kPa}$ per \qty{10}{m}), compute
$P = P_0 + \rho g h$ --- and in Boyle's law always use this
\emph{absolute} pressure, never the gauge term alone. A diver sits
under $2$ atmospheres at \qty{10}{m}, $3$ at \qty{20}{m}, $5$ at
\qty{40}{m}.
\end{method}

\begin{omfigure}
\begin{tikzpicture}[scale=1.0, font=\small]
  \shade[top color=omDef!8, bottom color=omDef!55] (0,0) rectangle (2.0,-4.4);
  \draw[thick] (2.0,0) -- (2.0,-4.4);
  \draw[thick, omDef] (-0.4,0) -- (2.4,0);
  \draw (1.85,0) -- (2.15,0) node[right] {\qty{0}{m}: \qty{101}{kPa} ($1$ atm)};
  \draw (1.85,-1.1) -- (2.15,-1.1) node[right] {\qty{10}{m}: \qty{202}{kPa} ($2$ atm)};
  \draw (1.85,-2.2) -- (2.15,-2.2) node[right] {\qty{20}{m}: \qty{302}{kPa} ($3$ atm)};
  \draw (1.85,-3.3) -- (2.15,-3.3) node[right] {\qty{30}{m}: \qty{403}{kPa} ($4$ atm)};
  \draw (1.85,-4.4) -- (2.15,-4.4) node[right] {\qty{40}{m}: \qty{503}{kPa} ($5$ atm)};
  \fill[omThm] (0.9,-2.15) circle (0.09);
  \draw[omThm, thick] (0.9,-2.24) -- (0.9,-2.6) (0.9,-2.35) -- (0.7,-2.5)
    (0.9,-2.35) -- (1.1,-2.5) (0.9,-2.6) -- (0.78,-2.85) (0.9,-2.6) -- (1.02,-2.85);
\end{tikzpicture}

{\small Absolute pressure in the sea: one atmosphere at the surface, plus one more for every ten metres.}
\end{omfigure}

\begin{remark}[Never hold your breath]\label{rem:g10:pressure:breath}
A regulator delivers air at ambient pressure: lungs fill normally at
any depth. But fill \qty{6.0}{L} at \qty{20}{m} ($\qty{302}{kPa}$),
hold that breath and surface: Boyle demands
$6.0 \times 302/101 \approx \qty{18}{L}$ --- three times what a chest
holds, and lungs tear well before that, even on a few metres of ascent.
Hence scuba's first rule: \emph{never hold your breath while ascending}
--- keep breathing and the expanding air simply flows out. A free diver
is safe: her air was taken at the surface, so on the way up it only
re-expands to its original volume.
\end{remark}

\begin{remark}[Decompression]\label{rem:g10:pressure:deco}
Breathing high-pressure air also dissolves extra nitrogen in the blood,
like gas in a capped soda bottle; surface too fast and it fizzes into
bubbles inside the body --- decompression sickness. Hence the slow
ascent (about \qty{10}{m} per minute) and shallow stops, which let the
gas leave quietly through the lungs. The theory waits for the Year 1
volume; the rule of conduct belongs to every diver.
\end{remark}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:g10:pressure:1}
A crate weighing \qty{600}{N} rests on a face of
$\qty{0.50}{m} \times \qty{0.30}{m}$, then on a face of
$\qty{0.30}{m} \times \qty{0.20}{m}$. Compute both pressures on the
ground. Which orientation is better on soft sand?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g10:pressure:2}
Express \qty{0.25}{MPa} in \unit{kPa} and in \unit{Pa}. Which presses
harder: \qty{3.0}{N} on \qty{2.0}{cm^2}, or \qty{60}{N} on
\qty{500}{cm^2}?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g10:pressure:3}
Compute the gauge pressure $\rho g h$ and the absolute pressure at the
bottom of a \qty{3.0}{m} deep swimming pool
($\rho = \qty{1000}{kg/m^3}$, $P_0 = \qty{101}{kPa}$).
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g10:pressure:4}
Air occupies \qty{4.0}{L} at \qty{100}{kPa}, temperature constant.
Its volume at \qty{250}{kPa}? At \qty{50}{kPa}?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g10:pressure:5}
A \qty{65}{kg} skater glides on one blade of contact area
\qty{12}{cm^2}, then stands on soles totalling \qty{360}{cm^2}.
Compute both pressures and their ratio.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:pressure:6}
A thumb pushes a drawing pin with \qty{20}{N}. The head has area
\qty{1.2}{cm^2}, the tip \qty{0.010}{mm^2}. Compute the pressures on
thumb and wall; why is only the wall pierced?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:pressure:7}
A submarine hatch of area \qty{0.50}{m^2} sits at \qty{30}{m} depth in
sea water ($\rho = \qty{1025}{kg/m^3}$); inside, the air is at
atmospheric pressure. Compute the net force holding it shut, and the
mass whose weight equals it.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:pressure:8}
A sensor \qty{5.00}{m} deep in an unknown liquid reads a gauge
pressure of \qty{66.2}{kPa}. Find the density. Is it fresh water?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:pressure:9}
How tall would Torricelli's barometer be with water
($\rho = \qty{1000}{kg/m^3}$) instead of mercury
($\rho = \qty{13600}{kg/m^3}$)? Recover both heights from
$\rho g h = P_0$; why did mercury win?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:pressure:10}
Using the syringe data of \cref{prop:g10:pressure:boyle}
($PV = \qty{6000}{kPa.mL}$), predict the volume at \qty{240}{kPa} and
the pressure at \qty{75}{mL}. What curve is $P$ against $V$? And $P$
against $1/V$?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:pressure:11}
A diver at \qty{20}{m} releases a bubble of volume \qty{0.50}{cm^3}.
What is its volume just below the surface? (Constant temperature; sea
water adds \qty{101}{kPa} per \qty{10}{m}.)
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g10:pressure:12}
A suction cup of area \qty{12}{cm^2} is pressed flat against a ceiling.
\begin{enumerate}
  \item Perfect vacuum inside: what force does the atmosphere exert on
        the cup, and what hanging mass can it hold?
  \item A real cup keeps \qty{20}{kPa} of air inside. What mass now?
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g10:pressure:13}
A scuba tank holds \qty{12}{L} of air at \qty{200}{bar}
(\qty{1}{bar} = \qty{100}{kPa} $\approx 1$ atmosphere); \qty{50}{bar}
must stay as reserve. A diver breathes \qty{15}{L} per minute at
ambient pressure. How many litres of surface-pressure air are usable?
How long does the tank last at \qty{30}{m} (absolute pressure
$\approx \qty{4}{bar}$)? And at \qty{10}{m}?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g10:pressure:14}
A free diver's lungs hold \qty{6.0}{L} at the surface and cannot
shrink below the residual volume \qty{1.5}{L}. At what sea depth
($\rho = \qty{1025}{kg/m^3}$) is that limit reached? (Once declared
the limit of free diving; records now pass \qty{100}{m}, blood
shifting into the chest taking up the missing volume.)
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g10:pressure:15}
In a hydraulic lift, a liquid transmits pressure unchanged from a
small piston ($s = \qty{2.0}{cm^2}$) to a large one
($S = \qty{400}{cm^2}$) carrying a \qty{1200}{kg} car.
\begin{enumerate}
  \item What force on the small piston holds the car up? What pressure
        does the liquid carry?
  \item To raise the car \qty{2.0}{cm}, how far must the small piston
        travel? Compare the work done on each piston.
\end{enumerate}
\end{exercise}

\section{Problem: One breath down, one breath up}

\begin{problem}[{Weekend problem --- one breath down, one breath up:
the physics of a free dive and a scuba ascent, from mask squeeze to the
rule that the last ten metres are the most dangerous}]
\label{pb:g10:pressure:1}
Lena free-dives: one surface breath, down to \qty{30}{m} and back.
Marco scuba-dives beside her, breathing from a tank. Same sea
($\rho = \qty{1025}{kg/m^3}$, $g = \qty{9.81}{N/kg}$,
$P_0 = \qty{101}{kPa}$), same \qty{6.0}{L} lungs, opposite dangers ---
all governed by $P = P_0 + \rho g h$ and $PV$ = constant (temperature
constant throughout).

\medskip
\noindent\textbf{Part I --- The weight of water.}
\begin{enumerate}
  \item Compute $\rho g$ for sea water in \unit{kPa} per metre, and
        check the rule of thumb: one atmosphere per \qty{10}{m}.
  \item Compute the absolute pressure at \qty{10}{m}, \qty{20}{m},
        \qty{30}{m} and \qty{40}{m}.
  \item Express each as a multiple of the surface pressure.
  \item Lena's mask covers \qty{150}{cm^2}. At \qty{30}{m}, if the air
        inside it were still at surface pressure, what net force would
        the water exert? (Divers exhale into the mask through the nose
        to prevent this \emph{mask squeeze}.)
  \item An eardrum has area about \qty{0.60}{cm^2}. Compute the net
        force on it at \qty{3.0}{m} if the middle ear stays at surface
        pressure; why do ears hurt in a mere pool, and what must
        ``equalizing'' achieve?
\end{enumerate}

\medskip
\noindent\textbf{Part II --- One breath down.}
Lena leaves the surface with \qty{6.0}{L} of air in her lungs.
\begin{enumerate}[resume]
  \item Using Boyle's law with absolute pressures, compute her lung
        volume at \qty{10}{m}, \qty{20}{m} and \qty{30}{m}.
  \item At what depth is her lung volume halved?
  \item Is lung volume proportional to depth? Describe the curve of
        $V$ against absolute pressure $P$.
  \item Her residual volume is \qty{1.5}{L}: lungs cannot shrink
        further. Show that this limit is reached near \qty{30}{m}.
  \item Free-diving records nonetheless exceed \qty{100}{m}: what
        fills the missing volume? (What else can flow into the chest?)
\end{enumerate}

\medskip
\noindent\textbf{Part III --- One breath up.}
At \qty{30}{m}, Marco's regulator fills his \qty{6.0}{L} lungs.
\begin{enumerate}[resume]
  \item At what pressure does the regulator deliver that air? How
        many times denser is it than surface air?
  \item Marco panics, holds his breath, and rises to \qty{10}{m}:
        what volume does his trapped air demand there?
  \item What volume at the surface --- how many times his capacity?
  \item State the scuba diver's first rule, and why breathing
        normally removes the danger.
  \item Lena also holds her breath from \qty{30}{m} to the surface,
        yet her lungs are perfectly safe. Explain the asymmetry.
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- Bubbles, and the last ten metres.}
\begin{enumerate}[resume]
  \item Marco releases a \qty{1.0}{cm^3} bubble at \qty{40}{m}.
        Compute its volume at \qty{30}{m}, \qty{20}{m}, \qty{10}{m}
        and just below the surface.
  \item For each \qty{10}{m} stage of the rise, compute the factor by
        which the bubble grows. Where is the growth fastest?
  \item Divers ascend at most \qty{10}{m} per minute: how long from
        \qty{40}{m}, and in which minute does any trapped or dissolved
        gas expand the most?
  \item At depth Marco's blood dissolves extra nitrogen, like gas in
        a capped soda bottle. What does a too-fast ascent do, and how
        do the slow ascent and a \qty{5}{m} safety stop prevent it?
  \item Punchline: in one sentence each, give the free diver's
        verdict, the scuba diver's verdict, and the quantified reason
        the last ten metres of any ascent deserve the most respect.
\end{enumerate}
\end{problem}
