\chapter{Signals and Waves}\label{ch:g10:signals-and-waves}

Every message you have ever received arrived as a signal: a voice pressed
on your eardrum, this page reached you as a flicker of radio. Behind the
variety sits one idea --- a physical quantity varying in time --- and one
toolkit: period, frequency, amplitude, and the art of timing a delay.
With it, a boat measures the sea floor and a doctor counts heartbeats.

\section{Signals carry information}

\begin{definition}[Signal]\label{def:g10:signals-and-waves:signal}
A \emph{signal}\index{signal} is a physical quantity that varies in time
to carry information: the air pressure at your eardrum, the voltage at a
microphone's terminals, the brightness of the light in an optical fiber.
To study a signal, record it as a function of time and read the
information off the curve.
\end{definition}

In a phone call the same information changes carrier repeatedly:
pressure in air, then voltage in the microphone, then radio at the
antenna --- and backwards at the other end. One vocabulary handles them
all; this chapter builds it.

\section{Periodic signals}

\begin{definition}[Periodic signal, period]\label{def:g10:signals-and-waves:period}
A signal is \emph{periodic}\index{periodic signal} when it repeats
identically at regular intervals. The \emph{period}\index{period} $T$ is
the duration of one complete cycle, in seconds.
\end{definition}

\begin{definition}[Frequency]\label{def:g10:signals-and-waves:frequency}
The \emph{frequency}\index{frequency} of a periodic signal is the number
of cycles per second:
\[
f = \frac{1}{T} .
\]
Its unit, one cycle per second, is the \emph{hertz}\index{hertz}
(\unit{Hz}), with the usual multiples \unit{kHz}, \unit{MHz}, \unit{GHz}.
\end{definition}

\begin{definition}[Amplitude]\label{def:g10:signals-and-waves:amplitude}
The \emph{amplitude}\index{amplitude} of a periodic signal is its maximal
deviation from the resting value. For a voltage oscillating symmetrically
about zero, the amplitude is the peak value; crest to trough measures
\emph{twice} the amplitude (the peak-to-peak value).
\end{definition}

\begin{example}[Mains and concert pitch]\label{ex:g10:signals-and-waves:mains}
The mains voltage oscillates at $f = \qty{50}{Hz}$: its period is
$T = 1/f = \qty{0.020}{s} = \qty{20}{ms}$. An orchestra's concert A is a
pressure signal at \qty{440}{Hz}: one cycle lasts
$T = 1/440 = \qty{2.3}{ms}$. Higher frequency, shorter cycle: $f$ and $T$
are inverses of each other.
\end{example}

\begin{omfigure}
\begin{tikzpicture}[font=\small]
  \draw[-Stealth] (-0.3,0) -- (7.7,0) node[below] {$t$};
  \draw[-Stealth] (0,-1.7) -- (0,2.0) node[left] {signal};
  \draw[omDef, thick, smooth, samples=120, domain=0:7.3]
    plot (\x, {1.3*sin(120*\x)});
  \draw[dashed, thin] (0.75,1.3) -- (0.75,1.6) (3.75,1.3) -- (3.75,1.6);
  \draw[omThm, Stealth-Stealth] (0.75,1.6) -- (3.75,1.6)
    node[midway, above] {period $T$};
  \draw[omProp, Stealth-Stealth] (5.25,0) -- (5.25,-1.3)
    node[midway, right] {amplitude};
\end{tikzpicture}

{\small A periodic signal: the pattern repeats every $T$ seconds, and the
amplitude is the maximal deviation from the resting value.}
\end{omfigure}

\section{Reading an oscillogram}

\begin{definition}[Oscilloscope]\label{def:g10:signals-and-waves:oscilloscope}
An \emph{oscilloscope}\index{oscilloscope} plots a voltage against time;
the curve on its gridded screen is an \emph{oscillogram}%
\index{oscillogram}. Two settings convert screen divisions into physical
values: the \emph{time base}\index{time base} (seconds per horizontal
division) and the vertical \emph{gain}\index{gain} (volts per vertical
division).
\end{definition}

\begin{method}[From screen to numbers]\label{met:g10:signals-and-waves:oscillogram}
\begin{enumerate}
  \item Count the horizontal divisions spanned by one complete cycle
        (crest to crest is safest); multiply by the time base to get
        $T$, then $f = 1/T$.
  \item Count the vertical divisions from the center line to a crest and
        multiply by the gain to get the amplitude (crest to trough gives
        the peak-to-peak value: halve it).
\end{enumerate}
\end{method}

\begin{example}[A full reading]\label{ex:g10:signals-and-waves:reading}
On the oscillogram below, the time base is \qty{5}{ms} per division and
the gain \qty{2}{V} per division. One cycle spans $4.0$ divisions:
$T = 4.0 \times \qty{5}{ms} = \qty{20}{ms}$ and
$f = 1/\qty{0.020}{s} = \qty{50}{Hz}$ --- the mains again. A crest sits
$3.0$ divisions above the axis: the amplitude is
$3.0 \times \qty{2}{V} = \qty{6.0}{V}$.
\end{example}

\begin{omfigure}
\begin{tikzpicture}[x=0.62cm, y=0.62cm, font=\small]
  \draw[gray!35, thin] (0,0) grid (10,8);
  \draw[gray!80] (0,4) -- (10,4);
  \draw[gray!80] (5,0) -- (5,8);
  \draw[omDef, thick, smooth, samples=120, domain=0:10]
    plot (\x, {4 + 3*sin(90*\x)});
  \draw[dashed, thin] (1,7) -- (1,7.6) (5,7) -- (5,7.6);
  \draw[omThm, Stealth-Stealth] (1,7.6) -- (5,7.6)
    node[midway, above] {$4.0$ div};
  \draw[omProp, Stealth-Stealth] (7,4) -- (7,1)
    node[midway, right] {$3.0$ div};
\end{tikzpicture}

{\small An oscillogram, time base \qty{5}{ms}/div and gain \qty{2}{V}/div:
one cycle spans $4.0$ divisions ($T = \qty{20}{ms}$, $f = \qty{50}{Hz}$)
and the crest sits $3.0$ divisions up (amplitude \qty{6.0}{V}).}
\end{omfigure}

\section{Sound and electromagnetic signals}

\begin{definition}[Sound]\label{def:g10:signals-and-waves:sound}
\emph{Sound}\index{sound} is a pressure signal: a vibrating object ---
string, membrane, vocal cords --- pushes rhythmically on the air, and
the compressions travel outward at about \qty{340}{m/s}. A microphone
converts the arriving pressure into a voltage; the eardrum, into nerve
impulses. Sound needs a medium: a bell rung under a vacuum jar falls
silent as the air is pumped out.
\end{definition}

\begin{definition}[Wave]\label{def:g10:signals-and-waves:wave}
A traveling perturbation --- a sound's compression pattern, the ripple
on a pond, a radio pulse --- is a \emph{wave}\index{wave}. The medium,
when there is one, stays put: each patch of water bobs in place while
the ripple crosses the pond. The geometry of waves is grade-12 business;
this chapter only needs their speeds.
\end{definition}

\begin{definition}[Ultrasound and infrasound]\label{def:g10:signals-and-waves:ultrasound}
The human ear hears sound between about \qty{20}{Hz} and \qty{20}{kHz}
(the ceiling drops with age). Sound above \qty{20}{kHz} is
\emph{ultrasound}\index{ultrasound} --- bats hunt at \qty{50}{kHz},
medical scanners image the body at a few megahertz. Sound below
\qty{20}{Hz} is \emph{infrasound}\index{infrasound}, felt by elephants
and seismometers.
\end{definition}

\begin{proposition}[Electromagnetic signals]\label{prop:g10:signals-and-waves:em}
Radio, Wi-Fi and light are \emph{electromagnetic
signals}\index{electromagnetic signal}: one family, all traveling
through vacuum --- no medium needed --- and all at one speed, the speed
of light of \cref{ch:g10:refraction},
\[
c = \qty{3.00e8}{m/s} \quad \text{(very nearly the same in air).}
\]
\end{proposition}
\admitted

\begin{remark}\label{rem:g10:signals-and-waves:honest}
Here this is an experimental fact --- sunlight crosses $150$ million
kilometres of empty space; it is honestly derived in the Year 2 volume
from the laws of electricity and magnetism. Keep the contrast: sound is
a vibration \emph{of} the medium and dies without it; an electromagnetic
signal carries its own field across vacuum, a million times faster.
\end{remark}

\begin{omfigure}
\begin{tikzpicture}[x=1.3cm, font=\small]
  \fill[omDef!15] (1.3,-0.14) rectangle (4.3,0.14);
  \draw[-Stealth] (-0.3,0) -- (7.7,0);
  \foreach \x/\l in {0/{$1$}, 1/{$10$}, 2/{$10^2$}, 3/{$10^3$},
      4/{$10^4$}, 5/{$10^5$}, 6/{$10^6$}, 7/{$10^7$}}
    \draw (\x,0.07) -- (\x,-0.07) node[below] {\l};
  \node[below] at (7.55,-0.07) {\unit{Hz}};
  \node[gray] at (0.4,-0.75) {infrasound};
  \node[omDef] at (2.8,-0.75) {audible};
  \node[gray] at (5.8,-0.75) {ultrasound};
  \draw[omThm] (1.7,0.14) -- (1.7,0.4) node[above] {mains};
  \draw[omThm] (2.64,0.14) -- (2.64,0.95) node[above] {concert A};
  \draw[omThm] (4.7,0.14) -- (4.7,0.4) node[above] {bat};
  \draw[omThm] (6.7,0.14) -- (6.7,0.95) node[above] {medical scan};
\end{tikzpicture}

{\small Sound frequencies on a logarithmic scale (each step $\times 10$).
The ear hears from \qty{20}{Hz} to \qty{20}{kHz}; below is infrasound,
above is ultrasound.}
\end{omfigure}

\section{Echoes: distances from delays}

A signal moving at speed $v$ covers $d = v\,t$ in a time $t$: every delay
is a distance in disguise. Better still, send a short pulse at an
obstacle and time its \emph{echo} --- the round trip covers $2d$.

\begin{method}[Echo ranging]\label{met:g10:signals-and-waves:echo}
\begin{enumerate}
  \item Send a short pulse; measure the round-trip time $t$ of the echo.
  \item Take the speed $v$ of the signal \emph{in the medium crossed}.
  \item The pulse covered $2d = v\,t$, so $d = \dfrac{v\,t}{2}$.
\end{enumerate}
Sound pulses in water make \emph{sonar}\index{sonar}
($v \approx \qty{1500}{m/s}$); radio pulses in air make
\emph{radar}\index{radar} ($v = c$); ultrasound pulses in the body make
the medical scanner ($v \approx \qty{1540}{m/s}$ in soft tissue).
\end{method}

\begin{example}[Sonar]\label{ex:g10:signals-and-waves:sonar}
A ship's sonar pings and the seabed echo returns after
$t = \qty{0.80}{s}$: the depth is
$d = \frac{1500 \times 0.80}{2} = \qty{600}{m}$. Forgetting the factor
$2$ doubles the ocean.
\end{example}

\begin{omfigure}
\begin{tikzpicture}[font=\small]
  \fill[blue!8] (-3.5,0) -- (3.5,0) -- (3.5,-2.95) -- (2.6,-3.1)
    -- (1.4,-2.9) -- (0.2,-3.05) -- (-1.2,-2.85) -- (-2.3,-3.1)
    -- (-3.5,-2.95) -- cycle;
  \draw[thick, blue!60] (-3.5,0) -- (3.5,0);
  \draw[thick] (-3.5,-2.95) -- (-2.3,-3.1) -- (-1.2,-2.85)
    -- (0.2,-3.05) -- (1.4,-2.9) -- (2.6,-3.1) -- (3.5,-2.95);
  \fill[pattern=north east lines] (-3.5,-2.95) -- (-2.3,-3.1)
    -- (-1.2,-2.85) -- (0.2,-3.05) -- (1.4,-2.9) -- (2.6,-3.1)
    -- (3.5,-2.95) -- (3.5,-3.4) -- (-3.5,-3.4) -- cycle;
  \fill[gray!70] (-0.9,0.3) -- (0.9,0.3) -- (0.55,-0.25)
    -- (-0.55,-0.25) -- cycle;
  \fill[gray!50] (-0.25,0.3) rectangle (0.25,0.62);
  \draw[-Stealth, omDef, thick] (-0.35,-0.35) -- (-0.35,-2.75)
    node[midway, left] {ping, $t/2$};
  \draw[-Stealth, omThm, thick] (0.35,-2.75) -- (0.35,-0.35)
    node[midway, right] {echo, $t/2$};
  \draw[Stealth-Stealth] (2.9,-0.05) -- (2.9,-2.9)
    node[midway, right] {$d = \dfrac{v\,t}{2}$};
\end{tikzpicture}

{\small A sonar ping travels down and back: the measured delay $t$ covers
$2d$, so the depth is $d = v\,t/2$ with $v \approx \qty{1500}{m/s}$ in
seawater.}
\end{omfigure}

\begin{example}[Radar]\label{ex:g10:signals-and-waves:radar}
An airport radar receives an aircraft's echo $\qty{0.30}{ms}$ after the
pulse:
$d = \frac{\qty{3.00e8}{m/s} \times \qty{3.0e-4}{s}}{2} = \qty{45}{km}$.
Because $c$ is so large, radar lives on microsecond clocks: timing light
is precision work.
\end{example}

\begin{example}[The heart as a signal]\label{ex:g10:signals-and-waves:ecg}
Each heartbeat sends a small voltage pulse across the chest; electrodes
record it as the \emph{electrocardiogram} (ECG), a nearly periodic
signal. If the tall spikes are $T = \qty{0.75}{s}$ apart, the heart
beats at $f = 1/0.75 = \qty{1.33}{Hz}$, i.e.\ $1.33 \times 60 = 80$
beats per minute: reading $T$ off a trace is how a cardiologist takes
your pulse.
\end{example}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:g10:signals-and-waves:1}
Compute the frequency of the periodic signal of period
(a) $T = \qty{20}{ms}$; (b) $T = \qty{4.0}{ms}$;
(c) $T = \qty{50}{\micro s}$.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g10:signals-and-waves:2}
Compute the period of (a) the \qty{50}{Hz} mains; (b) a \qty{440}{Hz}
concert A; (c) a \qty{2.4}{GHz} Wi-Fi signal.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g10:signals-and-waves:3}
Classify as infrasound, audible sound or ultrasound: \qty{12}{Hz};
\qty{440}{Hz}; \qty{18}{kHz}; \qty{40}{kHz} (a car parking sensor);
\qty{5}{MHz} (a medical probe).
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g10:signals-and-waves:4}
On an oscilloscope set to \qty{2}{ms} per division and \qty{0.5}{V} per
division, one cycle spans $5.0$ divisions and a crest sits $4.0$
divisions above the axis. Find $T$, $f$ and the amplitude.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g10:signals-and-waves:5}
You see the lightning, then hear the thunder \qty{6.0}{s} later. How far
away did it strike ($v_{\text{sound}} = \qty{340}{m/s}$)? Why may the
light's travel time be neglected?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:signals-and-waves:6}
A sonar echo returns from the seabed after \qty{0.60}{s}
($v = \qty{1500}{m/s}$ in seawater). Why must the product $v\,t$ be
halved? Compute the depth. How long would the echo take over a
\qty{1200}{m} deep trench?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:signals-and-waves:7}
An airport radar receives an aircraft echo after \qty{0.24}{ms} and,
exactly \qty{1.0}{s} later, a second echo after \qty{0.238}{ms}. Compute
the two distances, then the aircraft's speed. Is it approaching?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:signals-and-waves:8}
An ECG is printed at \qty{25}{mm/s}; the tall spikes are \qty{20}{mm}
apart. Find the period and the heart rate in beats per minute.
Tachycardia means over $100$ beats per minute: below what spike spacing
does it show on this paper?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:signals-and-waves:9}
You must display a \qty{200}{Hz} signal of amplitude \qty{3.0}{V} on a
screen of $10$ horizontal and $8$ vertical divisions ($4$ above the
center line). Choose a time base showing about two full cycles, and a
gain using most of the screen without clipping.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:signals-and-waves:10}
A medical probe sends an ultrasound pulse into the abdomen
($v = \qty{1540}{m/s}$ in soft tissue) and hears echoes after
\qty{40}{\micro s} (front wall of an organ) and \qty{60}{\micro s} (back
wall). Find the depth of each wall and the organ's thickness.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:signals-and-waves:11}
A concert is broadcast live. Who hears a drumbeat first: a listener
\qty{30}{m} from the stage, or a radio listener \qty{3000}{km} away? By
how much?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g10:signals-and-waves:12}
A bat emits ultrasound pulses lasting \qty{3.0}{ms} and cannot hear an
echo while still emitting. Closer than what distance is a moth
undetectable? What is the echo delay for a moth \qty{1.7}{m} away? Why
must the bat shorten its pulses during the final approach?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g10:signals-and-waves:13}
You clap facing a cliff and hear the echo \qty{1.5}{s} later. How far is
the cliff? You walk \qty{100}{m} straight toward it: new delay? The ear
no longer separates clap from echo below about \qty{0.1}{s}: within what
distance does the echo disappear into the clap?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g10:signals-and-waves:14}
At \qty{5}{ms} per division, one cycle of a signal spans $3.0$ divisions
of a $10$-division screen.
\begin{enumerate}
  \item Find $T$ and $f$.
  \item The time base is switched to \qty{2}{ms} per division: how many
        divisions does one cycle now span?
  \item How many \emph{complete} cycles fit on the screen at each
        setting?
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g10:signals-and-waves:15}
A ``ping'' measures the round-trip time of an internet packet to a
server \qty{1200}{km} away. What is the smallest conceivable round-trip
time (signals at $c$)? In optical fiber, light travels at about
\qty{2.0e8}{m/s}: recompute. The measured ping is \qty{40}{ms}: give two
reasons it exceeds your answers.
\end{exercise}

\section{Problem: The sonar, the storm and the cardiogram}

\begin{problem}[{Weekend problem --- reading the world's signals: one
law, $d = v\,t$, ranges a storm, maps a seabed, counts a heartbeat and
teases out how GPS works}]
\label{pb:g10:signals-and-waves:1}
A fishing boat works through a stormy night: lightning on the horizon,
the sonar sweeping the bottom, the skipper's heart on the doctor's paper
strip, a GPS receiver listening to satellites. Four instruments, one
law. Take $v_{\text{sound}} = \qty{340}{m/s}$ in air, \qty{1500}{m/s} in
seawater, and $c = \qty{3.00e8}{m/s}$.

\medskip
\noindent\textbf{Part I --- The storm.}
\begin{enumerate}
  \item A flash; thunder follows \qty{9.0}{s} later. How far away is it?
  \item Compute the light's travel time over that distance, and justify
        treating the flash as instantaneous.
  \item Sailors count the seconds between flash and thunder and divide
        by three to get kilometres. Justify the rule.
  \item Five minutes later, a flash gives \qty{6.0}{s}. How far now?
        Find the storm's average approach speed in \unit{m/s} and \unit{km/h}.
  \item Thunder rumbles instead of cracking: the lightning channel is
        kilometres long, so its parts lie at different distances. If the
        channel stretches from \qty{2.0}{km} to \qty{5.0}{km} from the
        boat, how long does the rumble last?
\end{enumerate}

\medskip
\noindent\textbf{Part II --- The sonar.}
\begin{enumerate}[resume]
  \item The sonar pings; the seabed answers in \qty{0.90}{s}. Depth?
  \item Over the shelf the echo shortens to \qty{0.20}{s}. Depth?
  \item One ping returns \emph{two} echoes, at \qty{0.16}{s} and
        \qty{0.20}{s}. Interpret them; how far above the bottom does the
        fish shoal swim?
  \item The sonar pings every \qty{0.50}{s}. What is the greatest depth
        it can measure without confusion, and what goes wrong beyond it?
  \item In air, the same \qty{0.60}{s} delay would mean what distance?
        Moral: what must you know before turning a delay into a
        distance?
\end{enumerate}

\medskip
\noindent\textbf{Part III --- The cardiogram.}
\begin{enumerate}[resume]
  \item The skipper's ECG spikes are \qty{0.86}{s} apart. Frequency?
        Beats per minute?
  \item The strip advances at \qty{25}{mm/s}. What spike spacing did the
        doctor measure?
  \item After hauling nets, the spacing is \qty{15}{mm}. New heart rate?
  \item The young deckhand, a trained rower, rests at $50$ beats per
        minute. Period? Spacing on the strip?
  \item Is an ECG strictly periodic? What, then, does the doctor read
        from the trace?
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- Homeward by light.}
\begin{enumerate}[resume]
  \item The boat radios the harbor, \qty{50}{km} away. How long does the
        message take? How long would sound take?
  \item A GPS satellite orbits about \qty{20200}{km} overhead. How long
        does its signal take to reach the boat?
  \item GPS turns time into position. What distance error does a clock
        error of \qty{1.0}{\micro s} cause? What timing precision does a
        \qty{3.0}{m} fix require?
  \item Why could no sound-based GPS exist, even in principle?
  \item Finale: list the three speeds used tonight and the single law
        behind all four instruments, then state the ranging habit in one
        sentence.
\end{enumerate}
\end{problem}
