\chapter{Preface}

This book has a simple ambition: a single, coherent physics book covering
everything from kindergarten to the end of the bachelor's degree, written
in English for readers anywhere in the world.

The contents follow the old French physics programs, which we hold to be
among the finest ever designed: the physics parts of the coll\`ege and
lyc\'ee ``S'' (scientifique) track for the school years, the PCSI and PC*
classes pr\'eparatoires for the first two university years, and a
Licence~3 de physique for the third. Unlike those programs, this course is
physics only --- the chemistry that accompanied them is left to its own
book --- and the early grades, where French schools taught no physics,
are covered here by age-adapted chapters so that the course truly starts
in Grade~1.

The style is concise and rigorous: courses built from definitions, examples,
propositions and theorems, with derivations whenever they are accessible at
the given level, followed by carefully chosen exercises. Results stated
without proof are explicitly marked as admitted. Full solutions to all
exercises are collected at the end of the book.

Each school year is a part of the book, and each part is split into
chapters. The earlier the year, the younger the reader it is written for:
more illustrations, more detailed worked steps, and gentler exercises.

\section*{How to read this book}

Statements are numbered within each chapter and color-coded: definitions in
blue, theorems and laws in red, propositions and their relatives in orange,
and methods --- short recipes for standard tasks --- in green. Exercises are
graded from ($\star$) for direct applications to ($\star\star\star$) for
harder problems.

\section*{Contributing}

The source of this book lives in a public git repository:
\begin{center}
\url{https://github.com/bvirrion/one-physics-book}
\end{center}
Contributions --- new chapters, corrections, better derivations, additional
exercises --- are welcome; see \texttt{CONTRIBUTING.md} at the root of the
repository.
