\section*{Chapter \ref{ch:g10:light-spectra} --- Light Spectra and the Message of Light}

\begin{solution}{exo:g10:light-spectra:1}
$\qty{254}{nm} < \qty{400}{nm}$: ultraviolet. $\qty{480}{nm}$: visible,
blue. $\qty{589}{nm}$: visible, yellow-orange. $\qty{656}{nm}$: visible,
red. $\qty{1064}{nm} > \qty{800}{nm}$: infrared --- which is what makes a
cutting laser doubly dangerous: powerful and invisible.
\end{solution}

\begin{solution}{exo:g10:light-spectra:2}
\emph{1.} A continuous band of colors, from red to violet: the spectrum
of sunlight.

\emph{2.} The violet end --- the prism deviates short wavelengths the
most.

\emph{3.} A single green spot: deviated, but not spread into a band.
Laser light is monochromatic --- one wavelength only, so there is nothing
to sort.
\end{solution}

\begin{solution}{exo:g10:light-spectra:3}
(a) Continuous (dense hot filament). (b) Line emission (low-pressure
excited gas). (c) Continuous (dense molten metal). (d) Line emission
(essentially the single line at $\qty{589}{nm}$). (e) Absorption: the
continuous spectrum of the hot surface, scored by the dark lines of the
Sun's cooler atmosphere.
\end{solution}

\begin{solution}{exo:g10:light-spectra:4}
\emph{1.} The raindrops: each one receives white sunlight, decomposes it
and sends it back sorted by color.

\emph{2.} The Sun supplies the white light; the rain supplies millions of
tiny prisms; the geometry works out only for light returned by the drops
toward the observer, which is why the Sun must be at the observer's back
and the rain in front.

\emph{3.} The drops add no light of their own --- they only sort what
they receive. If sorted sunlight displays every color, then every color
was already contained in the sunlight: sunlight is white light.
\end{solution}

\begin{solution}{exo:g10:light-spectra:5}
\emph{1.} Dull red $<$ bright orange $<$ almost white.

\emph{2.} Both: the glow gets \emph{brighter} (effect 1), and its color
drifts from red toward the blue side (effect 2, the peak $\lambda_{\max}$
shrinking), so that more and more of the visible band is strongly lit ---
a full band reads as white.

\emph{3.} A body glowing white emits strongly across the whole visible
band; a body glowing red only manages the long-wavelength end. By Wien's
law the first is the hotter one.
\end{solution}

\begin{solution}{exo:g10:light-spectra:6}
\emph{1.}
\[
T = \frac{\qty{2.90e-3}{m.K}}{\qty{5.00e-7}{m}} = \qty{5800}{K}.
\]

\emph{2.} $\theta = 5800 - 273 = \qty{5527}{\degreeCelsius} \approx
\qty{5500}{\degreeCelsius}$.

\emph{3.} $\qty{400}{nm} < \qty{500}{nm} < \qty{800}{nm}$: the peak is
visible, in the green.
\end{solution}

\begin{solution}{exo:g10:light-spectra:7}
\emph{1.} Candle:
$\lambda_{\max} = \qty{2.90e-3}{m.K} / \qty{1800}{K} = \qty{1.61e-6}{m}
= \qty{1.6}{\micro m}$, in the infrared. Hotplate:
$\lambda_{\max} = \qty{2.90e-3}{m.K} / \qty{500}{K} = \qty{5.8e-6}{m}
= \qty{5.8}{\micro m}$, far infrared.

\emph{2.} At $\qty{500}{K}$ essentially all the radiation is infrared:
the skin absorbs it and feels heat, but the eye receives nothing --- hot,
yet black.

\emph{3.} A continuous spectrum is broad: even peaked at
$\qty{1.6}{\micro m}$, the candle's spectrum keeps a visible tail at the
red-yellow end --- weak, but enough for the dark-adapted eye.
\end{solution}

\begin{solution}{exo:g10:light-spectra:8}
Lamp A: the lines $434$, $486$, $\qty{656}{nm}$ match hydrogen, and the
faint violet line completes the catalog at $\qty{410}{nm}$: hydrogen.

Lamp B: the line sits at $\qty{589}{nm}$, sodium's only strong visible
line. Helium is excluded not by the line itself (helium does own a line
at the nearby $\qty{588}{nm}$) but by the \emph{missing} ones: helium
would also show $447$, $502$ and $\qty{668}{nm}$, and none appears. An
element is identified by its full catalog: lamp B contains sodium.
\end{solution}

\begin{solution}{exo:g10:light-spectra:9}
\emph{1.} A continuous rainbow scored by one thin dark line: the
absorption spectrum of sodium.

\emph{2.} At $\qty{589}{nm}$: a gas absorbs exactly the wavelengths it is
able to emit (\cref{prop:g10:light-spectra:kirchhoff}), and sodium's
strong visible line is at $\qty{589}{nm}$.

\emph{3.} The emission spectrum of sodium: a single bright yellow-orange
line at $\qty{589}{nm}$ on a black background --- the same wavelength,
with the two spectra exchanging bright and dark.
\end{solution}

\begin{solution}{exo:g10:light-spectra:10}
\emph{1.} $410$, $434$, $486$, $\qty{656}{nm}$: the complete hydrogen
catalog. $517$ and $\qty{518}{nm}$: the magnesium pair. Both elements are
present.

\emph{2.} In the Sun's atmosphere, the cooler and thinner gas above the
glowing surface. The surface sends up a continuous spectrum; the
atmosphere removes its own wavelengths from the passing light, so the
lines appear as missing light --- dark against the bright background.
\end{solution}

\begin{solution}{exo:g10:light-spectra:11}
\emph{1.} $T = 37 + 273 = \qty{310}{K}$, so
\[
\lambda_{\max} = \frac{\qty{2.90e-3}{m.K}}{\qty{310}{K}}
= \qty{9.4e-6}{m} = \qty{9.4}{\micro m}.
\]

\emph{2.} Far infrared --- more than ten times the wavelength of red
light.

\emph{3.} At $\qty{310}{K}$ the visible part of the radiation is
utterly negligible: no visible glow, however dark the room. A thermal
camera carries a sensor tuned to the radiation actually emitted, around
$\qty{9}{\micro m}$, and sees warm bodies shine against cooler
surroundings.
\end{solution}

\begin{solution}{exo:g10:light-spectra:12}
\emph{1.} $\lambda_{\max} = \qty{2.90e-3}{m.K} / \qty{2700}{K}
= \qty{1.07e-6}{m} \approx \qty{1.1}{\micro m}$: near infrared.

\emph{2.} The spectrum peaks beyond the visible: most of the electric
power returns as invisible infrared --- that is, heat --- and only a
small visible tail lights the room. As a lamp the bulb wastes most of its
power; as a heater it is nearly perfect.

\emph{3.} $T = \qty{2.90e-3}{m.K} / \qty{5.50e-7}{m} \approx
\qty{5300}{K}$ --- far above tungsten's melting point of
$\qty{3700}{K}$. No solid filament can be pushed to daylight
temperatures, so daylight-like lamps had to abandon incandescence
altogether (fluorescent tubes, LEDs), producing visible light without
heating a body to $\qty{5300}{K}$.
\end{solution}

\begin{solution}{exo:g10:light-spectra:13}
\emph{1.} $T = \qty{2.90e-3}{m.K} / \qty{4.20e-7}{m} \approx
\qty{6900}{K}$: hotter than the Sun's $\qty{5800}{K}$.

\emph{2.} Hydrogen (the full set $410$, $434$, $486$, $\qty{656}{nm}$ is
present) and calcium (both lines, $393$ and $\qty{397}{nm}$).

\emph{3.} No. Sodium's line at $\qty{589}{nm}$ is absent, while abundant
sodium in the atmosphere would necessarily print it on the continuous
background. If sodium is there at all, it is only in traces too faint to
detect.
\end{solution}

\begin{solution}{exo:g10:light-spectra:14}
\emph{1.} Hydrogen explains $410$, $434$, $486$ and $\qty{656}{nm}$;
helium explains $447$, $502$, $588$ and $\qty{668}{nm}$. Every observed
line is accounted for: the mixture is hydrogen and helium.

\emph{2.} Helium is a \emph{sufficient} explanation because its three
other lines ($447$, $502$, $\qty{668}{nm}$) are all present --- the
$\qty{588}{nm}$ line is then expected from helium alone, and nothing in
the spectrum \emph{requires} sodium. To settle the matter one needs
resolution, not argument: a finer spectroscope separates helium's single
line at $\qty{588.0}{nm}$ from sodium's, which is in fact a close pair at
$589.0$ and $\qty{589.6}{nm}$. Seeing one line there acquits sodium;
seeing three convicts it.
\end{solution}

\begin{solution}{exo:g10:light-spectra:15}
\emph{1.} Betelgeuse:
$T = \qty{2.90e-3}{m.K} / \qty{8.5e-7}{m} \approx \qty{3400}{K}$. Rigel:
$T = \qty{2.90e-3}{m.K} / \qty{2.4e-7}{m} \approx \qty{12000}{K}$.

\emph{2.} Since $\lambda_{\max} \times T$ is the same constant for both,
$\dfrac{T_{\text{Rigel}}}{T_{\text{Betelgeuse}}}
= \dfrac{\qty{850}{nm}}{\qty{240}{nm}} \approx 3.5$.

\emph{3.} Betelgeuse peaks in the infrared: inside the visible band its
spectrum is strongest at the red end, so the star looks red. Rigel peaks
in the ultraviolet: its visible band is strongest at the blue end, hence
blue-white.

\emph{4.} A continuous spectrum covers a huge range of wavelengths. Even
with its peak outside the visible band, a star pours plenty of light
\emph{into} that band --- the peak decides the tint, not the visibility.
\end{solution}

\begin{solution}{pb:g10:light-spectra:1}
\textbf{1.} From about $\qty{400}{nm}$ (violet end) to about
$\qty{800}{nm}$ (red end); $\qty{550}{nm}$ sits near the middle, in the
green.

\textbf{2.} $\qty{410}{nm}$: violet; $\qty{434}{nm}$: violet-blue;
$\qty{486}{nm}$: blue-green; $\qty{656}{nm}$: red.

\textbf{3.} A monochromatic yellow-orange glow. The color of a car is
the mix of wavelengths its paint reflects; under a lamp offering
\emph{only} $\qty{589}{nm}$, every paint can only reflect more or less
of that one wavelength --- all cars appear in shades of orange and
gray.

\textbf{4.} Each element emits a fixed catalog of lines, and no two
elements share the same catalog
(\cref{prop:g10:light-spectra:fingerprint}): a complete set of lines
therefore identifies the element like a fingerprint. But a
\emph{partial} match proves little --- distinct elements can own nearly
coincident single lines (helium at $\qty{588}{nm}$, sodium at
$\qty{589}{nm}$) --- so the presence claim requires all the strong lines
of the catalog.

\textbf{5.} (a) A continuous rainbow with a dark line at
$\qty{589}{nm}$: the absorption spectrum. (b) A single bright line at
$\qty{589}{nm}$ on black: the emission spectrum --- same wavelength,
bright and dark exchanged
(\cref{prop:g10:light-spectra:kirchhoff}).

\textbf{6.} The spectrum brightens at every wavelength, and its peak
$\lambda_{\max}$ moves toward shorter wavelengths, following
$\lambda_{\max} \times T = \qty{2.90e-3}{m.K}$
(\cref{prop:g10:light-spectra:wien}).

\textbf{7.} $\lambda_{\max} = \qty{2.90e-3}{m.K} / \qty{3400}{K}
= \qty{8.5e-7}{m} = \qty{850}{nm}$: near infrared, just past the red
edge --- which is why halogen light is warmer-toned than daylight.

\textbf{8.} $T = \qty{2.90e-3}{m.K} / \qty{5.00e-7}{m} = \qty{5800}{K}$.

\textbf{9.} The peak is gentle, not a spike: the Sun emits the whole
visible band with comparable strength. A full band of colors mixed
together reads as white --- the Sun looks white (yellowish through our
atmosphere), never green.

\textbf{10.} Cold poker: peak deep in the infrared, no visible tail ---
radiation is felt, nothing is seen. Hotter: the brightening spectrum
pushes a first visible tail past $\qty{800}{nm}$ --- the red end lights
up alone: dull red. Hotter still: the whole visible band fills in while
everything brightens: orange, then toward white.

\textbf{11.} The continuous background comes from the dense, hot glowing
surface; the dark lines are printed by the cooler, thinner atmosphere
above it, which absorbs its own wavelengths from the light passing
through.

\textbf{12.} Hydrogen ($410$, $434$, $486$, $\qty{656}{nm}$: complete),
sodium ($\qty{589}{nm}$) and calcium ($393$ and $\qty{397}{nm}$: both)
--- all three are present in the Sun's atmosphere.

\textbf{13.} That the atmosphere is dominated by hydrogen. Modern
measurements agree and quantify: the Sun is roughly three quarters
hydrogen by mass, most of the remainder being helium.

\textbf{14.} The line matched no known element; since each element's
catalog is fixed and unique, a line belonging to no catalog must belong
to an element never yet seen in a laboratory. A new element was
announced and named helium, after \emph{helios}, the Sun. It was
isolated on Earth in 1895 --- the prediction confirmed, twenty-seven
years later.

\textbf{15.} Seen against the brilliant continuous background, the
atmosphere \emph{removes} light at its own wavelengths: dark lines.
During the eclipse the Moon blocks the background; the thin atmosphere
then stands alone against a dark sky and its re-emitted light is all
there is to see: the same wavelengths, now bright. One gas, one catalog
--- the sign depends on what stands behind it.

\textbf{16.} $T = \qty{2.90e-3}{m.K} / \qty{2.90e-7}{m}
= \qty{10000}{K}$.

\textbf{17.} Blue-white: within the visible band the spectrum is
strongest at the blue end. And no invisibility: the continuous spectrum
floods the entire visible band on its way to the ultraviolet peak ---
the peak fixes the tint, not the brightness.

\textbf{18.} The dark lines $410$, $434$, $486$, $\qty{656}{nm}$ are the
complete hydrogen fingerprint: hydrogen dominates Vega's atmosphere.

\textbf{19.} $T_{\text{Vega}} / T_{\text{Sun}} = 10000 / 5800 \approx
1.7$: Vega is hotter, hence bluer (peak at $\qty{290}{nm}$ against
$\qty{500}{nm}$) and, by the first part of
\cref{prop:g10:light-spectra:wien}, brighter per unit of glowing surface
at every wavelength.

\textbf{20.} Identity form of Vega --- surface temperature: about
$\qty{10000}{K}$; dominant element of the atmosphere: hydrogen; apparent
color: blue-white. All of it read from one beam of light, with a prism, a
catalog of laboratory lines and one division. Comte's ``never'' was
issued in 1835 and expired in 1859: it lasted twenty-four years.
\end{solution}
