\section*{Chapter \ref{ch:g10:refraction} --- Refraction of Light}

\begin{solution}{exo:g10:refraction:1}
Angles are measured from the normal: the angle of incidence is
$90^\circ - 25^\circ = 65^\circ$, so the angle of reflection is also
$65^\circ$. The incident and reflected beams make an angle of
$2 \times 65^\circ = 130^\circ$ with each other.
\end{solution}

\begin{solution}{exo:g10:refraction:2}
\emph{1.} $v = c/n$: in water
$v = \qty{3.00e8}{m/s}/1.33 = \qty{2.26e8}{m/s}$; in diamond
$v = \qty{3.00e8}{m/s}/2.42 = \qty{1.24e8}{m/s}$.

\emph{2.} $n = \dfrac{c}{v} = \dfrac{\qty{3.00e8}{m/s}}
{\qty{1.97e8}{m/s}} = 1.52$: glass.
\end{solution}

\begin{solution}{exo:g10:refraction:3}
$\sin i_2 = \dfrac{\sin 40^\circ}{1.50} = \dfrac{0.643}{1.50} = 0.429$,
so $i_2 = 25.4^\circ$. By reversibility of light paths, a ray inside the
glass hitting the surface at $25.4^\circ$ exits into the air at exactly
$40^\circ$.
\end{solution}

\begin{solution}{exo:g10:refraction:4}
$\sin i_2 = \dfrac{\sin 55^\circ}{1.33} = \dfrac{0.819}{1.33} = 0.616$,
so $i_2 = 38.0^\circ$. The ray bends \emph{toward} the normal, because
water is the slower medium ($n_2 > n_1$).
\end{solution}

\begin{solution}{exo:g10:refraction:5}
$n = \dfrac{\sin 45^\circ}{\sin 32^\circ} = \dfrac{0.707}{0.530}
= 1.33$: the liquid is water.
\end{solution}

\begin{solution}{exo:g10:refraction:6}
\emph{1.} The five ratios are
\[
\frac{0.342}{0.233} = 1.47,\quad
\frac{0.500}{0.334} = 1.50,\quad
\frac{0.643}{0.431} = 1.49,\quad
\frac{0.766}{0.515} = 1.49,\quad
\frac{0.866}{0.581} = 1.49 :
\]
constant within measurement error --- Snell's law.

\emph{2.} $n \approx 1.49$: plexiglas.

\emph{3.} Snell's law predicts $\sin i_2 = \sin i_1 / n$: sine against
sine, the data must align on a straight line through the origin, which a
ruler checks at a glance. Plotted angle against angle, the data lie on a
curve; at small angles the curve is deceptively close to a straight
line, which is exactly what fooled Ptolemy.
\end{solution}

\begin{solution}{exo:g10:refraction:7}
$\sin i_c = n_2/n_1$ with $n_2 = 1.00$:
water $\sin i_c = 1/1.33 = 0.752$, $i_c = 48.8^\circ$;
glass $\sin i_c = 1/1.50 = 0.667$, $i_c = 41.8^\circ$;
diamond $\sin i_c = 1/2.42 = 0.413$, $i_c = 24.4^\circ$.

From air into water, $\sin i_2 = \sin i_1/1.33 \leq 1/1.33 < 1$ for
every incidence: a refracted angle always exists, so nothing special
ever happens --- total internal reflection needs $n_1 > n_2$.
\end{solution}

\begin{solution}{exo:g10:refraction:8}
\emph{1.} $d' = \dfrac{d}{n} = \dfrac{\qty{2.0}{m}}{1.33}
= \qty{1.5}{m}$.

\emph{2.} The fish appears at $\qty{40}{cm}/1.33 = \qty{30}{cm}$: the
image is \emph{above} the fish, so the heron must strike \emph{below}
the image it sees. Herons, unburdened by Snell's law, learn this by
trial and error.
\end{solution}

\begin{solution}{exo:g10:refraction:9}
\emph{1.} $\sin r = \dfrac{\sin 50^\circ}{1.50} = \dfrac{0.766}{1.50}
= 0.511$, so $r = 30.7^\circ$.

\emph{2.} Glass to air at $30.7^\circ$:
$\sin i_2 = 1.50 \times 0.511 = 0.766$, so $i_2 = 50^\circ$ --- the
exit angle equals the entry angle.

\emph{3.} The two refractions undo each other: the ray leaves
\emph{parallel} to its original direction, merely shifted sideways.
That is why a window does not distort the scene, only (imperceptibly)
translates it.
\end{solution}

\begin{solution}{exo:g10:refraction:10}
$\sin 55^\circ = 0.819$. Red:
$\sin i_2 = 0.819/1.61 = 0.509$, $i_2 = 30.6^\circ$. Blue:
$\sin i_2 = 0.819/1.66 = 0.493$, $i_2 = 29.6^\circ$. The two colors are
separated by $1.0^\circ$; blue, with the larger index, bends more and
ends up closer to the normal.
\end{solution}

\begin{solution}{exo:g10:refraction:11}
\emph{1.} $\sin i_c = \dfrac{1.46}{1.48} = 0.986$, so
$i_c = 80.6^\circ$.

\emph{2.} A ray at $i_c$ from the normal makes $90^\circ - i_c$ with
the axis, so its path along an axis length $L$ is $\dfrac{L}{\sin i_c}$:
\[
\qty{1000}{m} \times \frac{1.48}{1.46} = \qty{1013.7}{m},
\]
about \qty{14}{m} more than the axial ray.
\end{solution}

\begin{solution}{exo:g10:refraction:12}
\emph{1.} Light grazing from the horizon arrives at $i_1$ close to
$90^\circ$; it refracts to $\sin i_2 = \sin 90^\circ / 1.33 = 0.752$,
i.e.\ $i_2 = 48.8^\circ$ --- the critical angle, as reversibility
demands. Rays from the whole sky therefore reach the eye within a cone
of half-angle $48.8^\circ$.

\emph{2.} The disc's edge is where that cone meets the surface:
\[
r = d \tan i_c = \qty{2.0}{m} \times \tan 48.8^\circ
= \qty{2.0}{m} \times 1.14 = \qty{2.3}{m}.
\]

\emph{3.} Outside the disc, light from below hits the surface beyond
the critical angle and is totally reflected: the diver sees a silvery
mirror showing the bottom of the pool.
\end{solution}

\begin{solution}{exo:g10:refraction:13}
\emph{1.} Red: $\sin i_2 = 1.51 \times \sin 30^\circ = 0.755$, so
$i_2 = 49.0^\circ$. Blue: $\sin i_2 = 1.53 \times 0.500 = 0.765$, so
$i_2 = 49.9^\circ$. The white beam fans out over
$49.9^\circ - 49.0^\circ = 0.9^\circ$ --- a prism spectrum from a
single refraction.

\emph{2.} At $45^\circ$: $\sin i_2 = 1.51 \times 0.707 = 1.07 > 1$
(and $1.53 \times 0.707 = 1.08$ for blue). No refracted angle exists
for either color: the critical angle is
$\arcsin(1/1.51) = 41.5^\circ < 45^\circ$, so the long face totally
reflects the whole beam.
\end{solution}

\begin{solution}{exo:g10:refraction:14}
\emph{1.} Critical angle glass--air: $\sin i_c = 1/1.50$, so
$i_c = 41.8^\circ < 45^\circ$: the beam is totally reflected --- $100\%$
of the light, with no coating to tarnish. A metallic mirror loses a few
percent at each bounce, and a periscope needs two.

\emph{2.} Glass--water: $\sin i_c = \dfrac{1.33}{1.50} = 0.887$, so
$i_c = 62.5^\circ$. Now $45^\circ < i_c$: the incidence is below
critical, most of the light refracts out into the water, and the
mirror --- hence the image --- fails.

\emph{3.} $\sin i_2 = \dfrac{1.50}{1.33} \times \sin 45^\circ
= 1.128 \times 0.707 = 0.798$, so $i_2 = 52.9^\circ$ into the water.
\end{solution}

\begin{solution}{exo:g10:refraction:15}
\emph{1.} $v = \dfrac{\qty{3.00e8}{m/s}}{1.47} = \qty{2.04e8}{m/s}$, so
\[
t = \frac{\qty{6.2e6}{m}}{\qty{2.04e8}{m/s}} = \qty{30.4}{ms} \approx
\qty{30}{ms}.
\]

\emph{2.} Radio: $t = \qty{6.2e6}{m}/\qty{3.00e8}{m/s} =
\qty{20.7}{ms}$. The glass costs $30.4 - 20.7 = \qty{9.7}{ms}$ each
way, about \qty{19}{ms} per round trip.

\emph{3.} Lower the index --- hollow-core fibers guide light through
air ($n \approx 1.00$), and microwave links through the atmosphere do
the same --- or shorten the path: a cable laid closer to the
great-circle route between the two cities. Both are actually sold, at
prices per millisecond.
\end{solution}

\begin{solution}{pb:g10:refraction:1}
\textbf{1.} $v_{\text{water}} = \qty{3.00e8}{m/s}/1.33 =
\qty{2.26e8}{m/s}$; $v_{\text{diamond}} = \qty{3.00e8}{m/s}/2.42 =
\qty{1.24e8}{m/s}$ --- diamond slows light by the factor $n = 2.42$.

\textbf{2.} $\sin i_2 = \sin 40^\circ / 1.33 = 0.643/1.33 = 0.483$, so
$i_2 = 28.9^\circ$.

\textbf{3.} $\sin i_2 = 1.33 \times \sin 28.9^\circ = 1.33 \times
0.483 = 0.643$, so $i_2 = 40^\circ$: light retraces its path ---
reversibility of light rays.

\textbf{4.} $\sin i_c = 1/1.33 = 0.752$, so $i_c = 48.8^\circ$. At
$60^\circ > i_c$, Snell's law would need $\sin i_2 = 1.33 \sin
60^\circ = 1.15 > 1$: no refracted ray; the surface totally reflects
the light back down.

\textbf{5.} $v = \qty{3.00e8}{m/s}/1.0003 = \qty{2.999e8}{m/s}$: light
in air is slower than in vacuum by $0.03\%$, far below the precision of
our angle measurements, so $n_{\text{air}} = 1.00$ is harmless.

\textbf{6.} $d' = \qty{3.0}{m}/1.33 = \qty{2.3}{m}$: the pool hides
\qty{70}{cm} of itself.

\textbf{7.} Two rays leaving the coin refract at the surface, bending
\emph{away} from the normal, so they diverge more steeply above the
water than below. The eye prolongs the two exiting rays backwards in
straight lines; they cross \emph{above} the coin, and that crossing is
where the eye places the image.

\textbf{8.} $r = d \tan i_c = \qty{3.0}{m} \times \tan 48.8^\circ =
\qty{3.0}{m} \times 1.14 = \qty{3.4}{m}$: a disc about \qty{6.9}{m}
across holds the entire sky.

\textbf{9.} At $30^\circ$: $\sin i_2 = 1.33 \times 0.500 = 0.665$, the
ray exits at $41.7^\circ$. At $45^\circ$: $\sin i_2 = 1.33 \times
0.707 = 0.940$, it exits at $70.1^\circ$, nearly grazing. At
$50^\circ$: $1.33 \times 0.766 = 1.02 > 1$ --- total internal
reflection, the ray dives back into the pool.

\textbf{10.} The image is above the real fish
(\cref{prop:g10:refraction:depth}), so aim \emph{below} it. And yes:
the fish sees the fisher displaced too, squeezed toward the vertical
inside Snell's window --- each sees the other where the other is not.

\textbf{11.} Diamond: $\sin i_c = 1/2.42 = 0.413$, so
$i_c = 24.4^\circ$, against $41.1^\circ$ for glass: diamond's escape
cone is not even half as wide.

\textbf{12.} In diamond, $30^\circ > 24.4^\circ$: total internal
reflection, the ray stays in the stone. In glass,
$30^\circ < 41.1^\circ$: the ray leaks out the back. The facets of a
brilliant cut are angled so that entering light almost always strikes
them beyond $24.4^\circ$ and bounces until it exits upward toward the
eye --- a glass imitation, with its wide escape cone, simply lets the
light fall through.

\textbf{13.} $\sin 45^\circ = 0.707$. Red: $0.707/2.41 = 0.293$, so
$i_2 = 17.1^\circ$. Blue: $0.707/2.45 = 0.289$, so $i_2 = 16.8^\circ$.
Split at entry: about $0.3^\circ$.

\textbf{14.} $\sin 17.0^\circ = 0.292$. Red: $\sin i_2 = 2.41 \times
0.292 = 0.705$, so $i_2 = 44.8^\circ$. Blue: $\sin i_2 = 2.45 \times
0.292 = 0.716$, so $i_2 = 45.8^\circ$. The fan is now $1.0^\circ$
wide: the exit refraction, working near the critical angle, stretched
the $0.3^\circ$ entry split threefold. That stretched spray of colors
is the diamond's \emph{fire}.

\textbf{15.} Diamond--water: $\sin i_c = 1.33/2.42 = 0.550$, so
$i_c = 33.3^\circ$. The $30^\circ$ ray, trapped when the facet faced
air, now escapes ($30^\circ < 33.3^\circ$) into the water film: every
drop, fingerprint or film of grease raises the outside index, widens
the escape cone and drains the sparkle --- hence the jeweler's cloth.

\textbf{16.} $\sin i_c = \dfrac{1.46}{1.48} = 0.986$, so
$i_c = 80.6^\circ$.

\textbf{17.} Total internal reflection happens \emph{at the surface},
so the surface must stay perfect: on a bare thread, every speck of
dust, scratch, water drop or touching finger raises the outside index
at that point and bleeds light out. The cladding buries the mirror
inside clean glass where nothing can reach it --- and, as a bonus,
protects the hair-thin core mechanically.

\textbf{18.} $v = \qty{3.00e8}{m/s}/1.48 = \qty{2.03e8}{m/s}$, so
\[
t = \frac{\qty{6.0e6}{m}}{\qty{2.03e8}{m/s}} = \qty{29.6}{ms} \approx
\qty{30}{ms}.
\]

\textbf{19.} Path factor $1/\sin i_c = 1.48/1.46 = 1.0137$: at worst
$\qty{6000}{km} \times 0.0137 \approx \qty{82}{km}$ of extra glass,
costing $\qty{8.2e4}{m} / \qty{2.03e8}{m/s} \approx \qty{0.4}{ms}$ ---
the zigzag is nearly free.

\textbf{20.} Ping: $2 \times (29.6 + 0.4) \approx \qty{60}{ms}$.
Throughput: $\dfrac{10^{13}}{2.4 \times 10^8} \approx 40\,000$ copies
of this book per second. In one sentence: by trapping light inside
glass through total internal reflection, Snell's law carries a library
across the ocean every second, sixty milliseconds there and back.
\end{solution}
