\section*{Chapter \ref{ch:g10:relative-motion} --- Relative Motion}

\begin{solution}{exo:g10:relative-motion:1}
Relative to the carriage: at rest. Relative to the ground and to the
tree: in motion at \qty{100}{km/h}. Relative to the oncoming train:
speeds in opposite directions add, $100 + 100 = \qty{200}{km/h}$. In the
frame of the passenger's train, the tree moves backward at
\qty{100}{km/h} along a straight line.
\end{solution}

\begin{solution}{exo:g10:relative-motion:2}
$72 / 3.6 = \qty{20}{m/s}$; $108 / 3.6 = \qty{30}{m/s}$;
$25 \times 3.6 = \qty{90}{km/h}$; $340 \times 3.6 = \qty{1224}{km/h}$.
In each pair the \unit{km/h} figure is the larger one, as the sanity
rule demands.
\end{solution}

\begin{solution}{exo:g10:relative-motion:3}
Cyclist: $v = \frac{\qty{36}{km}}{\qty{1.5}{h}} = \qty{24}{km/h}
= 24/3.6 \approx \qty{6.7}{m/s}$. Sprinter:
$v = \frac{\qty{100}{m}}{\qty{10.0}{s}} = \qty{10.0}{m/s}
= \qty{36}{km/h}$. The sprinter is half again as fast --- but only for
ten seconds, while the cyclist holds the pace for an hour and a half.
Average speeds compare fairly only over comparable durations.
\end{solution}

\begin{solution}{exo:g10:relative-motion:4}
Same direction: $1.4 + 0.9 = \qty{2.3}{m/s}$. Against the belt:
$1.4 - 0.9 = \qty{0.5}{m/s}$. Standing still on the belt: the belt's own
\qty{0.9}{m/s}.
\end{solution}

\begin{solution}{exo:g10:relative-motion:5}
Dropped pen: ground frame works, but the \emph{train} frame is better
still --- there the pen simply falls straight down. Space station:
geocentric frame (it orbits the Earth as a whole, ignoring its spin).
Martian year: heliocentric frame (Mars runs a near-circle around the
Sun). Sprint: ground frame (start line, finish line and stopwatch are
all fixed to the ground).
\end{solution}

\begin{solution}{exo:g10:relative-motion:6}
\emph{1.} $108/3.6 = \qty{30}{m/s}$. Toward the front:
$30 + 1.5 = \qty{31.5}{m/s}$; toward the rear:
$30 - 1.5 = \qty{28.5}{m/s}$.

\emph{2.} Relative to the train the passenger covers \qty{60}{m} at
\qty{1.5}{m/s}: $t = 60/1.5 = \qty{40}{s}$. Relative to the ground the
passenger covers $31.5 \times 40 = \qty{1260}{m}$ --- the \qty{1200}{m}
carried by the train plus the \qty{60}{m} walked, as the composition
rule predicts.
\end{solution}

\begin{solution}{exo:g10:relative-motion:7}
\emph{1.} Downstream: $5.0 + 2.0 = \qty{7.0}{m/s}$, so
$t_1 = 420/7.0 = \qty{60}{s}$. Upstream: $5.0 - 2.0 = \qty{3.0}{m/s}$,
so $t_2 = 420/3.0 = \qty{140}{s}$.

\emph{2.} Average speed: $v = \frac{840}{60 + 140}
= \frac{840}{200} = \qty{4.2}{m/s} < \qty{5.0}{m/s}$. The current's
help and hindrance are equal in \emph{speed} but not in \emph{time}:
the boat spends \qty{140}{s} being slowed and only \qty{60}{s} being
helped, so the slow leg dominates the average.
\end{solution}

\begin{solution}{exo:g10:relative-motion:8}
\emph{1.} Opposite directions: in the frame of A, B approaches at
$25 + 20 = \qty{45}{m/s}$. The crossing is complete when B has slid
past the combined length $180 + 90 = \qty{270}{m}$:
$t = 270/45 = \qty{6.0}{s}$.

\emph{2.} Same direction: relative speed $25 - 20 = \qty{5.0}{m/s}$,
and $t = 270/5.0 = \qty{54}{s}$ --- nine times longer, which is why
overtaking a truck feels endless.
\end{solution}

\begin{solution}{exo:g10:relative-motion:9}
Frame of the cyclist: a circle (the rim's circle, centred on the hub).
Ground frame: a cycloid --- a chain of arches, one per turn of the
wheel. Frame of the valve itself: the valve is at rest, a single point.
The hub in the ground frame: a horizontal straight line at the height
of one wheel radius.
\end{solution}

\begin{solution}{exo:g10:relative-motion:10}
\emph{1.} $v = \frac{\qty{40000}{km}}{\qty{24}{h}}
\approx \qty{1.7e3}{km/h} = 1667/3.6 \approx \qty{463}{m/s}$ ---
faster than sound.

\emph{2.} We do not feel it because everything around us --- ground,
air, buildings --- shares exactly the same motion, so nothing moves
relative to us (and the principle of inertia,
\cref{ch:g10:inertia}, will explain why shared uniform motion is
undetectable from inside). ``The Sun rises'' is the ground-frame
report and ``the Earth spins'' the geocentric-frame report of one and
the same observation.
\end{solution}

\begin{solution}{exo:g10:relative-motion:11}
Let $L$ be the escalator's length. Escalator speed: $L/60$; walking
speed: $L/90$. Walking on the moving escalator, the speeds add:
\[
v = \frac{L}{60} + \frac{L}{90} = L\left(\frac{3 + 2}{180}\right)
= \frac{L}{36},
\]
so the ride takes \qty{36}{s}. (Not the average of $60$ and $90$:
speeds add, times do not.)
\end{solution}

\begin{solution}{exo:g10:relative-motion:12}
\emph{1.} Over a distance $d$ each way:
$t = \frac{d}{60} + \frac{d}{90} = d\,\frac{3 + 2}{180} = \frac{d}{36}$,
so $v = \frac{2d}{d/36} = \qty{72}{km/h}$.

\emph{2.} The naive $75$ averages the two speeds as if the driver spent
equal \emph{times} at each; in fact the distances are equal, so the
slower leg lasts longer ($\frac{d}{60} > \frac{d}{90}$) and drags the
average toward \qty{60}{km/h}. The average speed is total distance over
total time --- nothing else.
\end{solution}

\begin{solution}{exo:g10:relative-motion:13}
\emph{1.} In the frame of the first cyclist, the second approaches in a
straight line at $15 + 20 = \qty{35}{km/h}$ from \qty{35}{km} away:
they meet after exactly \qty{1.0}{h}.

\emph{2.} The first has ridden $15 \times 1 = \qty{15}{km}$, the second
$20 \times 1 = \qty{20}{km}$ --- total \qty{35}{km}, as it must be.

\emph{3.} Same direction: closing speed $20 - 15 = \qty{5}{km/h}$, so
the chase lasts $35/5 = \qty{7}{h}$.
\end{solution}

\begin{solution}{exo:g10:relative-motion:14}
\emph{1.} The current pushes the swimmer along the river, not across
it: the two motions compose independently, and progress toward the far
bank is made at \qty{1.2}{m/s} regardless of the drift. Crossing time:
$t = 60/1.2 = \qty{50}{s}$.

\emph{2.} Drift: $0.9 \times 50 = \qty{45}{m}$ downstream.

\emph{3.} The ground-frame velocity is the hypotenuse of a right
triangle with legs $1.2$ and $0.9$:
$v = \sqrt{1.2^2 + 0.9^2} = \sqrt{2.25} = \qty{1.5}{m/s}$ (a $3$--$4$--$5$
triangle scaled by $0.3$). Path length: $1.5 \times 50 = \qty{75}{m}$
--- or directly $\sqrt{60^2 + 45^2} = \qty{75}{m}$.
\end{solution}

\begin{solution}{exo:g10:relative-motion:15}
\emph{1.} $v = \frac{2\pi \times \num{1.50e11}}{\num{3.16e7}}
= \frac{\num{9.42e11}}{\num{3.16e7}} \approx \qty{2.98e4}{m/s}
\approx \qty{30}{km/s}$.

\emph{2.} $30 / 0.46 \approx 65$: the Earth carries us around the Sun
about sixty-five times faster than it spins us around its axis.

\emph{3.} In the geocentric frame, Mars's apparent path combines its
own orbital motion with the Earth's, and when the faster Earth
overtakes Mars the combination runs backward for a few weeks. In the
heliocentric frame both planets simply circle the Sun at steady rates
--- the ``backward loop'' is not a motion of Mars but an artefact of
the frame.
\end{solution}

\begin{solution}{pb:g10:relative-motion:1}
\textbf{1.} $t = 90/1.0 = \qty{90}{s}$.

\textbf{2.} Ground speed $1.5 + 1.0 = \qty{2.5}{m/s}$:
$t = 90/2.5 = \qty{36}{s}$.

\textbf{3.} $t = 90/1.5 = \qty{60}{s}$.

\textbf{4.} $1.5 - 1.0 = \qty{0.5}{m/s}$ relative to the ground,
against the belt's direction: $t = 90/0.5 = \qty{180}{s}$ --- three
whole minutes of brisk walking.

\textbf{5.} Ground speed $1.0 - 1.0 = \qty{0}{m/s}$: the child runs
hard and goes nowhere, at rest in the ground frame while sprinting in
the belt frame. That is a treadmill.

\textbf{6.} Belt frame / ground frame: standing traveller $0$ /
\qty{1.0}{m/s}; walker with the belt $1.5$ / \qty{2.5}{m/s}; walker on
the floor $0.5$ (the floor slides backward under the belt frame) /
\qty{1.5}{m/s}; prankster $1.5$ / \qty{0.5}{m/s}; child $1.0$ /
\qty{0}{m/s}. The legs feel the belt-frame column: effort is walking
speed relative to what is underfoot.

\textbf{7.} Progress across the river is made at \qty{2.0}{m/s}
whatever the current does: $t = 120/2.0 = \qty{60}{s}$.

\textbf{8.} Drift: $1.5 \times 60 = \qty{90}{m}$ downstream.

\textbf{9.} Ground speed: hypotenuse of a right triangle with legs
$2.0$ and $1.5$: $v = \sqrt{2.0^2 + 1.5^2} = \sqrt{6.25}
= \qty{2.5}{m/s}$. Trajectory length: $2.5 \times 60 = \qty{150}{m}$
--- the $3$--$4$--$5$ triple again, scaled to $90$--$120$--$150$.

\textbf{10.} Water frame: a straight line, \qty{120}{m}, perpendicular
to the banks. Ground frame: also a straight line, but slanted, and
\qty{150}{m} long. Same shape (straight), different direction and
length: the trajectory belongs to the frame, not to the ferry.

\textbf{11.} The upstream component of the ferry's velocity through the
water must cancel the current: $2.0 \sin\theta = 1.5$, so
$\sin\theta = 0.75$ and $\theta \approx 49^\circ$. The remaining
crossing component is $\sqrt{2.0^2 - 1.5^2} = \sqrt{1.75}
\approx \qty{1.3}{m/s}$, giving $t = 120/1.32 \approx \qty{91}{s}$.
Landing on target is slower because part of the speed through the water
is spent fighting the current instead of crossing.

\textbf{12.} Ground frame: the log drifts at the current's
\qty{1.5}{m/s}. Water frame: it is at rest, floating with the water
around it. Whether the log ``moves'' has no frame-free answer --- which
is the whole point of the chapter.

\textbf{13.} The geocentric frame. Orbit radius
$r = \num{6.37e6} + \num{4.0e5} = \qty{6.77e6}{m}$, so
\[
v = \frac{2\pi r}{T} = \frac{2\pi \times \num{6.77e6}}{\num{5.56e3}}
\approx \qty{7.65e3}{m/s} \approx \qty{7.7}{km/s}.
\]

\textbf{14.} While the station completes one orbit, the Earth turns
underneath it by about $23^\circ$, so the track over the ground shifts
westward at every pass. In the ground frame the station's path is a
complicated wavy curve that never repeats exactly; in the geocentric
frame it is one fixed near-circle.

\textbf{15.} $7.65 / 0.46 \approx 17$: the station outruns the spinning
equator roughly seventeenfold, which is why it crosses the whole sky in
minutes.

\textbf{16.} $v = \frac{2\pi \times \num{4.22e7}}{\num{8.64e4}}
\approx \qty{3.07e3}{m/s} \approx \qty{3.1}{km/s}$ in the geocentric
frame. In the ground frame it is \emph{at rest}, hanging over one fixed
point of the equator --- it circles the Earth's axis in exactly the
time the ground itself does, so the two motions cancel in the ground
frame. At rest and at \qty{3.1}{km/s}: both true, one per frame.

\textbf{17.} In the heliocentric frame the Moon travels essentially
along the Earth's orbit --- a huge near-circle around the Sun ---
overlaid with a gentle wobble of the \qty{27}{day} cycle. The Sun's
influence dominates so strongly that the path always bends toward the
Sun; it never loops backward.

\textbf{18.} ``The Sun rises'' is the report in the ground frame, where
the observer is at rest and the Sun sweeps across the sky. ``The Earth
spins'' is the report in the geocentric frame, where the Sun's
direction is steady and the observer is carried eastward around the
axis. They describe the same observation; each is correct in its own
frame, and \cref{prop:g10:relative-motion:relativity} says no
observation of the motion alone can arbitrate.

\textbf{19.} Ptolemy's Earth-centred frame matched the sky's daily
motion naturally and predicted planetary positions serviceably, but
only at the price of loops upon loops (epicycles) to reproduce each
planet's wanderings. Copernicus's Sun-centred frame made every planet a
steady near-circle and turned Mars's backward loops into a frame
effect --- the Earth overtaking a slower runner on an inside lane. The
modern verdict is not that the Sun ``really'' stands still (it too
moves, around the centre of the galaxy) but that frames are tools: the
heliocentric one makes planetary motion incomparably simpler, and
simplicity, not truth, is what a frame is for.

\textbf{20.} Ground frame: \qty{0}{km/s} --- at rest in the armchair.
Geocentric frame: about \qty{0.46}{km/s}, carried around the Earth's
axis. Heliocentric frame: about \qty{30}{km/s}, carried around the Sun.
During a \qty{90}{min} film: $d = \num{3.0e4} \times 5400
\approx \qty{1.6e8}{m} = \qty{160000}{km}$ --- nearly halfway to the
Moon without leaving the armchair. Moral: motion is not a property of a
body but of a body and a frame; choose the frame that makes the story
simple, and always say which one you chose.
\end{solution}
