\section*{Chapter \ref{ch:g10:inertia} --- Forces and the Principle of Inertia}

\begin{solution}{exo:g10:inertia:1}
(a) \omterm{def:g10:universal-gravitation:weight}{Weight} (distance), \omterm{def:g10:inertia:inventory}{normal reaction} of the table (contact).
(b) \omterm{def:g10:universal-gravitation:weight}{Weight} (distance), \omterm{def:g10:inertia:inventory}{tension} of the wire (contact).
(c) \omterm{def:g10:universal-gravitation:weight}{Weight} (distance), \omterm{def:g10:inertia:inventory}{normal reaction} of the ice (contact); \omterm{def:g10:inertia:inventory}{friction}
negligible. (d) \omterm{def:g10:universal-gravitation:weight}{Weight} only --- nothing touches the apple.
\end{solution}

\begin{solution}{exo:g10:inertia:2}
Book: $P = 0.250 \times 9.81 \approx \qty{2.45}{N}$, an arrow about
\qty{2.5}{cm} long. Person: $P = 60 \times 9.81 \approx \qty{589}{N}$
--- an arrow of \qty{5.89}{m}. Pick a scale to fit the largest \omterm{def:g10:inertia:force}{force} on
the page.
\end{solution}

\begin{solution}{exo:g10:inertia:3}
\omterm{def:g10:inertia:force}{Point of application} $A$; direction horizontal, pointing left;
magnitude $3.5 \times 2 = \qty{7.0}{N}$.
\end{solution}

\begin{solution}{exo:g10:inertia:4}
\omterm{def:g10:universal-gravitation:weight}{Weight} $\vect P$: vertical, downward, $P = 1.2 \times 9.81 \approx
\qty{11.8}{N}$. \omterm{def:g10:inertia:inventory}{Normal reaction} $\vect R$ of the shelf: same line,
upward, $R = \qty{11.8}{N}$ (the dictionary is at rest, so the two
\omterm{def:g10:inertia:compensate}{compensate}).
\end{solution}

\begin{solution}{exo:g10:inertia:5}
(a) False: \omterm{def:g10:inertia:inventory}{friction} stops things; with no \omterm{def:g10:inertia:force}{force} at all the body keeps
its \omterm{def:g10:inertia:uniform}{straight uniform motion}. (b) True (principle of inertia).
(c) False: it is subject to \omterm{def:g10:inertia:force}{forces} that \emph{\omterm{def:g10:inertia:compensate}{compensate}} (e.g.\ \omterm{def:g10:universal-gravitation:weight}{weight}
and reaction). (d) True --- the contrapositive reading of the
principle.
\end{solution}

\begin{solution}{exo:g10:inertia:6}
\emph{1.} $v = 24/3.0 = \qty{8.0}{m/s}$. \omterm{def:g10:inertia:force}{Forces}: \omterm{def:g10:universal-gravitation:weight}{weight} and \omterm{def:g10:inertia:inventory}{normal
reaction}; they \omterm{def:g10:inertia:compensate}{compensate} (\omterm{def:g10:inertia:uniform}{straight uniform motion}, \omterm{def:g10:inertia:inventory}{friction}
negligible). \emph{2.} \omterm{def:g10:inertia:inventory}{Friction} from the concrete is no longer
negligible: the inventory gains an uncompensated backward \omterm{def:g10:inertia:force}{force}, and
indeed the velocity changes --- the puck slows and stops.
\end{solution}

\begin{solution}{exo:g10:inertia:7}
\omterm{def:g10:universal-gravitation:weight}{Weight} $P = 70 \times 9.81 \approx \qty{687}{N}$ down; \omterm{def:g10:inertia:force}{force} from the
floor up. \omterm{def:g10:inertia:uniform}{Straight uniform motion}, so they \omterm{def:g10:inertia:compensate}{compensate}: the floor pushes
with \qty{687}{N}. Nothing changes at rest --- rest and uniform motion
give the same \omterm{def:g10:inertia:diagram}{force diagram}.
\end{solution}

\begin{solution}{exo:g10:inertia:8}
\emph{1.} At rest, \omterm{def:g10:inertia:inventory}{tension} \omterm{def:g10:inertia:compensate}{compensates} \omterm{def:g10:universal-gravitation:weight}{weight}:
$T = 1.8 \times 9.81 \approx \qty{17.7}{N}$. \emph{2.} Each vertical
wire carries half: $T = \qty{8.8}{N}$.
\end{solution}

\begin{solution}{exo:g10:inertia:9}
Braking: the passenger's body keeps its velocity (inertia) while the
bus loses its own, so the passenger moves forward \emph{relative to the
bus}. Turning: the body tends to continue straight while the bus turns
underneath, so it drifts toward the outside of the curve. In both
cases the ``\omterm{def:g10:inertia:force}{force}'' is fictitious; what is missing is a real \omterm{def:g10:inertia:force}{force}
(grip, handrail) to change the passenger's velocity along with the
bus's.
\end{solution}

\begin{solution}{exo:g10:inertia:10}
\omterm{def:g10:inertia:uniform}{Straight uniform motion}, so \omterm{def:g10:universal-gravitation:weight}{weight} and air resistance \omterm{def:g10:inertia:compensate}{compensate}: drag
$= P = 85 \times 9.81 \approx \qty{834}{N}$, vertical and upward.
\end{solution}

\begin{solution}{exo:g10:inertia:11}
\emph{1.} \omterm{def:g10:inertia:uniform}{Straight uniform motion}: horizontal \omterm{def:g10:inertia:force}{forces} \omterm{def:g10:inertia:compensate}{compensate}, so the
drag is \qty{300}{N}, opposite the rope. \emph{2.} The skier's \omterm{def:g10:universal-gravitation:weight}{weight}
and the upward push of the water on the skis.
\end{solution}

\begin{solution}{exo:g10:inertia:12}
\emph{1.} $P = 2.4 \times 9.81 \approx \qty{23.5}{N}$; the vertical
components carry the \omterm{def:g10:universal-gravitation:weight}{weight}: $2T\cos 40^\circ = P$, so
$T = 23.5/(2 \times 0.766) \approx \qty{15.4}{N}$. \emph{2.}
$T = P/(2\cos\theta)$ grows without bound as $\theta \to 90^\circ$
since $\cos\theta \to 0$; exactly horizontal wires have \emph{no}
vertical component at all, so nothing would carry the \omterm{def:g10:universal-gravitation:weight}{weight}.
\end{solution}

\begin{solution}{exo:g10:inertia:13}
Push: \omterm{def:g10:universal-gravitation:weight}{weight}, reaction, \omterm{def:g10:inertia:inventory}{thrust} of the hand, small \omterm{def:g10:inertia:inventory}{friction}; the speed
increases, so the \omterm{def:g10:inertia:force}{forces} do not \omterm{def:g10:inertia:compensate}{compensate}
(\cref{prop:g10:inertia:converse}). Glide: \omterm{def:g10:universal-gravitation:weight}{weight}, reaction, small
\omterm{def:g10:inertia:inventory}{friction}; the stone slows, so again no compensation --- \omterm{def:g10:inertia:inventory}{friction} is the
uncompensated leftover (on ideal ice it would glide forever). Rest:
\omterm{def:g10:universal-gravitation:weight}{weight} and reaction alone, compensating
(\cref{thm:g10:inertia:principle}).
\end{solution}

\begin{solution}{exo:g10:inertia:14}
\emph{1.} By the principle of inertia, keeping a constant velocity
requires no \omterm{def:g10:inertia:force}{force} at all --- and in interstellar space almost none
acts. Aristotle would predict the probe stops when the ``mover''
quits. \emph{2.} One year $\approx \qty{3.156e7}{s}$, so
$d = 17 \times \num{3.156e7} \approx \qty{5.4e8}{km} \approx
\qty{3.6}{au}$. \emph{3.} Velocity is a vector: changing its direction
changes the velocity, which requires an uncompensated \omterm{def:g10:inertia:force}{force}.
\end{solution}

\begin{solution}{exo:g10:inertia:15}
At rest the three \omterm{def:g10:inertia:force}{forces} \omterm{def:g10:inertia:compensate}{compensate}:
$\vect F_3 = -(\vect F_1 + \vect F_2)$, so $\vect F_3\,(-2, -4)$,
magnitude $\sqrt{4 + 16} = \sqrt{20} \approx \qty{4.5}{N}$, pointing
opposite the resultant of the first two (down-left on the grid).
\end{solution}

\begin{solution}{pb:g10:inertia:1}
\textbf{1.} \omterm{def:g10:inertia:uniform}{Straight uniform motion} at \qty{0.75}{m/s}. \omterm{def:g10:inertia:force}{Forces}: \omterm{def:g10:universal-gravitation:weight}{weight}
and the \omterm{def:g10:inertia:inventory}{normal reaction} of the belt (no \omterm{def:g10:inertia:inventory}{friction} needed: the suitcase
does not slide on the belt).
\textbf{2.} Yes --- \omterm{def:g10:inertia:uniform}{straight uniform motion}
(\cref{thm:g10:inertia:principle}). $P = 23 \times 9.81 \approx
\qty{226}{N}$, so $R = \qty{226}{N}$.
\textbf{3.} The diagrams are identical. Rest and \omterm{def:g10:inertia:uniform}{straight uniform
motion} are one and the same mechanical state: inertia does not
distinguish them.
\textbf{4.} The suitcase keeps its \qty{0.75}{m/s} while the belt
stops: it slides or tips forward until \omterm{def:g10:inertia:inventory}{friction} absorbs its motion.
Nothing pushed it --- it merely kept the velocity it had.
\textbf{5.} Speeds along the same direction add: $0.75 + 1.2 =
\qty{1.95}{m/s}$. Her velocity is constant, so the \omterm{def:g10:inertia:force}{forces} on her
\omterm{def:g10:inertia:compensate}{compensate}.
\textbf{6.} $P = 0.160 \times 9.81 \approx \qty{1.57}{N}$ down,
$R = \qty{1.57}{N}$ up, and \emph{no} horizontal \omterm{def:g10:inertia:force}{force}.
\textbf{7.} The puck keeps \qty{8.0}{m/s} with no mover whatsoever.
Everyday surfaces hide \omterm{def:g10:inertia:inventory}{friction} --- an uncompensated backward \omterm{def:g10:inertia:force}{force} ---
so motion seems to die by itself.
\textbf{8.} Its velocity changes, so the \omterm{def:g10:inertia:force}{forces} no longer \omterm{def:g10:inertia:compensate}{compensate}
(\cref{prop:g10:inertia:converse}); the culprit is the \omterm{def:g10:inertia:inventory}{friction} of the
rough ice.
\textbf{9.} No: velocity is a \emph{vector}, and its direction is
changing, so the \omterm{def:g10:inertia:force}{forces} do not \omterm{def:g10:inertia:compensate}{compensate} --- the ice pushes sideways
on her blades.
\textbf{10.} (a)--(iii), (b)--(i), (c)--(ii).
\textbf{11.} \omterm{def:g10:universal-gravitation:weight}{Weight} $P = 80 \times 9.81 \approx \qty{785}{N}$, drag
zero: nothing \omterm{def:g10:inertia:compensate}{compensates} the \omterm{def:g10:universal-gravitation:weight}{weight}, so the velocity must change ---
the jumper speeds up downward.
\textbf{12.} While drag $<$ \omterm{def:g10:universal-gravitation:weight}{weight} the \omterm{def:g10:inertia:force}{forces} still do not \omterm{def:g10:inertia:compensate}{compensate}:
the speed keeps growing, and the drag grows with it.
\textbf{13.} Steady speed: \omterm{def:g10:universal-gravitation:weight}{weight} and drag \omterm{def:g10:inertia:compensate}{compensate}, so the drag is
\qty{785}{N}. The motion is straight and uniform --- a falling body in
the exact mechanical state of the suitcase on the walkway: inertia in
disguise.
\textbf{14.} Drag now exceeds \omterm{def:g10:universal-gravitation:weight}{weight}: the \omterm{def:g10:inertia:force}{forces} do not \omterm{def:g10:inertia:compensate}{compensate}, so
the velocity changes --- the jumper, still descending, slows down; as
the speed drops, the drag shrinks back toward the \omterm{def:g10:universal-gravitation:weight}{weight}.
\textbf{15.} Steady again: drag $= P = \qty{785}{N}$ --- exactly the
same as at \qty{50}{m/s}.
\textbf{16.} The terminal drag is always equal to the \omterm{def:g10:universal-gravitation:weight}{weight}, canopy or
not. The parachute changes the \emph{speed} at which the drag reaches
\qty{785}{N}: \qty{5}{m/s} instead of \qty{50}{m/s}.
\textbf{17.} ``\dots the \omterm{def:g10:inertia:force}{forces} acting on the body \omterm{def:g10:inertia:compensate}{compensate} (or none
act)'' --- the two readings of \cref{thm:g10:inertia:principle}. Rest
is just $\vect v = \vect 0$, a \omterm{def:g10:inertia:uniform}{straight uniform motion} like any other.
\textbf{18.} The rink: the puck cruises with no mover in sight. Galileo
points at the ever-smoother surface --- \omterm{def:g10:inertia:inventory}{friction}, not nature, is what
stops things.
\textbf{19.} \omterm{def:g10:inertia:force}{Forces} do act --- \omterm{def:g10:universal-gravitation:weight}{weight} \qty{785}{N} down and drag
\qty{785}{N} up; they \emph{\omterm{def:g10:inertia:compensate}{compensate}}, which is precisely why the
speed is constant.
\textbf{20.} Walkway: \omterm{def:g10:universal-gravitation:weight}{weight} and \omterm{def:g10:inertia:inventory}{normal reaction}. Rink: \omterm{def:g10:universal-gravitation:weight}{weight} and
\omterm{def:g10:inertia:inventory}{normal reaction}. Parachute: \omterm{def:g10:universal-gravitation:weight}{weight} and drag. The one law: the principle
of inertia, in its direct and converse readings.
\end{solution}
