\section*{Chapter \ref{ch:g12:mechanical-waves} --- Mechanical Waves}

\begin{solution}{exo:g12:mechanical-waves:1}
(a) $v = 320/40 = \qty{8.0}{m/s}$. (b) No spectator moves along the
ring: the disturbance (the act of standing) travels, not the crowd.
(c) $12/1.5 = \qty{8.0}{m/s}$ --- consistent.
\end{solution}

\begin{solution}{exo:g12:mechanical-waves:2}
(a) \omterm{def:g12:mechanical-waves:transverse}{transverse}; (b) \omterm{def:g12:mechanical-waves:transverse}{longitudinal}; (c) \omterm{def:g12:mechanical-waves:transverse}{longitudinal}; (d) \omterm{def:g12:mechanical-waves:transverse}{transverse};
(e) P \omterm{def:g12:mechanical-waves:transverse}{longitudinal}, S \omterm{def:g12:mechanical-waves:transverse}{transverse}.
\end{solution}

\begin{solution}{exo:g12:mechanical-waves:3}
(a) $d = 340 \times 4.0 \approx \qty{1.4}{km}$.
(b) $1500 \times 4.0 = \qty{6.0}{km}$. (c) The \omterm{def:g12:mechanical-waves:speed}{propagation speed} in
the medium crossed.
\end{solution}

\begin{solution}{exo:g12:mechanical-waves:4}
$\lambda = v/f$: $340/440 = \qty{0.77}{m}$ in air,
$1500/440 = \qty{3.4}{m}$ in water. The \omterm{def:g10:signals-and-waves:frequency}{frequency} stays fixed (the
source dictates it); $v$ and $\lambda$ both change.
\end{solution}

\begin{solution}{exo:g12:mechanical-waves:5}
$\lambda = \qty{12}{cm}$; $T = \lambda/v = 0.12/0.30 = \qty{0.40}{s}$;
$f = 1/T = \qty{2.5}{Hz}$.
\end{solution}

\begin{solution}{exo:g12:mechanical-waves:6}
(a) $v = 2.4/0.30 = \qty{8.0}{m/s}$. (b) From the first photo the
crest has $\qty{6.0}{m}$ to go: $t = 6.0/8.0 = \qty{0.75}{s}$ after
it. (c) The ribbon rises then falls, purely transversely, and stays
at $x = \qty{5.0}{m}$: the rope does not travel.
\end{solution}

\begin{solution}{exo:g12:mechanical-waves:7}
(a) Closing speed \qty{4.0}{m/s}, gap \qty{4.0}{m}: $t = \qty{1.0}{s}$.
(b) Superposition: $3.0 + 2.0 = \qty{5.0}{cm}$.
(c) $3.0 - 2.0 = \qty{1.0}{cm}$. (d) Two intact pulses emerge and
continue unchanged, as if they had never met.
\end{solution}

\begin{solution}{exo:g12:mechanical-waves:8}
(a) The \omterm{def:g11:lenses-and-eye:lens}{thin} one: less mass per \omterm{def:g10:orders-of-magnitude:unit}{metre} at equal \omterm{def:g10:inertia:inventory}{tension} means a faster
\omterm{def:g10:signals-and-waves:wave}{wave}. (b) It increases (higher \omterm{def:g10:inertia:inventory}{tension}). (c) No: the speed depends
only on the medium, not on the pulse's size.
\end{solution}

\begin{solution}{exo:g12:mechanical-waves:9}
Rail: $1800/5900 \approx \qty{0.31}{s}$; air:
$1800/340 \approx \qty{5.3}{s}$; gap $\approx \qty{5.0}{s}$.
\end{solution}

\begin{solution}{exo:g12:mechanical-waves:10}
With $\lambda = \qty{24}{cm}$: (a) $2\lambda$, \omterm{def:g12:mechanical-waves:phase}{in phase};
(b) $\lambda/2$, opposition; (c) $\tfrac32\lambda$, opposition;
(d) $1.25\lambda$, neither; (e) $\lambda/4$, neither.
\end{solution}

\begin{solution}{exo:g12:mechanical-waves:11}
$\lambda_P = 6.0 \times 2.0 = \qty{12}{km}$;
$\lambda_S = 3.5 \times 2.0 = \qty{7.0}{km}$ --- hundreds of times a
building's size. Points much closer than $\lambda$ are nearly \omterm{def:g12:mechanical-waves:phase}{in
phase}, so the whole neighborhood moves together.
\end{solution}

\begin{solution}{exo:g12:mechanical-waves:12}
(a) $T = \qty{0.50}{s}$, $f = \qty{2.0}{Hz}$.
(b) $\lambda = \qty{0.75}{m}$.
(c) $v = \lambda/T = 0.75/0.50 = \qty{1.5}{m/s}$.
(d) At $t = T = \qty{0.50}{s}$. (e) $1.875/0.75 = 2.5$ \omterm{def:g12:mechanical-waves:wavelength}{wavelengths}
$= (2 + \tfrac12)\lambda$: \omterm{def:g12:mechanical-waves:phase}{phase opposition}.
\end{solution}

\begin{solution}{exo:g12:mechanical-waves:13}
(a) $T = 80/10 = \qty{8.0}{s}$, $f = \qty{0.125}{Hz}$.
(b) $v = \lambda/T = 120/8.0 = \qty{15}{m/s} = \qty{54}{km/h}$.
(c) $t = 900/15 = \qty{60}{s}$. (d) No: \omterm{def:g10:signals-and-waves:wave}{waves} transport no matter ---
the anchored buoy only bobs.
\end{solution}

\begin{solution}{exo:g12:mechanical-waves:14}
(a) Closing at \qty{6.0}{m/s} over \qty{6.0}{m}: $t = \qty{1.0}{s}$;
the opposite displacements cancel --- the rope is momentarily flat.
(b) No: its points are moving (the \omterm{def:g12:mechanical-waves:transverse}{transverse} velocities add, not
cancel); the \omterm{def:g10:energy-conservation:energy}{energy} is all kinetic. (c) The two pulses reappear and
continue unchanged. (d) The flat rope carries velocity, hence \omterm{def:g10:energy-conservation:energy}{energy}:
the \omterm{def:g10:signals-and-waves:wave}{wave} is momentarily stored in motion, not in shape.
\end{solution}

\begin{solution}{exo:g12:mechanical-waves:15}
(a) $2\pi r$: \qty{3.1}{m} and \qty{50}{m} --- $\times 16$.
(b) \omterm{def:g10:energy-conservation:energy}{Energy} per \omterm{def:g10:orders-of-magnitude:unit}{metre} falls by $16$, so the \omterm{def:g10:signals-and-waves:amplitude}{amplitude} drops (the
\omterm{def:g10:energy-conservation:energy}{energy} grows with the \omterm{def:g10:signals-and-waves:amplitude}{amplitude}). (c) Straight fronts do not
lengthen: only absorption remains. (d) The medium is active: each
spectator stands on their own muscle \omterm{def:g10:energy-conservation:energy}{energy}, resupplying the \omterm{def:g10:signals-and-waves:wave}{wave}.
\end{solution}

\begin{solution}{pb:g12:mechanical-waves:1}
\textbf{1.} P compress along the travel: \omterm{def:g12:mechanical-waves:transverse}{longitudinal}; S shake
sideways: \omterm{def:g12:mechanical-waves:transverse}{transverse}.

\textbf{2.} P, the faster, arrives first everywhere.
$t_P = 120/6.0 = \qty{20}{s}$; $t_S = 120/3.5 \approx \qty{34}{s}$.

\textbf{3.} $\Delta t = d/v_S - d/v_P = d\,(1/v_S - 1/v_P)$; for
\qty{120}{km}: $34.3 - 20 \approx \qty{14}{s}$.

\textbf{4.} Solve for $d$: $d = \dfrac{v_P v_S}{v_P - v_S}\,\Delta t
= \dfrac{6.0 \times 3.5}{2.5}\,\Delta t = \qty{8.4}{km}$ per second
of gap.

\textbf{5.} $\Delta t \propto d$: one station's own recording gives
its distance to the epicenter without knowing the origin time.

\textbf{6.} $d_A = 8.4 \times 25 = \qty{2.1e2}{km}$.

\textbf{7.} The gap gives a distance, not a direction: the epicenter
lies somewhere on the circle of radius $d_A$ around A.

\textbf{8.} $d_B = 8.4 \times 15 = \qty{1.3e2}{km}$;
$d_C = 8.4 \times 32 \approx \qty{2.7e2}{km}$.

\textbf{9.} Two circles meet in two points; the third selects one:
the common intersection is the epicenter.

\textbf{10.} $t_P = 210/6.0 = \qty{35}{s}$: the quake began at
$09{:}14{:}32 - \qty{35}{s} = 09{:}13{:}57$.

\textbf{11.} $\qty{200}{m/s} = \qty{720}{km/h}$ --- a cruising jet
airliner.

\textbf{12.} $t = \num{6.0e5}/200 = \qty{3.0e3}{s} = \qty{50}{min}$.

\textbf{13.} $t_P = 600/6.0 = \qty{100}{s}$. Warning:
$3000 - 100 = \qty{2.9e3}{s} \approx \qty{48}{min}$.

\textbf{14.} $\lambda = vT = 200 \times 900 = \qty{1.8e5}{m} =
\qty{180}{km}$: a \omterm{def:g10:orders-of-magnitude:unit}{metre} or so of rise spread over \qty{180}{km} ---
the ship climbs an imperceptible slope over several minutes.

\textbf{15.} $\lambda = 10 \times 900 = \qty{9.0}{km}$: the \omterm{def:g10:signals-and-waves:wave}{wave}
bunches up twentyfold, its \omterm{def:g10:energy-conservation:energy}{energy} crams into a shorter length, so its
height must grow --- the tsunami rears up at the shore.

\textbf{16.} $2\pi r$: \qty{63}{km}, then $\approx \qty{1.3e3}{km}$
--- $\times 21$.

\textbf{17.} The same \omterm{def:g10:energy-conservation:energy}{energy} is shared over a $21\times$ longer
front, so the \omterm{def:g10:signals-and-waves:amplitude}{amplitude} drops even in a lossless crust; a real crust
adds absorption.

\textbf{18.} S \omterm{def:g10:signals-and-waves:wave}{waves}, \omterm{def:g12:mechanical-waves:transverse}{transverse} shear, cannot cross a liquid: the
\omterm{def:g10:signals-and-waves:wave}{S-wave} shadow shows the (outer) core is liquid.

\textbf{19.} $\num{1e15}/\num{1.5e9} \approx \num{6.7e5}$: about
seven hundred thousand trains at full speed.

\textbf{20.} The quake struck at $09{:}13{:}57$, offshore at the
crossing of the three circles, \qty{210}{km} from station A. The
port's \omterm{def:g10:signals-and-waves:wave}{P-wave} alarm sounded \qty{100}{s} after the origin, about
\qty{48}{min} before the tsunami arrived.
\end{solution}
