\section*{Chapter \ref{ch:g12:satellites-kepler} --- Satellites and Planetary Motion}

\begin{solution}{exo:g12:satellites-kepler:1}
On the pad: $F = GMm/R^2 = \num{3.98e17}/\num{4.06e13} \approx
\qty{9.8e3}{N}$. At $r = \qty{6.77e6}{m}$:
$F \approx \qty{8.7e3}{N}$ --- still $89\%$ of the ground value. The
occupants float because they \emph{fall} together with the station,
not because gravity is gone.
\end{solution}

\begin{solution}{exo:g12:satellites-kepler:2}
$v = \sqrt{GM/R} = \sqrt{\num{3.98e14}/\num{6.37e6}} \approx
\qty{7.9e3}{m/s}$; $T = 2\pi R/v \approx \qty{5.1e3}{s} \approx
\qty{84}{min}$. Exactly the cannonball's \qty{7.9}{km/s}: a skimming
\omterm{def:g10:universal-gravitation:satellite}{orbit} \emph{is} Newton's shot.
\end{solution}

\begin{solution}{exo:g12:satellites-kepler:3}
$v \propto 1/\sqrt{r}$: divided by $\sqrt 2 \approx 1.41$.
$T \propto r^{3/2}$: multiplied by $2^{3/2} \approx 2.8$. Neither
contains $m$: no.
\end{solution}

\begin{solution}{exo:g12:satellites-kepler:4}
$v = \sqrt{\num{3.98e14}/\num{6.77e6}} \approx \qty{7.67e3}{m/s}$;
$T = 2\pi r/v \approx \qty{5.55e3}{s} \approx \qty{92}{min}$;
$\num{86400}/5546 \approx 15.6$ \omterm{def:g10:universal-gravitation:satellite}{orbits} per day.
\end{solution}

\begin{solution}{exo:g12:satellites-kepler:5}
\omterm{def:g12:satellites-kepler:ellipse}{Ellipses} with the Sun at a focus; equal areas in equal times;
$T^2/a^3$ identical for all planets. Fastest at \omterm{def:g12:satellites-kepler:ellipse}{perihelion}, by the
second law (short sweep segment must sweep fast). No: the \omterm{def:g10:universal-gravitation:satellite}{satellite}'s
mass cancels --- only the central mass enters.
\end{solution}

\begin{solution}{exo:g12:satellites-kepler:6}
$T = \qty{2.36e6}{s}$: $M = 4\pi^2 r^3/(GT^2) =
\num{2.24e27}/371 \approx \qty{6.0e24}{kg}$ --- within $1\%$ of
\qty{5.97e24}{kg}.
\end{solution}

\begin{solution}{exo:g12:satellites-kepler:7}
Gravity points at the Earth's center, and in circular motion the \omterm{def:g11:forces-and-motion:netforce}{net
force} points at the circle's center: the two centers coincide. A
circle above Paris's parallel is centered on the axis, not the
Earth's center --- impossible. Hovering \omterm{def:g11:work-of-force:work}{works} only where the parallel
is a great circle centered at the center: over the \emph{equator}.
\end{solution}

\begin{solution}{exo:g12:satellites-kepler:8}
$T = \qty{2.754e4}{s}$: $M = 4\pi^2 r^3/(GT^2) =
\num{3.26e22}/\num{5.06e-2} \approx \qty{6.4e23}{kg}$ --- the data
card's Mars.
\end{solution}

\begin{solution}{exo:g12:satellites-kepler:9}
Earth: $T^2/a^3 = \num{9.96e14}/\num{3.35e33} =
\qty{2.98e-19}{s^2/m^3}$. Mars: $T = \qty{5.94e7}{s}$,
$\num{3.52e15}/\num{1.18e34} = \qty{2.98e-19}{s^2/m^3}$ --- equal, as
the third law demands. Then $M = 4\pi^2/(G \times \num{2.98e-19})
\approx \qty{1.99e30}{kg}$.
\end{solution}

\begin{solution}{exo:g12:satellites-kepler:10}
$v = 2\pi r/T = \qty{1.02e3}{m/s}$; $a = v^2/r =
\qty{2.72e-3}{m/s^2}$; $g/60^2 = 9.81/3600 = \qty{2.73e-3}{m/s^2}$.
The Moon sits $60$ Earth radii out, and its fall is $60^2$ times
weaker than the apple's: exactly the inverse square --- one law for
both.
\end{solution}

\begin{solution}{exo:g12:satellites-kepler:11}
Ceres: $T = 2.77^{3/2} = \sqrt{21.3} \approx \qty{4.6}{yr}$. Halley:
$T = 17.8^{3/2} \approx \qty{75}{yr}$, so next \omterm{def:g12:satellites-kepler:ellipse}{perihelion} around
$1986 + 75 \approx 2061$.
\end{solution}

\begin{solution}{exo:g12:satellites-kepler:12}
At \qty{400}{km}: $v \approx \qty{7.67e3}{m/s}$; at \qty{350}{km}
($r = \qty{6.72e6}{m}$): $v \approx \qty{7.70e3}{m/s}$. \omterm{def:g10:inertia:inventory}{Friction} does
drain \omterm{def:g11:mechanical-energy:em}{mechanical energy} --- the \omterm{def:g10:universal-gravitation:satellite}{orbit} shrinks; but on the way down
gravity's \omterm{def:g11:work-of-force:work}{work} more than repays the loss of speed to drag:
$E_m$ falls while $E_k$ rises.
\end{solution}

\begin{solution}{exo:g12:satellites-kepler:13}
$T = \qty{3.024e5}{s}$: $M = 4\pi^2 r^3/(GT^2) =
\num{1.65e31}/6.10 \approx \qty{2.7e30}{kg} \approx 1.4$ Suns. The
\emph{planet's} mass is out of reach: it cancelled from the \omterm{def:g10:universal-gravitation:satellite}{orbit}
equations.
\end{solution}

\begin{solution}{exo:g12:satellites-kepler:14}
$T = \qty{8.862e4}{s}$, $GM = \qty{4.28e13}{m^3/s^2}$:
$r^3 = GMT^2/(4\pi^2) = \num{8.52e21}$, so $r \approx
\qty{2.04e7}{m}$ and $h = r - R \approx \qty{1.7e7}{m} \approx
\qty{17000}{km}$ above the Martian ground.
\end{solution}

\begin{solution}{exo:g12:satellites-kepler:15}
$v_a = r_p v_p / r_a = \num{8.77e10} \times \num{5.45e4} /
\num{5.25e12} \approx \qty{9.1e2}{m/s}$. Since $r_a/r_p \approx 60$,
the comet is $60$ times farther and $60$ times slower: it spends
decades dawdling in the dark and weeks sprinting past the Sun.
\end{solution}

\begin{solution}{pb:g12:satellites-kepler:1}
\textbf{1.} $T = \qty{86164}{s}$: the ground turns with the Earth
\emph{relative to the stars} once per \omterm{def:g12:satellites-kepler:geo}{sidereal day}; the \qty{24}{h}
day adds the degree per day the Earth advances around the Sun.

\textbf{2.} The \omterm{def:g11:forces-and-motion:netforce}{net force} of circular motion aims at the circle's
center; the only \omterm{def:g10:inertia:force}{force}, gravity, aims at the Earth's center: the two
centers are the same point.

\textbf{3.} A Paris-hovering circle would be centered on the axis
north of the center --- excluded by 2. The \omterm{def:g10:universal-gravitation:satellite}{orbit} must lie in the
\emph{equatorial plane}.

\textbf{4.} Normal component of \omterm{thm:g12:newtons-laws:second}{Newton's second law}:
$m v^2/r = GmM/r^2$, so $v = \sqrt{GM/r}$ ($m$ cancels).

\textbf{5.} $T = 2\pi r/v = 2\pi\sqrt{r^3/(GM)}$; squaring,
$T^2/r^3 = 4\pi^2/(GM)$.

\textbf{6.} $r = \left(GMT^2/4\pi^2\right)^{1/3} =
(\num{7.49e22})^{1/3} \approx \qty{4.22e7}{m}$.

\textbf{7.} $h = r - R = \num{4.22e7} - \num{6.37e6} \approx
\qty{3.58e7}{m} \approx \qty{35800}{km}$.

\textbf{8.} $v = 2\pi \times \num{4.22e7}/\num{86164} \approx
\qty{3.07e3}{m/s}$.

\textbf{9.} $\sqrt{\num{3.98e14}/\num{4.22e7}} \approx
\qty{3.07e3}{m/s}$ --- consistent.

\textbf{10.} Identical radii: the mass cancelled in 4. The heavier
\omterm{def:g10:universal-gravitation:satellite}{satellite} changes only the launch bill --- more fuel for the same
\omterm{def:g10:universal-gravitation:satellite}{orbit}.

\textbf{11.} $v = \sqrt{\num{3.98e14}/\num{6.77e6}} \approx
\qty{7.67e3}{m/s}$.

\textbf{12.} $T = 2\pi r/v \approx \qty{5.55e3}{s} \approx
\qty{92.4}{min}$.

\textbf{13.} $\num{86400}/5546 \approx 15.6$ \omterm{def:g10:universal-gravitation:satellite}{orbits}: about $16$
sunrises a day.

\textbf{14.} $r_{\text{geo}}/r_{\text{ISS}} = \num{4.22e7}/\num{6.77e6}
= 6.23$; $6.23^{3/2} \approx 15.5 = \num{86164}/5546$ --- Kepler's
third law, verified on the spot.

\textbf{15.} The station is faster (\qty{7.67}{km/s} against
\qty{3.07}{km/s}), yet the \omterm{def:g12:satellites-kepler:geo}{geostationary} cost more \omterm{def:g10:energy-conservation:energy}{energy} per
\omterm{def:g10:orders-of-magnitude:unit}{kilogram}: the climb in potential \omterm{def:g10:energy-conservation:energy}{energy} outweighs the loss of speed.

\textbf{16.} $M_J = 4\pi^2 r^3/(G T^2)$.

\textbf{17.} $T = \qty{1.529e5}{s}$: $M_J =
\num{2.97e27}/1.56 \approx \qty{1.90e27}{kg}$.

\textbf{18.} $\num{1.90e27}/\num{5.97e24} \approx 318$ Earths ---
and about $1/1000$ of the Sun.

\textbf{19.} $T_J = \qty{3.576e4}{s}$: $r^3 = GM_J T_J^2/(4\pi^2) =
\num{4.11e24}$, $r \approx \qty{1.60e8}{m}$ --- Io, at
\qty{4.22e8}{m}, flies $2.6$ times higher and drifts across Jupiter's
sky.

\textbf{20.} Park the relay on the equatorial circle of altitude
\qty{35800}{km}, where it rides at \qty{3.07}{km/s} and never leaves
the dish's aim; and the same law, fed Io's month, weighed Jupiter at
$\qty{1.90e27}{kg}$ --- $318$ Earths --- for free.
\end{solution}
