\section*{Chapter \ref{ch:g11:color-light-sources} --- Color and Light Sources}

\begin{solution}{exo:g11:color-light-sources:1}
(a) yellow; (b) cyan; (c) white. Complementary to blue: yellow
(blue $+$ yellow $=$ blue $+$ red $+$ green $=$ white).
\end{solution}

\begin{solution}{exo:g11:color-light-sources:2}
Yellow: R, G. Magenta: R, B. White: R, G, B. Black: none. Orange: R at
full \omterm{def:g10:energy-conservation:power}{power}, G at reduced \omterm{def:g10:energy-conservation:power}{power}.
\end{solution}

\begin{solution}{exo:g11:color-light-sources:3}
Magenta absorbs the green band: $(R, B)$ emerges, magenta light. The
cyan \omterm{def:g11:color-light-sources:subtractive}{filter} then absorbs the red band: only $(B)$ survives --- blue.
\end{solution}

\begin{solution}{exo:g11:color-light-sources:4}
Yellow stripe: \omterm{def:g11:color-light-sources:objectcolor}{diffuses} R and G, absorbs B. (a) Under red light it
\omterm{def:g11:color-light-sources:objectcolor}{diffuses} R: red. (b) Under blue light it absorbs everything: black.
\end{solution}

\begin{solution}{exo:g11:color-light-sources:5}
$T = \dfrac{\qty{2.90e-3}{m.K}}{\qty{5.80e-7}{m}} = \qty{5000}{K}$.
\end{solution}

\begin{solution}{exo:g11:color-light-sources:6}
\emph{(a)} $\lambda_{\max} = \qty{2.90e-3}{m.K}/\qty{1800}{K}
= \qty{1.61e-6}{m} = \qty{1610}{nm}$: \omterm{def:g10:light-spectra:wavelength}{infrared}.

\emph{(b)} The \omterm{def:g10:light-spectra:white}{spectrum} is continuous: its visible tail is small but not
zero, and the blue end fades first --- what remains visible is mostly
red-orange, hence the flame's color.
\end{solution}

\begin{solution}{exo:g11:color-light-sources:7}
Jacket (\omterm{def:g11:color-light-sources:objectcolor}{diffuses} R): (a) red; (b) red; (c) black. Gloves (diffuse G and
B, absorb R): (a) cyan; (b) black; (c) green.
\end{solution}

\begin{solution}{exo:g11:color-light-sources:8}
\emph{(a)} An \omterm{def:g10:light-spectra:emission}{emission line spectrum}. Every object can only diffuse
$\qty{589}{nm}$ or nothing: the street is shades of yellow-orange and
black.

\emph{(b)} The blue car \omterm{def:g11:color-light-sources:objectcolor}{diffuses} only the blue band. Daylight contains
it: blue car. The sodium line carries no blue: everything is absorbed,
the car looks black.
\end{solution}

\begin{solution}{exo:g11:color-light-sources:9}
\emph{(a)} The white beam fans out into a rainbow; the \omterm{def:g11:color-light-sources:lamps}{laser} beam is
deviated but stays a single beam of the same green: it is \omterm{def:g10:light-spectra:white}{monochromatic}
--- one \omterm{def:g10:light-spectra:wavelength}{wavelength}, nothing for the prism to sort.

\emph{(b)} The red \omterm{def:g11:color-light-sources:subtractive}{filter} transmits only the red band and absorbs
$\qty{532}{nm}$: almost nothing comes out; the spot disappears.
\end{solution}

\begin{solution}{exo:g11:color-light-sources:10}
\omterm{def:g10:light-spectra:white}{Spectrum}: a narrow blue peak near $\qty{450}{nm}$ plus the \omterm{def:g11:color-light-sources:led}{phosphor}'s
broad yellow band (roughly $500$--$\qty{700}{nm}$). Blue and yellow are
complementary, so the \omterm{def:g11:color-light-sources:cones}{cone} triple reads white. The band fades toward
deep red, so a deep red object receives little it can diffuse: it looks
duller than in daylight, whose \omterm{def:g10:light-spectra:continuous}{continuous spectrum} feeds it fully.
\end{solution}

\begin{solution}{exo:g11:color-light-sources:11}
(a) $\lambda_{\max} = \qty{2.90e-3}{m.K}/\qty{1800}{K}
= \qty{1610}{nm}$, \omterm{def:g10:light-spectra:wavelength}{infrared};
(b) $\qty{2.90e-3}{m.K}/\qty{6500}{K} = \qty{446}{nm}$, blue-violet.
The candle light is called ``warm'' --- the source with the
\emph{lower} temperature: vocabulary runs opposite to the \omterm{def:g10:light-spectra:kelvin}{kelvin} scale.
\end{solution}

\begin{solution}{exo:g11:color-light-sources:12}
\emph{(a)} \omterm{def:g11:color-light-sources:perception}{Metamers}. Their existence proves that color is a \omterm{def:g11:color-light-sources:perception}{perception}
--- three \omterm{def:g11:color-light-sources:cones}{cone} responses --- and not the \omterm{def:g10:light-spectra:white}{spectrum} itself: infinitely
many spectra map to the same triple.

\emph{(b)} Prism or \omterm{def:g10:light-spectra:white}{spectroscope}: one line at $\qty{589}{nm}$ against
two bands at $540$ and $\qty{610}{nm}$. Red \omterm{def:g11:color-light-sources:subtractive}{filter}: it absorbs
$\qty{589}{nm}$, the sodium spot goes dark, while the screen's
$\qty{610}{nm}$ subpixel survives --- the screen patch turns red.
\end{solution}

\begin{solution}{exo:g11:color-light-sources:13}
\emph{(a)} $\lambda_{\max} = \qty{2.90e-3}{m.K}/\qty{3200}{K} =
\qty{906}{nm}$. \omterm{def:g11:color-light-sources:colortemp}{Color temperature} $\qty{3200}{K} < \qty{6500}{K}$:
warmer (more orange) than daylight.

\emph{(b)} The gel absorbs part of the red-orange excess to rebalance
the \omterm{def:g10:light-spectra:white}{spectrum} toward blue; since a \omterm{def:g11:color-light-sources:subtractive}{filter} only removes light, the total
\omterm{def:g10:energy-conservation:power}{power} on stage drops.

\emph{(c)} A \omterm{def:g11:color-light-sources:subtractive}{filter} is passive: it can absorb at each \omterm{def:g10:light-spectra:wavelength}{wavelength} but
never emit, so no gel can raise the blue intensity above what the spot
already produces.
\end{solution}

\begin{solution}{exo:g11:color-light-sources:14}
\emph{(a)} $\qty{2700}{K}$: warm, red-rich light --- the meat's red band
is well fed, the meat looks vividly fresh. $\qty{6500}{K}$: cool
bluish-white --- the fish \omterm{def:g11:color-light-sources:objectcolor}{diffuses} the full triple and looks bright
white, freshly iced.

\emph{(b)} Neither changed his goods: the absorption habits of meat and
fish are untouched. What changed is the illumination, hence the diffused
light --- the \omterm{def:g11:color-light-sources:objectcolor}{color of an object} is defined only under a given
illumination.
\end{solution}

\begin{solution}{exo:g11:color-light-sources:15}
\emph{(a)} Sodium: all the city's glow sits in one line at
$\qty{589}{nm}$, which a single narrow \omterm{def:g11:color-light-sources:subtractive}{filter} removes from telescope
images. A broad LED \omterm{def:g10:light-spectra:white}{spectrum} cannot be filtered out without discarding
the starlight too.

\emph{(b)} Drivers and pedestrians prefer the LED: under its full
\omterm{def:g10:light-spectra:white}{spectrum} objects keep distinguishable colors (a red coat stays red),
while sodium light makes everything yellow-orange or black.

\emph{(c)} Warm white LEDs, around $\qty{2700}{K}$: acceptable color
rendering for the street, and a \omterm{def:g10:light-spectra:white}{spectrum} poor in the blue end, the part
that pollutes the night sky most.
\end{solution}

\begin{solution}{pb:g11:color-light-sources:1}
\textbf{1.} R $+$ G $=$ yellow, R $+$ B $=$ magenta, G $+$ B $=$ cyan;
triple overlap white.

\textbf{2.} Orange: red at full \omterm{def:g10:energy-conservation:power}{power}, green at partial \omterm{def:g10:energy-conservation:power}{power}, blue
off. Pale pink: all three on (white base) with red slightly dominant.

\textbf{3.} The shadow cast from the red spot receives only the green
spot: green shadow; symmetrically the other shadow is red.

\textbf{4.} Deep blue paint \omterm{def:g11:color-light-sources:objectcolor}{diffuses} only B. Under the red spot alone:
black. Under all three: it \omterm{def:g11:color-light-sources:objectcolor}{diffuses} the blue band --- blue.

\textbf{5.} The eye reduces any \omterm{def:g10:light-spectra:white}{spectrum} to three \omterm{def:g11:color-light-sources:cones}{cone} responses;
matching the triple is enough to match the color. Three \omterm{def:g11:color-light-sources:cones}{cone} types fix
the number of spotlights.

\textbf{6.} Yellow $=$ white $-$ blue: the dye absorbs the blue band.

\textbf{7.} Red spot: \omterm{def:g11:color-light-sources:objectcolor}{diffuses} R --- red. Green spot: \omterm{def:g11:color-light-sources:objectcolor}{diffuses} G ---
green. Blue spot: absorbs everything --- black.

\textbf{8.} Light the scene with the blue spot alone; the flick is
white $\to$ blue. The yellow costume \omterm{def:g11:color-light-sources:objectcolor}{diffuses} R and G but absorbs B:
bright under white, black under blue.

\textbf{9.} Magenta gel: $(R, G, B) \to (R, B)$. The costume \omterm{def:g11:color-light-sources:objectcolor}{diffuses}
R and G of what it receives: it \omterm{def:g11:color-light-sources:objectcolor}{diffuses} R, absorbs B --- red costume.

\textbf{10.} A red gel keeps one band of three and dumps two thirds of
the white spot's \omterm{def:g10:energy-conservation:power}{power}; each C, M, Y gel removes a single band, keeps
two thirds, and stacking two of them still yields any primary.

\textbf{11.} An \omterm{def:g10:light-spectra:emission}{emission line spectrum}. The street becomes a monochrome
world: every surface is yellow-orange, dimmer or brighter, or black.

\textbf{12.} White wall: yellow-orange (\omterm{def:g11:color-light-sources:objectcolor}{diffuses} the line). Yellow
marking: bright yellow-orange. Red car: dark --- the $\qty{589}{nm}$
line lies outside the band it \omterm{def:g11:color-light-sources:objectcolor}{diffuses}. Blue car: black.

\textbf{13.} One narrow \omterm{def:g11:color-light-sources:subtractive}{filter} centered on $\qty{589}{nm}$ removes the
entire city glow from an image while sacrificing almost none of the
stars' broad spectra.

\textbf{14.} $\lambda_{\max} = \qty{2.90e-3}{m.K}/\qty{2700}{K} =
\qty{1.07e-6}{m} = \qty{1070}{nm}$: the peak, and most of the \omterm{def:g10:energy-conservation:power}{power},
lies in the \omterm{def:g10:light-spectra:wavelength}{infrared} --- heat, not light.

\textbf{15.} Sodium: $0.30 \times \qty{100}{W} = \qty{30}{W}$ of
visible light; tungsten: $0.05 \times \qty{100}{W} = \qty{5}{W}$. Six
times more light for the same bill.

\textbf{16.} The \omterm{def:g11:color-light-sources:led}{phosphor} converts part of the blue into a broad yellow
band; blue and yellow are complementary, so blue $+$ yellow completes
the triple: white.

\textbf{17.} A narrow blue spike plus a broad yellow hump --- against
daylight's even continuum. The deep red end (and a dip between blue and
yellow) is under-represented.

\textbf{18.} Deep red objects are rendered dull and brownish. Hence the
butcher's $\qty{2700}{K}$ warm LEDs, whose reinforced red content keeps
the meat vivid.

\textbf{19.} Nothing inside is at $\qty{2700}{K}$: the chip runs near
room temperature. The label promises only the \emph{hue} of a
$\qty{2700}{K}$ incandescent body, whose peak would be
$\qty{2.90e-3}{m.K}/\qty{2700}{K} = \qty{1.07e-6}{m} = \qty{1070}{nm}$.

\textbf{20.} Bedroom: about $\qty{2700}{K}$ --- warm, low-blue light
for rest. Jeweler: $5500$--$\qty{6500}{K}$ --- daylight-like white, so
whites read white and stones return every band. The law: perceived
color $=$ source \omterm{def:g10:light-spectra:white}{spectrum} $\times$ object absorption, read through
three \omterm{def:g11:color-light-sources:cones}{cones}.
\end{solution}
