\section*{Chapter \ref{ch:g11:fundamental-interactions} --- The Fundamental Interactions}

\begin{solution}{exo:g11:fundamental-interactions:1}
Proton: $10^{-15}$ (nucleon floor); water molecule: $10^{-10}$
(molecule floor --- single molecules sit at atomic sizes); sand:
$10^{-4}$ (everyday matter); Earth: $10^{7}$ (planet); Milky Way:
$10^{21}$ (galaxy). All in metres.
\end{solution}

\begin{solution}{exo:g11:fundamental-interactions:2}
Negative charge: it \emph{gained}
$N = \num{4.8e-9}/\num{1.6e-19} = \num{3.0e10}$ electrons. The hair
carries $+\qty{4.8e-9}{C}$, by conservation of charge
(\cref{prop:g11:fundamental-interactions:conservation}).
\end{solution}

\begin{solution}{exo:g11:fundamental-interactions:3}
$F = \num{8.99e9} \times \dfrac{\num{2.0e-6} \times \num{3.0e-6}}{0.30^2}
= \dfrac{\num{5.39e-2}}{0.090} \approx \qty{0.60}{N}$, attractive
(unlike signs). The two forces have the same magnitude, opposite
directions.
\end{solution}

\begin{solution}{exo:g11:fundamental-interactions:4}
$F_0/4$; \ $9F_0$; \ $2F_0$; \ $F_0$ (numerator $\times 4$, denominator
$\times 4$).
\end{solution}

\begin{solution}{exo:g11:fundamental-interactions:5}
(a) gravitation; (b) electromagnetic; (c) strong; (d) weak;
(e) electromagnetic.
\end{solution}

\begin{solution}{exo:g11:fundamental-interactions:6}
$F_E = \num{8.99e9} \times (\num{1.60e-19})^2/(\num{2.0e-15})^2
= \num{2.30e-28}/\num{4.0e-30} \approx \qty{58}{N}$;
$F_G = \num{6.67e-11} \times (\num{1.67e-27})^2/\num{4.0e-30}
\approx \qty{4.7e-35}{N}$. Ratio $F_E/F_G \approx \num{1.2e36}$. The
strong interaction holds the nucleus, gravity is irrelevant.
\end{solution}

\begin{solution}{exo:g11:fundamental-interactions:7}
$F_E = \num{2.30e-28}/(\num{5.3e-11})^2 = \num{2.30e-28}/\num{2.81e-21}
\approx \qty{8.2e-8}{N}$;
$F_G = \num{6.67e-11} \times \num{1.67e-27} \times \num{9.11e-31} /
\num{2.81e-21} \approx \qty{3.6e-47}{N}$. Ratio $\approx \num{2.3e39}$:
the electromagnetic interaction holds atoms together.
\end{solution}

\begin{solution}{exo:g11:fundamental-interactions:8}
By contact the charge is shared: $q_A = q_B = \qty{+4.0e-9}{C}$ (total
still $\qty{8.0e-9}{C}$). Then
$F = \num{8.99e9} \times (\num{4.0e-9})^2/0.10^2 =
\qty{1.4e-5}{N}$, repulsive (like signs).
\end{solution}

\begin{solution}{exo:g11:fundamental-interactions:9}
$d = \sqrt{k q^2/F} = \sqrt{\num{8.99e9} \times \num{1.0e-12}/0.90}
\approx \qty{0.10}{m}$. A force $9$ times smaller needs a distance $3$
times larger: $d = \qty{0.30}{m}$.
\end{solution}

\begin{solution}{exo:g11:fundamental-interactions:10}
$F = \num{2.30e-28}/(\num{1.0e-10})^2 = \qty{2.3e-8}{N}$. The strong
interaction contributes nothing: $10^{-10}$~m is $10^{5}$ times its
range. Molecules are bound by the electromagnetic interaction.
\end{solution}

\begin{solution}{exo:g11:fundamental-interactions:11}
$q = \num{e23}/\num{e6} \times \num{1.6e-19} = \qty{1.6e-2}{C}$ on each
coin. $F = \num{8.99e9} \times (\num{1.6e-2})^2/1.0^2 \approx
\qty{2.3e6}{N}$ --- the weight of $\num{2.3e6}/9.81 \approx
\qty{2.3e5}{kg}$, a loaded freight train, from a millionth of the
electrons. Everyday matter must be (and is) neutral to far better than
one part in a million.
\end{solution}

\begin{solution}{exo:g11:fundamental-interactions:12}
$F = \num{2.30e-28}/(\num{1.4e-14})^2 = \num{2.30e-28}/\num{1.96e-28}
\approx \qty{1.2}{N}$ --- still enormous at this scale. No: they are
$14$ times the strong range apart; each is glued only to its nearest
neighbours. Pairs: $\frac{92 \times 91}{2} = 4186$, \emph{all}
repelling at any distance, while the glue does not grow with size.
Neutrons add attraction without repulsion and dilute the protons; but
repulsion grows with the square of the proton number and glue only with
the number of neighbours, so beyond about $80$ protons no neutron
budget balances the books --- such nuclei decay
(\cref{ch:g11:nucleus-radioactivity}).
\end{solution}

\begin{solution}{exo:g11:fundamental-interactions:13}
Equilibrium: vertically $T\cos\theta = mg$, horizontally
$T\sin\theta = F$, so $F = mg\tan\theta$. Here
$\sin\theta = 3.0/50 = 0.060$, and for small angles
$\tan\theta \approx \sin\theta = 0.060$. Then
$F = \num{1.0e-3} \times 9.81 \times 0.060 \approx \qty{5.9e-4}{N}$;
$q = d\sqrt{F/k} = 0.060 \times \sqrt{\num{5.9e-4}/\num{8.99e9}}
\approx \qty{1.5e-8}{C}$; $N = \num{1.5e-8}/\num{1.6e-19} \approx
\num{9.6e10}$ elementary charges.
\end{solution}

\begin{solution}{exo:g11:fundamental-interactions:14}
$F = \num{6.67e-11} \times \num{5.97e24} \times \num{7.35e22} /
(\num{3.84e8})^2 = \num{2.93e37}/\num{1.47e17} \approx \qty{2.0e20}{N}$.
Setting $kQ^2/d^2 = F$:
$Q = d\sqrt{F/k} = \num{3.84e8} \times \sqrt{\num{2.0e20}/\num{8.99e9}}
\approx \qty{5.7e13}{C}$. That is
$N = \num{5.7e13}/\num{1.6e-19} \approx \num{3.6e32}$ electrons, of
mass $\num{3.6e32} \times \num{9.11e-31} \approx \qty{3.2e2}{kg}$: a
few hundred kilograms of electrons could do the work of
$\qty{7.35e22}{kg}$ of Moon --- electricity's strength and gravity's
weakness in one number.
\end{solution}

\begin{solution}{exo:g11:fundamental-interactions:15}
$F = \num{2.30e-28}/(\num{2.8e-10})^2 = \num{2.30e-28}/\num{7.84e-20}
\approx \qty{2.9e-9}{N}$; weight
$P = \num{5.9e-26} \times 9.81 \approx \qty{5.8e-25}{N}$; ratio
$F/P \approx \num{5e15}$. Each bond has a \emph{fixed} strength, but
weight grows with the whole mass above: pile up rock and the load on
the bottom layer grows without limit while its bonds do not. Around
\qty{e4}{m} of rock the load crushes the electric bonds --- gravity
wins not by strength per particle but by \emph{adding up} when charge
cancels.
\end{solution}

\begin{solution}{pb:g11:fundamental-interactions:1}
\textbf{1.} Quark $< \qty{e-18}{m}$; nucleon $\qty{e-15}{m}$; nucleus
$10^{-15}$--$\qty{e-14}{m}$; atom $\qty{e-10}{m}$; molecule
$\qty{e-9}{m}$; everyday object $\qty{1}{m}$; planet $\qty{e7}{m}$;
planetary system $\qty{e11}{m}$; galaxy $\qty{e21}{m}$.

\textbf{2.} Scale factor $10^{5}$: the atom is
$\qty{1}{cm} \times 10^{5} = \qty{1}{km}$ wide.

\textbf{3.} $(10^{-15}/10^{-10})^3 = 10^{-15}$: matter is
$\num{99.9999999999999}\,\%$ empty.

\textbf{4.} $\num{7e-5}/\num{e-10} = \num{7e5}$ atoms --- about a
million across one hair.

\textbf{5.} From $10^{-18}$ to $10^{21}$: $39$ powers of ten.

\textbf{6.} $F_E = \num{8.99e9} \times (\num{1.60e-19})^2 /
(\num{5.3e-11})^2 \approx \qty{8.2e-8}{N}$.

\textbf{7.} $F_G = \num{6.67e-11} \times \num{1.67e-27} \times
\num{9.11e-31}/(\num{5.3e-11})^2 \approx \qty{3.6e-47}{N}$; ratio
$F_E/F_G \approx \num{2.3e39}$. Electricity holds the atom; gravity is
a spectator.

\textbf{8.} Without gravity: nothing measurable changes (it is $10^{39}$
below). Without electricity: no atoms, no molecules, no matter at all.

\textbf{9.} ``Contact'' is the electric repulsion between the electron
clouds of atoms brought to about $\qty{e-10}{m}$: nothing ever touches.
Floor reaction and friction are Coulomb's law in bulk.

\textbf{10.} $q = \num{e23} \times \num{e-9} \times \num{1.6e-19} =
\qty{1.6e-5}{C}$ each;
$F = \num{8.99e9} \times (\num{1.6e-5})^2 \approx \qty{2.3}{N}$ --- a
plainly visible force from one electron in a billion. Everyday matter
is neutral to far better than $10^{-9}$.

\textbf{11.} The electromagnetic interaction. The strong force is out
of range beyond about $\qty{e-14}{m}$ --- $10^{4}$ times smaller than a
single atom.

\textbf{12.} $F = \num{8.99e9} \times (\num{1.60e-19})^2 /
(\num{1.0e-15})^2 \approx \qty{2.3e2}{N}$.

\textbf{13.} $a = F/m_p = 230/\num{1.67e-27} \approx
\qty{1.4e29}{m/s^2}$, about $\num{1.4e28}$ times $g$: no everyday force
comes close.

\textbf{14.} Stronger than \qty{230}{N} at $\qty{e-15}{m}$ (nuclei
exist); acting on protons and neutrons alike, blind to charge (neutrons
are gripped); range about $\qty{e-15}{m}$ (nuclei stay tiny --- the
glue cannot reach further).

\textbf{15.} $F = \num{2.30e-28}/(\num{1.4e-14})^2 \approx
\qty{1.2}{N}$. No: $14$ times the range. Repulsion acts across
\emph{all} $\frac{92\times91}{2} = 4186$ proton pairs; the glue binds
only nearest neighbours, so growth favours repulsion. Neutrons add glue
and spacing without adding repulsion --- until, beyond about $80$
protons, no mixture balances and the nucleus decays.

\textbf{16.} The weak interaction turns a neutron into a proton (and
back), causing $\beta$ decay --- it transforms; it never binds.

\textbf{17.} $F = \num{6.67e-11} \times \num{5.97e24} \times
\num{7.35e22}/(\num{3.84e8})^2 \approx \qty{2.0e20}{N}$.

\textbf{18.} $Q_E = \num{1.8e51} \times \num{1.6e-19} =
\qty{2.9e32}{C}$, $Q_M = \qty{3.5e30}{C}$. Matching forces means
$f^2\,k\,Q_E Q_M = G M_E M_M$:
\[
f = \sqrt{\frac{\num{2.93e37}}{\num{8.99e9} \times \num{2.9e32}
\times \num{3.5e30}}} = \sqrt{\num{3.2e-36}} \approx \num{1.8e-18} :
\]
an imbalance of \emph{two parts in $10^{18}$} would cancel gravity.

\textbf{19.} Charge comes in two signs and cancels (question 18 shows
how perfectly), so the stronger force switches itself off at large
scales; mass has one sign, adds forever, and nothing screens it.

\textbf{20.} Nucleus: the strong interaction outmuscles the protons'
repulsion at $\qty{e-15}{m}$. Atom to mountain: the electromagnetic
interaction binds electrons to nuclei and atoms to each other. Planet
and up: gravitation, unopposed once charge cancels, holds planets,
orbits and galaxies. The world does not collapse because each floor is
propped by a binder stiffer than the load, and does not fly apart
because matter is neutral to about $10^{-18}$ --- so every scale has
exactly one force in charge, and it is attractive where it must be.
\end{solution}
