\section*{Chapter \ref{ch:g11:forces-and-motion} --- Forces and Motion}

\begin{solution}{exo:g11:forces-and-motion:1}
(a) Constant velocity: net force zero. (b) Speed changes: nonzero,
opposite the motion. (c) Direction changes (velocity is a vector):
nonzero, toward the inside of the bend. (d) Constant velocity: zero ---
weight and drag compensate.
\end{solution}

\begin{solution}{exo:g11:forces-and-motion:2}
$d_1/d_2 = 1.0/3.0$ with $d_1 + d_2 = \qty{0.80}{m}$:
$d_1 = \qty{0.20}{m}$ from the \qty{3.0}{kg} ball. Equal masses: the
middle, \qty{0.40}{m} from each.
\end{solution}

\begin{solution}{exo:g11:forces-and-motion:3}
Same $F\Delta t$, triple mass: $\Delta v$ three times smaller loaded.
To match the empty trolley's $\Delta v$, push three times harder (or
three times longer).
\end{solution}

\begin{solution}{exo:g11:forces-and-motion:4}
$\Delta \vect v$: \qty{2}{m/s}, directed \emph{backward} (from
\num{10} to \qty{8}{m/s}). The net force points the same way:
opposite $\vect v$ --- case 2, slowing down.
\end{solution}

\begin{solution}{exo:g11:forces-and-motion:5}
(a) Static, opposite your push (no sliding yet). (b) Kinetic, opposite
the sliding. (c) Static, pointing \emph{up} the slope --- against the
slide that gravity is trying to start.
\end{solution}

\begin{solution}{exo:g11:forces-and-motion:6}
$\Delta v \propto F\Delta t/m$: (a) \qty{8.0}{m/s}; (b) \qty{2.0}{m/s};
(c) \qty{2.0}{m/s}; (d) $4.0 \times \frac{3 \times 2}{6} =
\qty{4.0}{m/s}$ --- the three changes cancel.
\end{solution}

\begin{solution}{exo:g11:forces-and-motion:7}
$v' = \sqrt{8.0^2 + 6.0^2} = \qty{10.0}{m/s}$, at $\tan\theta =
6.0/8.0$, i.e.\ $\theta \approx 37^\circ$ north of east. Velocities add
as \emph{vectors}: perpendicular contributions combine by Pythagoras,
not by addition of magnitudes.
\end{solution}

\begin{solution}{exo:g11:forces-and-motion:8}
Fixed beforehand: the ball's mass and its $\Delta v$ (incoming speed to
zero), hence the product $F\Delta t$. Stretching $\Delta t$ from
\qty{0.02}{s} to \qty{0.08}{s} divides the average force by $4$.
\end{solution}

\begin{solution}{exo:g11:forces-and-motion:9}
The static friction of the track on the shoe: the foot pushes backward
on the ground, and the mutual contact force pushes the sprinter
forward. Spikes raise the maximum static friction (no slip at full
effort). On frictionless ice no external horizontal force exists, so
the velocity of the center of mass cannot change --- no start, no walk.
\end{solution}

\begin{solution}{exo:g11:forces-and-motion:10}
A rolling wheel's contact point does not slide: static friction,
stronger than the kinetic friction of a skid
(\cref{def:g11:forces-and-motion:friction}) --- shorter stop. A
skidding tyre's kinetic friction points opposite the sliding, whatever
the wheels' angle: turning the wheel changes nothing, steering is lost.
\end{solution}

\begin{solution}{exo:g11:forces-and-motion:11}
The tension of the wire, pointing inward to the athlete. After release
the ball flies off in a straight line along the tangent, at the speed
it had (inertia). The Moon's ``wire'' is the Earth's gravitational
pull; switch gravity off and it leaves along its tangent, straight and
uniform, forever.
\end{solution}

\begin{solution}{exo:g11:forces-and-motion:12}
From the top: $x = \dfrac{1.0 \times 0.60 + 2.0 \times 1.20}{3.0}
= \qty{1.00}{m}$. Distances to the two part-centers: \qty{0.40}{m}
(stick) and \qty{0.20}{m} (head), ratio $2 : 1$ --- the inverse of the
mass ratio $1.0 : 2.0$, as \cref{def:g11:forces-and-motion:com}
requires.
\end{solution}

\begin{solution}{exo:g11:forces-and-motion:13}
\emph{1.} Same $F$ and $\Delta v$, double $m$: the van needs twice the
time. \emph{2.} Same average speed for twice the time: twice the
distance. \emph{3.} Double $\Delta v$ at the same $F$ and $m$: twice
the time; but the average speed also doubles, so the distance is
multiplied by $4$. Doubling your speed quadruples your braking
distance.
\end{solution}

\begin{solution}{exo:g11:forces-and-motion:14}
\emph{1.} $T = 27.3 \times \qty{86400}{s} \approx \qty{2.36e6}{s}$, so
$v = \dfrac{2\pi \times \num{3.84e8}}{\num{2.36e6}} \approx
\qty{1.0e3}{m/s} \approx \qty{1}{km/s}$. \emph{2.} Toward the Earth
(inward); the Earth's gravitational pull. \emph{3.} The Moon \emph{is}
falling: its velocity change points earthward at every instant. Its
tangential speed merely ensures that it keeps missing --- it falls
\emph{around} the Earth instead of into it.
\end{solution}

\begin{solution}{exo:g11:forces-and-motion:15}
Same $m$ and $\Delta v$, so $F\Delta t$ is fixed: $\Delta t$ fifty
times longer on foam means an average force fifty times smaller.
Bending the knees stretches the landing's $\Delta t$ from a jolt to a
long flex; a crash mat is, mechanically speaking, made of extra
milliseconds.
\end{solution}

\begin{solution}{pb:g11:forces-and-motion:1}
\textbf{1.} $50/3.6 \approx \qty{13.9}{m/s}$ (call it \qty{14}{m/s}).
\textbf{2.} $\Delta v \approx \qty{14}{m/s}$: the occupant moves at
\qty{14}{m/s} before and at zero after --- both ends are fixed by the
crash, so no device can touch $\Delta v$.
\textbf{3.} $\Delta v \propto F\Delta t/m$
(\cref{thm:g11:forces-and-motion:secondlaw}) with $\Delta v$ and $m$
fixed forces $F\Delta t$ to be fixed. The designer's only dial is
$\Delta t$.
\textbf{4.} $0.150/0.050 = 3$: the crumple zone divides the average
force by $3$.
\textbf{5.} Backward, opposite the motion --- the velocity shrinks from
\qty{14}{m/s} to zero.
\textbf{6.} The folding front is a machine for making the stop last
longer: metal that crumples buys milliseconds.
\textbf{7.} Same $\Delta v$ and $m$, $\Delta t$ three times shorter:
the vintage dummy takes three times the force.
\textbf{8.} Average speed $\approx \qty{6.9}{m/s}$; distance $\approx
6.9 \times 0.150 \approx \qty{1.0}{m}$ --- just about the metre of
deformable car a real crumple zone provides.
\textbf{9.} A \qty{10}{m} nose is undrivable and unparkable, and its
own mass would worsen every crash. The passenger cell must not
crumple: it preserves the survival space.
\textbf{10.} ``\dots so you must choose $\Delta t$.''
\textbf{11.} $0.150/0.005 = 30$: the windshield stop is thirty times
more violent than the belted one.
\textbf{12.} A slightly stretching belt lengthens the occupant's own
$\Delta t$ a little more (and spreads the force); a steel harness
would impose the car body's shortest stop on the softest passenger.
\textbf{13.} $0.080/0.010 = 8$: the airbag divides the head's force by
$8$.
\textbf{14.} Same $\Delta v$ and $\Delta t$: the force scales with the
mass, $70/20 = 3.5$ times larger for the adult.
\textbf{15.} Required change: $14/0.150 \approx \qty{93}{m/s}$ each
second --- about $9.5$ times what gravity produces
(\qty{9.8}{m/s} per second). The arms must pull about $9.5$ times the
baby's weight: like holding a \qty{95}{kg} load. No arms can; the baby
flies forward.
\textbf{16.} The belt of an adult routes force to hips and chest; on a
child it would load belly and neck. The child seat applies the
(smaller) force to strong body parts and its padding stretches
$\Delta t$ further.
\textbf{17.} If the cell deforms, the occupants lose their survival
space and are struck by the structure itself. Nothing should hit the
rigid cell directly: the occupants' stops are handled by belt and
airbag (long $\Delta t$), while the cell anchors the crumpling ends.
\textbf{18.} By symmetry both cars stop at the contact plane: each
driver undergoes $\Delta v \approx \qty{14}{m/s}$ over a crumple-zone
stop --- essentially the wall test, not ``double the crash''.
\textbf{19.} The forces are equal and opposite, but
$\Delta v \propto F\Delta t/m$: the small car, with the small $m$,
suffers the large velocity change.
\textbf{20.} Crumple zone, stretching seat belt, airbag, child-seat
padding --- every one of them stretches the same quantity: $\Delta t$.
\textbf{21.} The crash fixes the mass and the velocity change, hence
the product force $\times$ time: a gentle collision is a long one ---
stretch the time, and the force falls in the same ratio.
\end{solution}
