\chapter{Color and Light Sources}\label{ch:g11:color-light-sources}

Ask a physicist for the color of a tomato and you get a counter-question:
under which light? The tomato only diffuses what it is given, the lamp
only gives what its spectrum contains, and the eye compresses whatever
arrives into three numbers. Color is a negotiation between source,
object and eye --- this chapter interrogates all three parties in turn.

\section{Seeing in three numbers}

\begin{definition}[Cones and trichromatic vision]\label{def:g11:color-light-sources:cones}
The retina (\cref{ch:g11:lenses-and-eye}) carries two families of
light-sensitive cells: the \emph{rods}\index{rod}, which respond to
faint light but ignore wavelength (night vision, in gray), and the
\emph{cones}\index{cone}, in three types most sensitive near
$\qty{420}{nm}$ (blue), $\qty{530}{nm}$ (green) and $\qty{560}{nm}$
(red). Human color vision is
\emph{trichromatic}\index{trichromatic vision}: whatever the spectrum,
the brain receives only three cone responses, and color is what it
makes of that triple.
\end{definition}

\begin{definition}[Color as a perception]\label{def:g11:color-light-sources:perception}
Color is therefore a \emph{perception}, not a property of light alone:
a spectrum specifies an intensity at every wavelength, the eye keeps
three numbers. Lights with different spectra but the same three cone
responses look strictly the same: they are \emph{metamers}\index{metamer}.
\end{definition}

\begin{example}[Two yellows]\label{ex:g11:color-light-sources:yellows}
The $\qty{589}{nm}$ line of a sodium lamp excites the red and green
cones almost equally: the brain says \emph{yellow}. A screen lighting
its red ($\qty{610}{nm}$) and green ($\qty{540}{nm}$) subpixels gives
the same yellow with no light near $\qty{589}{nm}$. A prism separates
the two instantly; the eye never can.
\end{example}

\section{Adding lights}

\begin{definition}[Additive synthesis]\label{def:g11:color-light-sources:additive}
Superposing light beams on a white screen adds their spectra: this is
\emph{additive synthesis}\index{additive synthesis}. Three well-chosen
lights, red, green and blue --- the \emph{additive
primaries}\index{primary colors} --- mix into almost every perceivable
color: R $+$ G $=$ yellow, R $+$ B $=$ magenta, G $+$ B $=$ cyan, and
R $+$ G $+$ B $=$ white.
\end{definition}

\begin{proposition}[Complementary colors]\label{prop:g11:color-light-sources:complementary}
Two lights whose sum is perceived as white are
\emph{complementary}\index{complementary colors}. The three pairs
yellow/blue, cyan/red and magenta/green are complementary.
\end{proposition}

\begin{proof}
Yellow is R $+$ G; adding blue completes the triple to white. The other
two pairs are the same argument, one primary at a time.
\end{proof}

\begin{example}[Screens]\label{ex:g11:color-light-sources:screen}
Each pixel of a phone screen is three subpixels --- red, green, blue,
under $\qty{0.1}{mm}$ wide --- too close for the eye to separate, so
their lights add on the retina: yellow is R and G lit, orange is R at
full power and G at half, white is all three. A magnifying glass on a
white area reveals the trick: no white anywhere, only R, G, B.
\end{example}

\begin{omfigure}
\begin{tikzpicture}[scale=0.72, font=\small]
  \fill[red] (90:0.8) circle (1.25);
  \fill[green!70!black] (210:0.8) circle (1.25);
  \fill[blue] (330:0.8) circle (1.25);
  \begin{scope}\clip (90:0.8) circle (1.25);\fill[yellow] (210:0.8) circle (1.25);\end{scope}
  \begin{scope}\clip (90:0.8) circle (1.25);\fill[magenta] (330:0.8) circle (1.25);\end{scope}
  \begin{scope}\clip (210:0.8) circle (1.25);\fill[cyan] (330:0.8) circle (1.25);\end{scope}
  \begin{scope}\clip (90:0.8) circle (1.25);\clip (210:0.8) circle (1.25);
    \fill[white] (330:0.8) circle (1.25);\end{scope}
  \node at (90:2.5) {R};\node at (210:2.5) {G};\node at (330:2.5) {B};
\end{tikzpicture}\qquad
\begin{tikzpicture}[scale=0.72, font=\small]
  \fill[cyan] (90:0.8) circle (1.25);
  \fill[magenta] (210:0.8) circle (1.25);
  \fill[yellow] (330:0.8) circle (1.25);
  \begin{scope}\clip (90:0.8) circle (1.25);\fill[blue] (210:0.8) circle (1.25);\end{scope}
  \begin{scope}\clip (90:0.8) circle (1.25);\fill[green!70!black] (330:0.8) circle (1.25);\end{scope}
  \begin{scope}\clip (210:0.8) circle (1.25);\fill[red] (330:0.8) circle (1.25);\end{scope}
  \begin{scope}\clip (90:0.8) circle (1.25);\clip (210:0.8) circle (1.25);
    \fill[black] (330:0.8) circle (1.25);\end{scope}
  \node at (90:2.5) {C};\node at (210:2.5) {M};\node at (330:2.5) {Y};
\end{tikzpicture}

{\small Additive synthesis of overlapping light beams (left) and
subtractive synthesis of stacked filters in white light (right).}
\end{omfigure}

\section{Subtracting from white light}

\begin{definition}[Subtractive synthesis]\label{def:g11:color-light-sources:subtractive}
A \emph{filter}\index{filter} (colored glass, a gel, a layer of ink or
paint) transmits part of the spectrum it receives and absorbs the rest.
Producing colors by \emph{removing} bands from white light is
\emph{subtractive synthesis}\index{subtractive synthesis}. Its
primaries, complements of the additive ones, each subtract one band:
cyan $=$ white $-$ R, magenta $=$ white $-$ G, yellow $=$ white $-$ B.
Stacked filters subtract jointly: cyan $+$ yellow leaves green, cyan
$+$ magenta blue, magenta $+$ yellow red, all three black. Printing
and painting work this way, with C, M, Y inks.
\end{definition}

\begin{method}[Spectrum bookkeeping]\label{met:g11:color-light-sources:bookkeeping}
To predict a color through any chain of sources, filters and objects,
track the three primary bands: write the source's light as its content
in R, G, B; at each filter or surface, delete the absorbed bands; name
the color of what survives (nothing left $=$ black).
\end{method}

\begin{example}[A filter chain]\label{ex:g11:color-light-sources:chain}
White light $(R, G, B)$ crosses a yellow filter: the blue band is
absorbed, leaving $(R, G)$ --- yellow. A cyan filter then absorbs the
red band, leaving $(G)$: green. Swapping the filters changes nothing:
each absorbs its own band regardless of order.
\end{example}

\section{The color of an object}

\begin{definition}[Diffusion and object color]\label{def:g11:color-light-sources:objectcolor}
An opaque object absorbs part of the light it receives and
\emph{diffuses}\index{diffusion} the rest --- re-emits it in every
direction, toward our eyes among others. The \emph{color of an
object}\index{color of an object} is the color of the light it diffuses
\emph{under the given illumination}: the object supplies its absorption
habits, the source the raw material. Change the lamp, change the color.
\end{definition}

\begin{example}[The red shirt]\label{ex:g11:color-light-sources:redshirt}
A shirt looks red in daylight: it absorbs the green and blue bands and
diffuses the red one. Under a pure green light there is no red band to
diffuse: the shirt absorbs everything and appears black. Under magenta
light $(R, B)$ it diffuses the red band alone: still red. A white
object diffuses every band it receives (under green light it looks
green); a black object absorbs them all under every light --- which is
also why black clothing heats up in the sun.
\end{example}

\begin{omfigure}
\begin{tikzpicture}[scale=0.9, font=\small]
  \fill[red!80] (0,0) rectangle (1.6,1.0);
  \node at (0.8,0.5) {shirt};
  \draw[-Stealth, red, thick] (-1.9,1.45) -- (-0.1,0.9);
  \draw[-Stealth, green!65!black, thick] (-1.9,0.95) -- (-0.1,0.62);
  \draw[-Stealth, blue, thick] (-1.9,0.45) -- (-0.1,0.34);
  \draw[-Stealth, red, very thick] (1.7,0.9) -- (3.1,1.45)
    node[above right=-3pt] {red diffused};
  \node[anchor=west] at (1.7,0.3) {G, B absorbed};
  \node at (0.8,-0.55) {white light: red shirt};
\end{tikzpicture}\qquad
\begin{tikzpicture}[scale=0.9, font=\small]
  \fill[black!85] (0,0) rectangle (1.6,1.0);
  \node[white] at (0.8,0.5) {shirt};
  \draw[-Stealth, green!65!black, thick] (-1.9,0.95) -- (-0.1,0.62);
  \node[anchor=west] at (1.7,0.75) {nothing diffused};
  \node[anchor=west] at (1.7,0.3) {G absorbed};
  \node at (0.8,-0.55) {green light: black shirt};
\end{tikzpicture}

{\small The same red shirt under two illuminations: it can only diffuse
the red band --- present in white light, absent from green light.}
\end{omfigure}

\section{Light sources and color temperature}

\begin{definition}[Incandescent sources]\label{def:g11:color-light-sources:incandescent}
A source glowing because it is hot --- flame, filament, star --- shines
by \emph{incandescence}\index{incandescence}. Its spectrum is the
continuous thermal spectrum of \cref{ch:g10:light-spectra}, and its peak
obeys Wien's law (\cref{prop:g10:light-spectra:wien}),
\[
\lambda_{\max} \times T = \qty{2.90e-3}{m.K},
\]
our quantitative tool from now on: peak measured, temperature known.
\end{definition}

\begin{example}[A very inefficient lamp]\label{ex:g11:color-light-sources:bulb}
A tungsten filament at $T = \qty{2700}{K}$ peaks at
$\lambda_{\max} = \qty{2.90e-3}{m.K}/\qty{2700}{K} = \qty{1.07e-6}{m}
= \qty{1070}{nm}$, deep in the infrared: most of the electric power
leaves as invisible radiation, only about $5\%$ as light --- the
incandescent bulb was mainly a heater.
\end{example}

\begin{omfigure}
\begin{tikzpicture}
\begin{axis}[omaxis, width=9.4cm, height=5.4cm,
    xlabel={$\lambda$ (\unit{nm})}, ylabel={intensity (arbitrary units)},
    xmin=0, xmax=2400, ymin=0, ymax=10,
    xtick={400, 800, 1200, 1600, 2000}, ytick=\empty]
  \fill[gray, opacity=0.12] (axis cs:400,0) rectangle (axis cs:800,10);
  \node[gray, anchor=south] at (axis cs:600,8.7) {visible};
  \addplot[omThm, thick, domain=200:2300, samples=160]
    {3.2e16 * x^(-5) / (exp(2400/x) - 1)};
  \addplot[omDef, thick, domain=250:2300, samples=160]
    {3.84e17 * x^(-5) / (exp(4800/x) - 1)};
  \draw[dashed, omThm] (axis cs:483,0) -- (axis cs:483,8.5);
  \draw[dashed, omDef] (axis cs:967,0) -- (axis cs:967,3.2);
  \node[omThm, anchor=west] at (axis cs:520,9.0) {$\qty{6000}{K}$};
  \node[omDef, anchor=west] at (axis cs:1010,3.4) {$\qty{3000}{K}$ ($\times 12$)};
\end{axis}
\end{tikzpicture}

{\small Thermal spectra at $\qty{6000}{K}$ and $\qty{3000}{K}$, peaks
$\lambda_{\max}$ dashed at $483$ and $\qty{967}{nm}$; the cooler curve,
magnified twelve times, would otherwise hug the axis.}
\end{omfigure}

\begin{definition}[Spectral lamps and lasers]\label{def:g11:color-light-sources:lamps}
A \emph{spectral lamp}\index{spectral lamp} --- a low-pressure gas
excited by a discharge: sodium, mercury, neon --- emits a line spectrum
(\cref{ch:g10:light-spectra}): a few isolated wavelengths, nothing in
between. A \emph{laser}\index{laser} goes further: one single
wavelength, monochromatic light, in one narrow beam.
\end{definition}

\begin{definition}[LEDs]\label{def:g11:color-light-sources:led}
A \emph{light-emitting diode}\index{light-emitting diode} (LED) emits a
\emph{quasi-monochromatic}\index{quasi-monochromatic} band a few tens of nanometers wide: narrow
enough to look a definite color, far broader than a laser line. A white
LED is a blue LED (peak near $\qty{450}{nm}$) coated with a
\emph{phosphor}\index{phosphor}, a material absorbing part of the blue
and re-emitting it as a broad yellow band: blue and yellow being
complementary (\cref{prop:g11:color-light-sources:complementary}), the
mixture reads as white.
\end{definition}

\begin{definition}[Color temperature]\label{def:g11:color-light-sources:colortemp}
The \emph{color temperature}\index{color temperature} of a white-ish
source is the temperature of the incandescent body whose hue its light
matches: candle $\qty{1800}{K}$, incandescent bulb $\qty{2700}{K}$,
halogen $\qty{3400}{K}$, midday sun $\qty{5500}{K}$, overcast daylight
$\qty{6500}{K}$. It is a color label, not a thermometer reading --- a
$\qty{6500}{K}$ LED runs at room temperature --- and vocabulary runs
backwards: low color temperature reads ``warm'' orange, high ``cool''
bluish.
\end{definition}

\begin{omfigure}
\begin{tikzpicture}[font=\small]
  \shade[left color=orange!95!red, right color=white] (0,0) rectangle (5.4,0.7);
  \shade[left color=white, right color=blue!25] (5.4,0) rectangle (9,0.7);
  \draw (0,0) rectangle (9,0.7);
  \draw[-Stealth] (0,-0.5) -- (9.6,-0.5)
    node[right] {color temperature (\unit{K})};
  \foreach \x/\l in {0.5/1800, 2.0/2700, 3.1/3400, 6.5/5500, 8.2/6500}
    \draw (\x,-0.4) -- (\x,-0.6) node[below] {\l};
  \foreach \x/\n in {0.5/candle, 2.0/bulb, 3.1/halogen, 6.5/sun,
      8.2/overcast}
    \node[above] at (\x,0.7) {\n};
\end{tikzpicture}

{\small The color-temperature scale of common sources: ``warm'' orange
hues sit at low $T$, ``cool'' bluish hues at high $T$.}
\end{omfigure}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:g11:color-light-sources:1}
Give the color perceived when a white screen receives simultaneously:
(a) red and green light; (b) green and blue light; (c) all three
primaries. Which light is complementary to blue?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g11:color-light-sources:2}
Which subpixels (R, G, B) does a screen light up to display: yellow,
magenta, white, black, orange?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g11:color-light-sources:3}
White light crosses a magenta filter. Which band is absorbed, and what
color emerges? A cyan filter is then added behind it: what emerges now?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g11:color-light-sources:4}
Under white light a flag shows a yellow stripe. Which bands does it
diffuse, which does it absorb? Its color under (a) red; (b) blue light?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g11:color-light-sources:5}
A star's continuous spectrum peaks at $\lambda_{\max} = \qty{580}{nm}$.
Compute its surface temperature.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:color-light-sources:6}
A candle flame sits at about $\qty{1800}{K}$. (a) Compute
$\lambda_{\max}$ and give its domain. (b) The flame nevertheless looks
orange: explain.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:color-light-sources:7}
A cyclist wears a red jacket and cyan gloves. Using
\cref{met:g11:color-light-sources:bookkeeping}, give the apparent color
of each item under (a) white light; (b) red light; (c) green light.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:color-light-sources:8}
An old street is lit by low-pressure sodium lamps (a single line at
$\qty{589}{nm}$). (a) What type of spectrum is this, and what colors
can \emph{any} object take under it? (b) A car looks black under these
lamps and blue in daylight: explain.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:color-light-sources:9}
A green laser pointer emits at $\qty{532}{nm}$. (a) Its beam crosses a
prism: compare the outcome with that of a white beam, and name the
property illustrated. (b) It then meets a red filter: what comes out?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:color-light-sources:10}
Describe the spectrum of a white LED (blue diode plus phosphor), and
explain why its light is perceived as white. Why does a deep red object
look duller under a cheap white LED than in daylight?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:color-light-sources:11}
Compute $\lambda_{\max}$ for an incandescent body at (a) a candle's
color temperature ($\qty{1800}{K}$); (b) overcast daylight's
($\qty{6500}{K}$). Which of the two lights is called ``warm''? Comment.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g11:color-light-sources:12}
A sodium lamp ($\qty{589}{nm}$) and a screen displaying yellow
(subpixels at $\qty{610}{nm}$ and $\qty{540}{nm}$) look exactly the same
color. (a) What are two such lights called, and what does their
existence prove about color? (b) Propose two different experiments that
tell them apart, and predict the outcome of each.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g11:color-light-sources:13}
A theater halogen spot runs at $\qty{3200}{K}$. (a) Compute
$\lambda_{\max}$: is this light warmer or cooler than daylight? (b) To
imitate daylight, the spot is covered with a pale blue gel. What does
the gel remove, and why does the stage get dimmer? (c) Explain why no
gel could instead \emph{strengthen} the blue end of the spot's spectrum.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g11:color-light-sources:14}
A butcher lights the meat display with LEDs of color temperature
$\qty{2700}{K}$; the fishmonger next door uses $\qty{6500}{K}$.
(a) Describe the hue of each lighting and its effect on red meat and on
white fish. (b) Has either merchant changed his goods? Say exactly
where the change happens (\cref{def:g11:color-light-sources:objectcolor}).
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g11:color-light-sources:15}
A town next to an observatory must relight its streets and hesitates
between low-pressure sodium lamps and broad-spectrum white LEDs.
(a) Which choice do the astronomers prefer, and why? (Think of what a
single filter can remove from their images.) (b) Which choice do
drivers and pedestrians prefer? Justify with object colors.
(c) Propose a compromise, phrased with color temperature.
\end{exercise}

\section{Problem: Engineering color}

\begin{problem}[{Weekend problem --- stage lights, sodium streets and
the white LED: how color is engineered, from three spotlights that can
paint any hue to the chip that lights this page}]
\label{pb:g11:color-light-sources:1}
A lighting designer owns only red, green and blue spotlights; a street
engineer owns a lamp that emits one single wavelength; a chip maker owns
a diode that only shines blue. All three sell color. This problem runs
their trades in turn --- additive mixing on stage, monochrome economy in
the street, phosphor alchemy in the chip --- and ends by choosing the
light for a bedroom and for a jeweler's window.

\medskip
\noindent\textbf{Part I --- The stage: painting with three spotlights.}
\begin{enumerate}
  \item The three spots overlap on a white backdrop. Name the color of
        each pairwise overlap and of the triple overlap.
  \item How does the designer make orange? And a pale pink? (Describe
        the spot powers qualitatively.)
  \item A dancer stands before the backdrop, lit by the red and green
        spots only. Two colored shadows appear: color of each? Explain.
  \item The backdrop is repainted deep blue. What does it look like
        under the red spot alone? Under all three?
  \item Why can three spots reproduce ``almost every'' color? Which
        piece of anatomy fixes the number three?
\end{enumerate}

\medskip
\noindent\textbf{Part II --- Costumes: subtraction at work.}
\begin{enumerate}[resume]
  \item A costume is yellow in white light. Which band does its dye
        absorb?
  \item Predict its color under the red spot, the green spot, the blue
        spot.
  \item The designer wants the costume to flip from bright to black at
        one flick of a switch. Which spot does the flick switch to?
  \item A magenta gel is slid in front of a white spot lighting the
        yellow costume. Track the bands
        (\cref{met:g11:color-light-sources:bookkeeping}): costume color?
  \item Why do stage gels come in cyan, magenta and yellow rather than
        red, green and blue? (What would a red gel do to a white spot's
        power?)
\end{enumerate}

\medskip
\noindent\textbf{Part III --- Sodium streets: the monochrome economy.}
\begin{enumerate}[resume]
  \item A low-pressure sodium lamp emits essentially one line at
        $\qty{589}{nm}$. What type of spectrum is this, and what does
        the whole street look like under it?
  \item Under this lamp, give the apparent color of a white wall, a
        yellow road marking, a red car, a blue car.
  \item Why did astronomers love these lamps? (What can one filter do
        to the whole city's glow?)
  \item A tungsten street bulb at $\qty{2700}{K}$ is proposed instead.
        Compute its $\lambda_{\max}$: where does most of its power go?
  \item The sodium lamp converts about $30\%$ of its electric power
        into visible light, the tungsten bulb about $5\%$. How many
        watts of light does each deliver per $\qty{100}{W}$ of power?
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- The white LED and the choice of a light.}
\begin{enumerate}[resume]
  \item A white LED is a blue diode ($\qty{450}{nm}$) under a yellow
        phosphor. Explain, with complementary colors, why its output
        reads as white.
  \item Describe its spectrum and compare it with the continuous
        spectrum of daylight. Which band is under-represented?
  \item Deduce which objects such an LED renders poorly, and connect
        this with the butcher of \cref{exo:g11:color-light-sources:14}.
  \item An LED bulb is sold as ``$\qty{2700}{K}$''. Is anything inside
        it at $\qty{2700}{K}$? What does the label promise? Compute the
        $\lambda_{\max}$ of the incandescent body it imitates.
  \item Finale --- choose, with one sentence of justification each, a
        color temperature for a bedroom lamp and one for a jeweler's
        window; then state the chapter's law of color in one line.
\end{enumerate}
\end{problem}
