\chapter{Electric and Gravitational Fields}\label{ch:g11:electric-gravitational-fields}

How does the Earth pull an apple it never touches? Newton himself called
action at a distance ``so great an absurdity'' that he refused to defend
it. The modern answer: a mass, or a charge, modifies the space around it
--- it creates a \emph{field} --- and whatever is placed there responds
to the field at its own location. We build the two great examples side
by side, until the analogy becomes an identity of form.

\section{From action at a distance to the field}

\begin{definition}[Field]\label{def:g11:electric-gravitational-fields:field}
A \emph{field}\index{field} assigns to every point of space a vector. A
source (a charge, a mass) creates its field everywhere around it,
whether or not anything is there to feel it; a test object then feels a
force determined by the field \emph{at its own location} alone. This
two-step picture replaces action at a distance.
\end{definition}

\begin{remark}
The field is not a bookkeeping trick: it carries energy, and changes in
the source travel outward through it at finite speed --- light itself is
such a traveling field (\cref{ch:g12:light-as-wave}). Here we study
fields that do not change in time.
\end{remark}

\section{The electric field}

\begin{definition}[Electric field]\label{def:g11:electric-gravitational-fields:efield}
If a small test charge $q$ placed at a point $M$ feels an electric force
$\vect F$, the \emph{electric field}\index{electric field} at $M$ is
\[
\vect E = \frac{\vect F}{q}
\qquad \text{(in \unit{N/C}, also written \unit{V/m})}.
\]
$\vect E$ depends only on the sources and on $M$, not on the test
charge: doubling $q$ doubles $\vect F$, leaving the quotient unchanged.
\end{definition}

\begin{proposition}[Force on a charge]\label{prop:g11:electric-gravitational-fields:force}
A charge $q$ placed where the field is $\vect E$ feels the force
$\vect F = q \vect E$: along $\vect E$ if $q > 0$, opposite to $\vect E$
if $q < 0$, of magnitude $F = \abs{q}\,E$.
\end{proposition}

\begin{proof}
Rearrange the definition; the sign rule is the algebra of a negative multiple.
\end{proof}

\begin{example}[Reading a field]\label{ex:g11:electric-gravitational-fields:reading}
A test charge $q = \qty{+2.0e-8}{C}$ at $M$ feels a force of
\qty{6.0e-4}{N} pointing east. The field at $M$ is
$E = \num{6.0e-4} / \num{2.0e-8} = \qty{3.0e4}{N/C}$, pointing east;
$q' = \qty{-1.0e-8}{C}$ placed at the same point feels
$F = \num{1.0e-8} \times \num{3.0e4} = \qty{3.0e-4}{N}$ --- pointing west.
\end{example}

\begin{proposition}[Field of a point charge]\label{prop:g11:electric-gravitational-fields:point}
A point charge $Q$ creates, at distance $d$, a field along the line
joining charge to point --- away from $Q$ if $Q > 0$, toward it if
$Q < 0$ --- of magnitude
\[
E = k\,\frac{\abs{Q}}{d^2},
\qquad k = \qty{8.99e9}{N.m^2/C^2}.
\]
\end{proposition}

\begin{proof}
Coulomb's law (\cref{ch:g11:fundamental-interactions}) gives the force
on a test charge $q$ at distance $d$: magnitude $k \abs{qQ} / d^2$,
along the joining line, repulsive for like signs. Divide by $q$ and
apply \cref{prop:g11:electric-gravitational-fields:force}.
\end{proof}

\begin{example}[A charged sphere]\label{ex:g11:electric-gravitational-fields:sphere}
A Van de Graaff sphere carries $Q = \qty{+5.0e-7}{C}$ (a small charged
sphere acts like a point charge at its center). At $d = \qty{0.50}{m}$:
\[
E = \num{8.99e9} \times \num{5.0e-7} / 0.50^2 \approx \qty{1.8e4}{N/C},
\]
pointing radially away from the sphere. At \qty{1.0}{m} it has dropped
to a quarter of this: the $1/d^2$ of the force survives in the field.
\end{example}

\begin{definition}[Field lines]\label{def:g11:electric-gravitational-fields:lines}
A \emph{field line}\index{field line} is a curve everywhere tangent to
the field, an arrow giving the field's direction. Three rules:
\begin{itemize}
  \item they leave positive charges and arrive at negative charges;
  \item where lines crowd the field is strong; where they spread, weak;
  \item two lines never cross: the field has one direction at each point.
\end{itemize}
\end{definition}

\begin{omfigure}
\begin{tikzpicture}[scale=0.82]
  \fill[omDef!20] (0,0) circle (0.32);
  \draw[omDef, thick] (0,0) circle (0.32);
  \node[font=\small] at (0,0) {$+$};
  \foreach \a in {0,45,...,315}
    \draw[-Stealth, omThm, thick] (\a:0.32) -- (\a:1.5);
\end{tikzpicture}\qquad\qquad
\begin{tikzpicture}[scale=0.82]
  \fill[omProp!20] (0,0) circle (0.32);
  \draw[omProp, thick] (0,0) circle (0.32);
  \node[font=\small] at (0,0) {$-$};
  \foreach \a in {0,45,...,315}
    \draw[-Stealth, omThm, thick] (\a:1.5) -- (\a:0.32);
\end{tikzpicture}

{\small Field lines of a point charge: radially out of $+$, into $-$;
the crowding near the charge pictures the $1/d^2$ growth of the field.}
\end{omfigure}

\section{The uniform field of a capacitor}

\begin{definition}[Parallel-plate capacitor]\label{def:g11:electric-gravitational-fields:capacitor}
Two facing parallel metal plates, close together and carrying opposite
charges, form a \emph{capacitor}\index{capacitor}. Between the plates
the field is \emph{uniform}\index{uniform field}: same magnitude and
direction everywhere --- perpendicular to the plates, from the positive
plate to the negative.
\end{definition}

\begin{proposition}[Field between the plates]\label{prop:g11:electric-gravitational-fields:uniform}
If the voltage between the plates (\cref{ch:g11:circuits-and-power}) is
$U$ and their separation is $d$, the field between them has magnitude
\[
E = \frac{U}{d} \qquad \text{(in \unit{V/m})}.
\]
\end{proposition}
\admitted

\begin{remark}
Uniformity and $E = U/d$ are derived honestly in the Year 1 volume; the
formula explains the unit \unit{V/m}. Thin gaps make strong fields:
\qty{100}{V} across a millimetre is already \qty{e5}{V/m}.
\end{remark}

\begin{example}[Steering an electron beam]\label{ex:g11:electric-gravitational-fields:beam}
In an oscilloscope, the beam passes between plates with
$U = \qty{100}{V}$ and $d = \qty{5.0}{mm}$:
$E = 100 / \num{5.0e-3} = \qty{2.0e4}{V/m}$, and each electron feels
$F = \num{1.60e-19} \times \num{2.0e4} = \qty{3.2e-15}{N}$ toward the
positive plate --- $\num{e14}$ times its weight of \qty{8.9e-30}{N},
so the beam deflects visibly (the curved path is
\cref{ch:g12:newtons-laws}'s business).
\end{example}

\begin{omfigure}
\begin{tikzpicture}[scale=0.9]
  \draw[omDef, very thick] (0,2.2) -- (5.2,2.2)
    node[right, font=\small] {$+$};
  \draw[omProp, very thick] (0,0) -- (5.2,0)
    node[right, font=\small] {$-$};
  \draw[-Stealth, omThm] (0.8,2.05) -- (0.8,0.15);
  \draw[-Stealth, omThm] (1.8,2.05) -- (1.8,0.15);
  \draw[-Stealth, omThm] (2.8,2.05) -- (2.8,0.15)
    node[pos=0.25, right, font=\small] {$\vect E$};
  \draw[-Stealth, omThm] (4.6,2.05) -- (4.6,0.15);
  \fill[omDef] (3.7,1.4) circle (2.5pt)
    node[right=2pt, font=\small] {$q>0$};
  \draw[-Stealth, omMeth, very thick] (3.7,1.4) -- (3.7,0.6)
    node[left, font=\small] {$q\vect E$};
  \draw[Stealth-Stealth, thin] (-0.35,0) -- (-0.35,2.2)
    node[midway, left, font=\small] {$d$};
\end{tikzpicture}

{\small The uniform field of a capacitor: parallel, equally spaced
lines from the $+$ plate to the $-$ plate, $E = U/d$; a positive charge
is pushed along the field, a negative one against it.}
\end{omfigure}

\section{The gravitational field}

\begin{definition}[Gravitational field]\label{def:g11:electric-gravitational-fields:gfield}
If a test mass $m$ placed at a point feels a gravitational force (a
weight) $\vect P$, the \emph{gravitational field}\index{gravitational
field} there is
\[
\vect g = \frac{\vect P}{m} \qquad \text{(in \unit{N/kg})},
\]
and conversely $\vect P = m \vect g$. Grade 10 introduced the number
$g$; $\vect g$ adds the direction a dropped stone starts to move.
\end{definition}

\begin{proposition}[Field of the Earth]\label{prop:g11:electric-gravitational-fields:earth}
At distance $d$ from the Earth's center (mass $M$), the gravitational
field points toward the center and has magnitude
\[
g = G\,\frac{M}{d^2},
\qquad G = \qty{6.67e-11}{N.m^2/kg^2}.
\]
\end{proposition}

\begin{proof}
Universal gravitation (\cref{ch:g10:universal-gravitation}) gives the
force on a test mass $m$: magnitude $G M m / d^2$, toward the center;
divide by $m$. At the surface, $d = R$ and this is the
$g = GM/R^2 \approx \qty{9.81}{N/kg}$ of grade 10.
\end{proof}

\begin{example}[The field where the Moon lives]\label{ex:g11:electric-gravitational-fields:moon}
At the Moon's distance $d = \qty{3.84e8}{m}$:
\[
g = \num{6.67e-11} \times \num{5.97e24} / (\num{3.84e8})^2
\approx \qty{2.7e-3}{N/kg}.
\]
Earth's field never stops --- it only thins out as $1/d^2$; this is the
field that holds the Moon on its orbit.
\end{example}

\begin{remark}[Uniform near the surface]\label{rem:g11:electric-gravitational-fields:uniformg}
Over a laboratory or even a mountain, $d$ barely changes compared with
$R = \qty{6.37e6}{m}$, so $\vect g$ is uniform to high accuracy:
vertical lines, constant \qty{9.81}{N/kg} --- the Earth's field looks
like the capacitor's, just as the round Earth looks flat from a garden.
\end{remark}

\begin{omfigure}
\begin{tikzpicture}[scale=0.9]
  \fill[omExo!15] (0,0) circle (0.85);
  \draw[omExo, thick] (0,0) circle (0.85);
  \node[font=\small] at (0,0) {Earth};
  \foreach \a in {0,60,...,300}
    \draw[-Stealth, omThm, thick] (\a:2.1) -- (\a:0.95);
  \node[omThm, font=\small] at (30:1.9) {$\vect g$};
\end{tikzpicture}

{\small The Earth's gravitational field: radial, pointing at the
center, of magnitude $GM/d^2$ --- the exact portrait of the field of a
negative point charge.}
\end{omfigure}

\section{One idea, two forces}

The two fields of this chapter are the same mathematics in two costumes:

\begin{center}
\begin{tabular}{lcc}
 & \textbf{electric} & \textbf{gravitational} \\
\hline
source & charge $Q$ & mass $M$ \\
field & $\vect E = \vect F / q$ & $\vect g = \vect P / m$ \\
unit & \unit{N/C} = \unit{V/m} & \unit{N/kg} \\
point source & $E = k \abs{Q} / d^2$ & $g = G M / d^2$ \\
constant & $k = \qty{8.99e9}{N.m^2/C^2}$ & $G = \qty{6.67e-11}{N.m^2/kg^2}$ \\
force on a test & $\vect F = q \vect E$ & $\vect P = m \vect g$ \\
field lines & out of $+$, into $-$ & always into the mass \\
\end{tabular}
\end{center}

Every formula on the left becomes the one on the right under the
dictionary $q \leftrightarrow m$, $\vect E \leftrightarrow \vect g$, $k \leftrightarrow G$.

\begin{remark}[Where the analogy breaks]\label{rem:g11:electric-gravitational-fields:noanalogue}
Charge comes in two signs; mass in one. So electric fields can be
canceled: the rearranged charges of a metal box wipe out any external
field inside it (a \emph{Faraday cage} --- why an elevator kills your
phone signal), and bulk matter, with balanced $+$ and $-$, is
electrically silent. Gravity can be neither screened nor neutralized:
every kilogram adds its pull --- which is why the force weaker by
$\num{e39}$ (\cref{exo:g11:electric-gravitational-fields:10}) runs the
universe at large scales.
\end{remark}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:g11:electric-gravitational-fields:1}
A test charge $q = \qty{+2.0e-6}{C}$ at a point $M$ feels a force of
\qty{5.0e-3}{N} pointing north. Give the field at $M$ (magnitude and
direction), then the force on $q' = \qty{-4.0e-6}{C}$ placed at $M$.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g11:electric-gravitational-fields:2}
A point charge $Q = \qty{+5.0}{nC}$ sits at $O$.
\begin{enumerate}
  \item Compute the field at $d = \qty{30}{cm}$: magnitude, direction.
  \item At what distance is the field half as strong?
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g11:electric-gravitational-fields:3}
A capacitor has $U = \qty{600}{V}$ across plates \qty{3.0}{mm} apart.
Compute the field, then the force on one electron there
($e = \qty{1.60e-19}{C}$).
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g11:electric-gravitational-fields:4}
On Mars, a \qty{250}{kg} probe weighs \qty{930}{N}. Compute the Martian
surface field $g_{\text{Mars}}$, then the weight of an \qty{80}{kg}
astronaut there.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g11:electric-gravitational-fields:5}
True or false, with one reason each: (a) two field lines may cross
where the field is strong; (b) field lines leave negative charges and
arrive at positive ones; (c) bunched field lines signal a strong field.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:electric-gravitational-fields:6}
The International Space Station orbits at altitude \qty{420}{km}
($R = \qty{6.37e6}{m}$, $M = \qty{5.97e24}{kg}$).
\begin{enumerate}
  \item Compute $g$ at the station; what fraction of the surface value?
  \item Astronauts float. Reconcile this with your answer in one
        sentence.
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:electric-gravitational-fields:7}
What field magnitude would balance an electron's weight
($m_e = \qty{9.11e-31}{kg}$)? Compare with the \qty{2.0e4}{V/m} of
\cref{ex:g11:electric-gravitational-fields:beam}; conclude about the
place of gravity in electron physics.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:electric-gravitational-fields:8}
A dust grain of mass \qty{2.0e-6}{kg} floats at rest between horizontal
plates \qty{2.0}{cm} apart with $U = \qty{5.0}{kV}$ across them.
\begin{enumerate}
  \item Compute the field, then the magnitude of the grain's charge.
  \item The grain's charge is positive: which plate is the positive one?
  \item How many elementary charges does the grain carry?
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:electric-gravitational-fields:9}
Charges $+Q$ and $+4Q$ sit \qty{30}{cm} apart. Where on the segment
joining them is the total field zero? Explain first why the point must
lie between the charges, closer to the smaller one.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:electric-gravitational-fields:10}
In a hydrogen atom, the proton ($m_p = \qty{1.67e-27}{kg}$, charge
$+e$) and the electron ($m_e = \qty{9.11e-31}{kg}$, charge $-e$) are
$d = \qty{5.3e-11}{m}$ apart. Compute the electric and the
gravitational forces between them, and their ratio.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:electric-gravitational-fields:11}
At a point $M$, one source alone would create a field of \qty{300}{N/C}
pointing east; a second, \qty{400}{N/C} pointing north. Fields add as
vectors: give the total field at $M$ (magnitude and direction), then
the force on $q = \qty{-2.0e-6}{C}$ placed there.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g11:electric-gravitational-fields:12}
At what altitude has the Earth's field dropped to half its surface
value? And at altitude $h = R$, to what fraction?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g11:electric-gravitational-fields:13}
Charges of $\qty{+2.0}{nC}$ sit at $A$ and $B$, \qty{6.0}{cm} apart.
$M$ is on the perpendicular bisector of $[AB]$, \qty{4.0}{cm} from its
midpoint.
\begin{enumerate}
  \item Compute the distance $AM$, then the magnitude of the field each
        charge creates at $M$.
  \item Add the two fields as vectors (use the symmetry) and give the
        total field at $M$, magnitude and direction.
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g11:electric-gravitational-fields:14}
Explain, using the two signs of charge, how the free charges of a metal
box cancel an external electric field everywhere inside it --- and why
no arrangement of masses can do the same for gravity. What impossible
ingredient would a gravity shield require?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g11:electric-gravitational-fields:15}
A small ball ($m = \qty{0.50}{g}$, $q = \qty{+2.0e-7}{C}$) hangs from a
thread in a horizontal uniform field $E = \qty{1.0e4}{V/m}$. At rest,
the thread makes an angle $\theta$ with the vertical.
\begin{enumerate}
  \item List the three forces; explain why $\tan\theta = qE / (mg)$.
  \item Compute $\theta$, then the tension of the thread.
\end{enumerate}
\end{exercise}

\section{Problem: Millikan's droplet}

\begin{problem}[{Weekend problem --- weighing the electron's charge: an
oil droplet parked in mid-air between two plates reveals that charge
comes in indivisible steps of $\num{1.60e-19}$ coulombs}]
\label{pb:g11:electric-gravitational-fields:1}
In 1909, Robert Millikan sprayed oil droplets between the horizontal
plates of a capacitor. By tuning the voltage until a chosen droplet
hung motionless --- electric force exactly balancing weight --- he
could weigh its charge. Our droplet has mass $m = \qty{4.90e-15}{kg}$
(measured from its slow fall with the field off; we take it as given);
the plates are $d = \qty{8.0}{mm}$ apart, the \emph{upper} one
positive; the droplet's charge is negative.

\medskip
\noindent\textbf{Part I --- The stage.}
\begin{enumerate}
  \item Describe the field between the plates; what does ``uniform'' buy?
  \item For $U = \qty{800}{V}$, compute $E$.
  \item In which direction is the electric force on the droplet? Why
        did the \emph{upper} plate have to be the positive one?
  \item Compute the droplet's weight.
  \item Write the at-rest condition as an equality of two force
        magnitudes.
\end{enumerate}

\medskip
\noindent\textbf{Part II --- The balance.}
\begin{enumerate}[resume]
  \item From it, express $\abs{q}$ in terms of $m$, $g$, $d$ and $U$.
  \item The droplet hangs at rest for $U = \qty{800}{V}$: compute
        $\abs{q}$.
  \item Divide by $e = \qty{1.60e-19}{C}$. How many excess electrons
        does the droplet carry?
  \item The voltage is nudged up to \qty{900}{V}. In which direction
        does the net force now point?
  \item Show that the balancing voltage is $U = mgd/\abs{q}$: the
        bigger the charge, the smaller the voltage.
\end{enumerate}

\medskip
\noindent\textbf{Part III --- The steps.}
A burst of radiation can knock electrons onto or off the droplet; after
each change, Millikan re-tunes the voltage to balance.
\begin{enumerate}[resume]
  \item After one burst, balance requires $U = \qty{1200}{V}$. Compute
        the new $\abs{q}$. Did the droplet gain or lose an electron?
  \item Successive balances are found at $U = 1200$, $800$, $600$,
        $480$ and \qty{400}{V}. Compute the five charges.
  \item Show that all five are integer multiples of one quantity; give
        its value.
  \item Between consecutive balances, by how much does the charge
        change? Interpret.
  \item Explain why this droplet can never hang at rest at
        $U = \qty{700}{V}$, however patiently one tunes.
  \item Millikan repeated this on hundreds of droplets: every measured
        charge was an integer multiple of the same value. State the
        conclusion, and name the quantity measured.
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- Why there is no gravitational Millikan.}
\begin{enumerate}[resume]
  \item In this chapter's dictionary, what plays the roles of $q$, $E$
        and $qE$ in the balance of Part II?
  \item Could a mass overhead hold the droplet up instead? Compute the
        field of a \qty{1000}{kg} lead sphere \qty{1.0}{m} above the
        droplet, the force on the droplet, and compare with its weight.
  \item The balance needs an upward pull strong enough to fight the
        whole Earth. Why can a plate of charge provide it and no
        arrangement of masses can?
  \item The droplet's \emph{mass} also changes in steps (one oil
        molecule: about \qty{5e-25}{kg}). Compare the relative jumps in
        $m$ and in $\abs{q}$ when one electron lands, and conclude: why
        does this balance see the atoms of electricity but not the
        molecules of oil? Punchline: charge is quantized, and its atom
        is $e = \qty{1.60e-19}{C}$.
\end{enumerate}
\end{problem}
