\chapter{Magnetism and Magnetic Fields}\label{ch:g11:magnetic-fields}

Sailors steered by magnetized needles for a thousand years before
anyone knew why they point north. The answer came in 1820, when a
current made a compass twitch: magnetism is made by moving charges.
This chapter maps the magnetic field --- of magnets, of the Earth, of
currents --- then closes the loop: the field pushes back on currents,
which is how every electric motor turns.

\section{Magnets and the magnetic field}

\begin{definition}[Magnet and poles]\label{def:g11:magnetic-fields:magnet}
A \emph{magnet}\index{magnet} attracts iron and, hung by a thread,
turns to face north. Its action concentrates at two \emph{magnetic
poles}\index{magnetic pole}, \emph{north} (the end that seeks
geographic north) and \emph{south}: like poles repel, unlike attract.
Sawing a magnet in half yields two complete magnets --- no experiment
has ever isolated a pole.
\end{definition}

\begin{definition}[Magnetic field]\label{def:g11:magnetic-fields:bfield}
A magnet modifies the space around it
(\cref{ch:g11:electric-gravitational-fields}): it creates everywhere a
\emph{magnetic field}\index{magnetic field} $\vect B$. A compass reads
its direction (the needle aligns with $\vect B$, north end forward); a
\emph{Hall probe}\index{Hall probe} measures its magnitude, in
\emph{teslas}\index{tesla} (\unit{T}).
\end{definition}

\begin{definition}[Magnetic field lines]\label{def:g11:magnetic-fields:lines}
\emph{Magnetic field lines}\index{magnetic field line} are curves
tangent to $\vect B$, crowding where the field is strong. Outside a
magnet they run from north pole to south; unlike electric field lines
they have no ends: each closes through the magnet's body.
\end{definition}

\begin{omfigure}
\begin{tikzpicture}[scale=0.9,
  fline/.style={omThm, postaction={decorate},
    decoration={markings, mark=at position 0.5 with {\arrow{Stealth}}}}]
  \fill[omProp!15] (-1.4,-0.35) rectangle (0,0.35);
  \fill[omDef!15] (0,-0.35) rectangle (1.4,0.35);
  \draw[thick] (-1.4,-0.35) rectangle (1.4,0.35);
  \draw (0,-0.35) -- (0,0.35);
  \node[font=\small] at (-0.7,0) {S};
  \node[font=\small] at (0.7,0) {N};
  \draw[fline] (1.4,0.2) .. controls (2.6,0.9) and (-2.6,0.9) .. (-1.4,0.2);
  \draw[fline] (1.4,0.3) .. controls (3.8,2.0) and (-3.8,2.0) .. (-1.4,0.3);
  \draw[fline] (1.4,-0.2) .. controls (2.6,-0.9) and (-2.6,-0.9) .. (-1.4,-0.2);
  \draw[fline] (1.4,-0.3) .. controls (3.8,-2.0) and (-3.8,-2.0) .. (-1.4,-0.3);
  \draw[fline] (1.4,0) -- (2.95,0);
  \draw[omExo, thick] (0,1.1) circle (0.22);
  \draw[-Stealth, omDef, thick] (0.22,1.1) -- (-0.22,1.1);
  \draw[omExo, thick] (2.45,0) circle (0.22);
  \draw[-Stealth, omDef, thick] (2.23,0) -- (2.67,0);
\end{tikzpicture}

{\small Field lines of a bar magnet: out of N, into S, closed through
the body. A compass placed anywhere settles along the line through it.}
\end{omfigure}

\section{The Earth is a magnet}

\begin{definition}[The Earth's magnetic field]\label{def:g11:magnetic-fields:earth}
Currents in the Earth's molten iron core make the planet a giant
magnet: at the surface $B \approx \qty{50}{\micro T}$, and a free
needle aligns with the field's horizontal component --- this is the
compass. The geomagnetic poles are not the geographic ones; the angle
between compass north and true north is the
\emph{declination}\index{declination}, corrected on every bearing.
\end{definition}

\begin{example}[Correcting a bearing]\label{ex:g11:magnetic-fields:bearing}
Where the declination is $\ang{7}$ west, a ship steering ``compass
north'' for \qty{5.0}{km} drifts $5000 \times \sin\ang{7} \approx
\qty{6.1e2}{m}$ west of its track: to sail true north, steer $\ang{7}$
\emph{east} of the needle.
\end{example}

\section{The magnetic field of a current}

\begin{remark}[Oersted, 1820]\label{rem:g11:magnetic-fields:oersted}
During a lecture in 1820, Hans Christian Oersted saw a compass needle
swing when he closed a circuit nearby: electricity and magnetism were
one subject --- \emph{currents create magnetic fields}.
\end{remark}

\begin{proposition}[Field of a straight wire]\label{prop:g11:magnetic-fields:wire}
A long straight wire carrying a current $I$ creates a field whose
lines are circles centered on the wire, in planes perpendicular to it.
At distance $d$ from the wire,
\[
B = \frac{\mu_0\, I}{2\pi d},
\qquad \mu_0 = 4\pi\times 10^{-7}\ \unit{T.m/A} \approx \qty{1.26e-6}{T.m/A}.
\]
\end{proposition}
\admitted

\begin{omfigure}
\begin{tikzpicture}[scale=0.85]
  \node[font=\large] at (0,0) {$\odot$};
  \foreach \r in {0.55,1.05,1.55}
    \draw[omThm,-Stealth] (\r,0) arc (0:350:\r);
  \node[omThm, font=\small] at (45:1.9) {$\vect B$};
  \node[font=\small] at (0,-2.1) {current toward you};
\end{tikzpicture}\qquad\qquad
\begin{tikzpicture}[scale=0.85]
  \node[font=\large] at (0,0) {$\otimes$};
  \foreach \r in {0.55,1.05,1.55}
    \draw[omThm,-Stealth] (\r,0) arc (360:10:\r);
  \node[omThm, font=\small] at (45:1.9) {$\vect B$};
  \node[font=\small] at (0,-2.1) {current away from you};
\end{tikzpicture}

{\small Circular field lines around a straight wire, seen head-on
($\odot$: current out of the page; $\otimes$: into it). Grip the wire
with the right hand, thumb along the current: the fingers curl the way
$\vect B$ turns.}
\end{omfigure}

\begin{example}[A wire is a weak magnet]\label{ex:g11:magnetic-fields:wire}
At $d = \qty{1.0}{cm}$ from a wire carrying $I = \qty{15}{A}$:
$B = \num{2e-7} \times 15 / 0.010 = \qty{0.30}{mT}$ --- six Earth
fields, from a current that would trip a household breaker. Strong
fields need a better geometry.
\end{example}

\begin{definition}[Coil and solenoid]\label{def:g11:magnetic-fields:solenoid}
Winding the wire concentrates its field. A \emph{flat
coil}\index{coil} ($N$ turns in a disc) behaves like a thin magnet,
one face north, the other south; a \emph{solenoid}\index{solenoid} ---
a long cylindrical winding, $n$ turns per metre --- like a bar magnet.
\end{definition}

\begin{proposition}[Field inside a solenoid]\label{prop:g11:magnetic-fields:solenoidfield}
Inside a long solenoid, away from the ends, the field is
\emph{uniform}: parallel to the axis and the same at every interior
point, of magnitude $B = \mu_0\, n\, I$, independent of the bore's
width. Outside, the field is weak.
\end{proposition}
\admitted

\begin{remark}
Both formulas are derived honestly in the Year 2 volume, from the law
relating a field's circulation to the current it encircles. Note what
matters: $n$, turns \emph{per metre} --- wind tighter, not longer.
\end{remark}

\begin{method}[The right hand, twice]\label{met:g11:magnetic-fields:righthand}
\begin{itemize}
  \item \emph{Wire}: thumb = current; curled fingers = turning of the lines.
  \item \emph{Coil, solenoid}: curled fingers = current in the turns;
        thumb = $\vect B$ inside, pointing at the north face.
\end{itemize}
\end{method}

\begin{omfigure}
\begin{tikzpicture}[scale=0.9]
  \foreach \x in {0,0.65,...,3.9} {
    \node[font=\small] at (\x,0.85) {$\odot$};
    \node[font=\small] at (\x,-0.85) {$\otimes$};}
  \draw[-Stealth, omThm, thick] (-0.4,0.4) -- (4.3,0.4);
  \draw[-Stealth, omThm, thick] (-0.4,0) -- (4.3,0)
    node[right, font=\small] {$\vect B$};
  \draw[-Stealth, omThm, thick] (-0.4,-0.4) -- (4.3,-0.4);
  \draw[omExo, -Stealth, dashed]
    (4.5,0.7) .. controls (5.9,2.3) and (-2.0,2.3) .. (-0.6,0.7);
\end{tikzpicture}

{\small A solenoid cut lengthwise ($\odot$/$\otimes$: the turns
crossing the page): inside, a uniform $B = \mu_0 n I$ along the axis;
outside, a weak, spread-out return field (dashed).}
\end{omfigure}

\begin{example}[Solenoid numbers]\label{ex:g11:magnetic-fields:solenoid}
At $10$ turns per centimetre ($n = \qty{1000}{m^{-1}}$) and
$I = \qty{5.0}{A}$: $B = \num{1.26e-6} \times 1000 \times 5.0 \approx
\qty{6.3}{mT}$ --- a hundred Earth fields, adjustable by a knob.
\end{example}

\section{Electromagnets}

\begin{definition}[Electromagnet]\label{def:g11:magnetic-fields:electromagnet}
An \emph{electromagnet}\index{electromagnet} is a solenoid wound on a
soft iron core: the iron magnetizes in the coil's field and multiplies
it, typically by $100$ to $1000$. Unlike a permanent magnet, it
switches off with the current.
\end{definition}

\begin{remark}[Three machines]\label{rem:g11:magnetic-fields:machines}
The scrapyard crane lifts a car with an electromagnet and --- the
whole point --- \emph{drops} it by opening a switch. A relay's small
electromagnet pulls a contact closed: a weak current switches a
strong one (\cref{ch:g11:circuits-and-power}). An MRI scanner holds a uniform
\qty{1.5}{T} around the patient with a superconducting solenoid.
\end{remark}

Orders of magnitude, from the galaxy to a dead star:
\begin{center}
\small
\begin{tabular}{lr}
interstellar space & $\sim \qty{e-10}{T}$ \\
the Earth's surface field & \qty{5e-5}{T} \\
a fridge magnet & $\sim \qty{5e-3}{T}$ \\
a neodymium magnet, at its pole & $\sim \qty{0.5}{T}$ \\
an MRI solenoid & \qty{1.5}{T} \\
surface of a neutron star & $\sim \qty{e8}{T}$ \\
\end{tabular}
\end{center}

\section{The Laplace force}

\begin{proposition}[Laplace force]\label{prop:g11:magnetic-fields:laplace}
A straight wire of length $L$, carrying a current $I$ in a uniform
field $\vect B$ \emph{perpendicular} to the wire, feels the
\emph{Laplace force}\index{Laplace force} of magnitude $F = I\,L\,B$,
perpendicular to both wire and $\vect B$: flat right hand, thumb along
the current, straight fingers along $\vect B$ --- the palm pushes along
$\vect F$. A wire \emph{parallel} to the field feels no force.
\end{proposition}
\admitted

\begin{remark}[The jumping rail]\label{rem:g11:magnetic-fields:rail}
The Year 1 volume derives this from the magnetic force on each moving
charge. In the laboratory: a copper rod rests across two horizontal
rails between the poles of a magnet (field into the page in the
figure); close the switch, and the rod shoots along the rails. Reverse
the current, or flip the magnet: it shoots the other way.
\end{remark}

\begin{omfigure}
\begin{tikzpicture}[scale=0.95]
  \foreach \x in {0.6,1.5,2.9,3.8}
    \foreach \y in {0.35,1.05}
      \node[omThm, font=\small] at (\x,\y) {$\otimes$};
  \node[omThm, font=\small] at (4.35,1.05) {$\vect B$};
  \draw[very thick] (0,0) -- (4.6,0);
  \draw[very thick] (0,1.4) -- (4.6,1.4);
  \draw[omDef, line width=2.5pt] (2.1,-0.12) -- (2.1,1.52);
  \draw[-Stealth, omDef, thick] (0.1,1.6) -- (1.6,1.6)
    node[midway, above, font=\small] {$I$};
  \draw[-Stealth, omDef, thick] (2.32,1.15) -- (2.32,0.45);
  \draw[-Stealth, omDef, thick] (1.6,-0.22) -- (0.1,-0.22);
  \draw[-Stealth, omMeth, very thick] (2.1,0.7) -- (3.4,0.7)
    node[pos=0.85, above, font=\small] {$\vect F$};
\end{tikzpicture}

{\small The rail experiment: current down the rod, field into the page
($\otimes$) --- the Laplace force $F = ILB$ drives the rod along the
rails.}
\end{omfigure}

\begin{example}[Laplace numbers]\label{ex:g11:magnetic-fields:laplace}
A rod of length $L = \qty{5.0}{cm}$ carrying $I = \qty{8.0}{A}$ across
$B = \qty{0.50}{T}$: $F = 8.0 \times 0.050 \times 0.50 = \qty{0.20}{N}$
--- the weight of \qty{20}{g}, on a wire weighing a few grams.
\end{example}

\begin{remark}[The DC motor]\label{rem:g11:magnetic-fields:motor}
Bend the wire into a loop in the field: the two sides perpendicular to
$\vect B$ carry opposite currents, so their Laplace forces are opposite
--- a pair that spins the loop. Half a turn later the pair would spin
it back, so a rotating switch (the \emph{commutator}) reverses the
current every half-turn. Every fan and drill is this loop, multiplied.
\end{remark}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:g11:magnetic-fields:1}
A bar magnet is sawn in half between its poles. What poles does each
piece carry? Brought back together, do the freshly cut faces attract
or repel? Can any cutting scheme isolate the north pole?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g11:magnetic-fields:2}
True or false, with one reason each: (a) two magnetic field lines
cross where the field is strong; (b) outside a magnet, field lines run
from north pole to south pole; (c) every magnetic field line is a
closed loop; (d) a needle settles perpendicular to the line through it.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g11:magnetic-fields:3}
Compute the field \qty{2.0}{cm} from a straight wire carrying
\qty{10}{A}. How many times the Earth's \qty{50}{\micro T} is that?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g11:magnetic-fields:4}
A solenoid of $800$ turns wound over \qty{40}{cm} carries
$I = \qty{1.5}{A}$. Compute $n$, then $B$ inside; what does $B$ become
if the current doubles?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g11:magnetic-fields:5}
A wire segment of length \qty{10}{cm} carries \qty{5.0}{A}
perpendicular to a \qty{0.20}{T} field. Compute the Laplace force; the
weight of what mass equals it?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:magnetic-fields:6}
Where a hiker stands, the declination is $\ang{8}$ west. She walks
\qty{2.0}{km} ``north by compass'': how far, and to which side of true
north, does she drift? What heading would have walked her true north?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:magnetic-fields:7}
At what distance from a straight wire carrying \qty{20}{A} does its
field equal the Earth's \qty{50}{\micro T}? What does this say about
trusting a compass near live cables?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:magnetic-fields:8}
Design a solenoid producing $B = \qty{10}{mT}$ with $I = \qty{4.0}{A}$:
compute the required $n$, then the turn count for a \qty{25}{cm} coil.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:magnetic-fields:9}
In the rail experiment, the rod (mass \qty{20}{g}, length
$L = \qty{12}{cm}$ between the rails) carries \qty{10}{A} in a
\qty{0.50}{T} perpendicular field. Compute the Laplace force, then the
rod's initial acceleration; compare with $g$.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:magnetic-fields:10}
A horizontal wire of mass \qty{5.0}{g} and length \qty{20}{cm} sits in
a horizontal field $B = \qty{0.10}{T}$ perpendicular to it. What
current makes the Laplace force balance the wire's weight, so that it
levitates? Which way must the force point?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g11:magnetic-fields:11}
Give the direction of the Laplace force (or say it vanishes):
(a) current to the right of the page, $\vect B$ into the page;
(b) current toward the top of the page, $\vect B$ out of the page;
(c) current parallel to $\vect B$. Use the flat right hand.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g11:magnetic-fields:12}
A horizontal wire runs geographic north--south, \qty{3.0}{cm} above a
compass; the Earth's horizontal component there is \qty{20}{\micro T}.
The wire carries $I = \qty{5.0}{A}$: compute its field at the compass,
give its direction, then the needle's deflection (fields add as vectors).
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g11:magnetic-fields:13}
A scrapyard electromagnet is a coil of $400$ turns over \qty{20}{cm},
resistance \qty{2.0}{\ohm}, on a \qty{12}{V} supply; its iron core
multiplies the bare field by $100$. Compute the current, the bare-coil
field, the field with the core, and the power the coil dissipates
(\cref{ch:g11:circuits-and-power}). Why does the scrap fall the
instant the switch opens?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g11:magnetic-fields:14}
A motor's rectangular loop has $N = 50$ turns carrying
$I = \qty{2.0}{A}$; its two sides of length \qty{5.0}{cm} are
perpendicular to a field $B = \qty{0.30}{T}$. Compute the force on each
of these sides; why are the two forces opposite, and what do they do to
the loop? Why do the other two sides sometimes feel no force? Why must
the current be reversed every half-turn?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g11:magnetic-fields:15}
How close to a cable carrying \qty{100}{A} would you have to be for
its field to reach \qty{1}{T}? Compare with the cable's millimetre
radius, and conclude why tesla-strength fields are built from
solenoids, not single wires. Then: a neutron star's surface field of
\qty{e8}{T} is how many powers of ten above the Earth's?
\end{exercise}

\section{Problem: The compass, the crane and the MRI}

\begin{problem}[{Weekend problem --- three magnets at work: a compass
crossing an ocean, an electromagnet lifting a car, and an MRI solenoid
--- one field, spanning ten powers of ten, doing three jobs}]
\label{pb:g11:magnetic-fields:1}
The same vector $\vect B$ steers a sailboat with \qty{50}{\micro T},
lifts scrap iron with \qty{1}{T}, and images a knee with \qty{1.5}{T}
--- three machines for this chapter's three formulas.

\medskip
\noindent\textbf{Part I --- The compass.}
At the ship's position, the Earth's field has horizontal component
\qty{20}{\micro T} and vertical component \qty{44}{\micro T};
the declination is $\ang{6}$ west.
\begin{enumerate}
  \item Why does a compass needle point (magnetic) north at all?
  \item Compute the total field's magnitude and its angle with the horizontal.
  \item Steering ``compass north'' for \qty{30}{km}, how far west of
        the true-north track does the ship end up?
  \item Steel cargo deflects the needle a further $\ang{4}$ west: same question.
  \item Near the geomagnetic pole a compass is useless: which component
        is to blame, and why?
\end{enumerate}

\medskip
\noindent\textbf{Part II --- The crane.}
The scrapyard electromagnet is a solenoid of $600$ turns wound over
\qty{30}{cm}, of resistance \qty{1.5}{\ohm}, fed by a \qty{12}{V}
supply; its iron core multiplies the bare field by $50$.
\begin{enumerate}[resume]
  \item Compute $n$, the turn density.
  \item Compute the current in the coil (\cref{ch:g11:circuits-and-power}).
  \item Compute the bare-coil field $\mu_0 n I$.
  \item Compute the field with the core; compare with neodymium's \qty{0.5}{T}.
  \item Compute the power dissipated in the coil, then the energy for
        a \qty{10}{\minute} lifting shift.
  \item A permanent magnet this strong exists. Why an electromagnet anyway?
\end{enumerate}

\medskip
\noindent\textbf{Part III --- The winch motor.}
The crane's winch is a DC motor: a rectangular loop of $N = 100$ turns,
sides of length \qty{4.0}{cm} perpendicular to $B = \qty{0.25}{T}$,
carrying $I = \qty{3.0}{A}$.
\begin{enumerate}[resume]
  \item Compute the Laplace force on each side bundle perpendicular to $\vect B$.
  \item The two forces are equal and opposite, yet their effect does not
        cancel. What do they produce, and why must they be \emph{opposite}?
  \item What force acts on the other two sides, when parallel to $\vect B$?
  \item After half a turn, the same forces would undo the rotation.
        What does the commutator do, and when?
  \item Give three separate design changes that each double the force pair.
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- The MRI.}
The scanner's solenoid holds a uniform \qty{1.5}{T} with a current
$I = \qty{500}{A}$.
\begin{enumerate}[resume]
  \item How many times the Earth's field is \qty{1.5}{T}?
  \item Compute the turn density $n$ needed \emph{without} any iron core.
  \item If the winding had resistance \qty{0.20}{\ohm}, what power
        would it dissipate? What property of the actual winding avoids
        this, at the price of extreme cold?
  \item Punchline: place the ship's, the crane's and the MRI's fields
        on the powers-of-ten ladder from interstellar space
        (\qty{e-10}{T}) to a neutron star (\qty{e8}{T}); what changed
        between the machines --- the physics, or the amperes?
\end{enumerate}
\end{problem}
